### 2023-02-27 data-analytics.qmd
starting data crunching pipeline

---
title: "Data analytics"
format: pdf
---

```{r setup, include=FALSE}
library(tidyverse)
```

This script summarises the data analytical steps for the ICNS2023 paper.

```{mermaid}
%%| fig-width: 6
flowchart LR
  OSN[(Opensky \nNetwork)]
  OSN --> A[Data \nExtraction] --> B(Arrival \nTrajectories)
  B --> C{Decision}
  C --> D[Result one]
  C --> E[Result two]
```

## Opensky Data Extraction

* use/refine code developed for 3Di to download OSN-Monday-Dumps

```{r}
source("./R/osn_dumps.R")
source("./R/osn_read_tardump.R")
source("./R/osn_make_nice_names.R")
```

* OSN dumps have been stored for re-usage in a folder above the project hierarchy
* load a single day, e.g. 1. August 2019
* read in the chosen day
* standardise OSN naming conventions

```{r}
#| message: false
pth_to_dumps <- "../__DATA/OSN-dumps"
ptn <- "2019-08-01"

# list downloaded dumps
fns <- list.files(path = pth_to_dumps, pattern = ptn, full.names = TRUE)

# load list of chosen dumps
ds  <- fns |> 
  purrr::map_dfr(.f = read_osn_tar_file) |> 
  make_nice_names_osn()
```

* we want to study arrivals at certain airports, i.e. 200NM
* let's take London Heathrow (EGLL)
  + conversions done by hand & https://www.fcc.gov/media/radio/dms-decimal

EGLL Lat: 512839N Long: 0002741W   51.4775 -0461389
09L

089.67°

3901 x 50 M	
RWY surface: Asphalt, Grooved
PCN 83/F/A/W/T

SWY surface: Asphalt, Grooved
PCN 83/F/A/W/T

512839.00N 0002905.97W THR 78.6 FT
TDZ 81.3 FT	
27R

269.71°

3901 x 50 M	
RWY surface: Asphalt, Grooved
PCN 83/F/A/W/T

SWY surface: Asphalt, Grooved
PCN 83/F/A/W/T

512839.63N 0002559.82W
151.0 FT	
THR 78.1 FT
TDZ 79.3 FT	
09R

089.68°

3658 x 50 M	
RWY surface: Asphalt, Grooved
PCN 83/F/A/W/T

SWY surface: Asphalt, Grooved
PCN 83/F/A/W/T

512753.25N 0002856.33W
151.0 FT	
THR 75.3 FT
TDZ 76.2 FT	
27L

269.72°

3658 x 50 M	
RWY surface: Asphalt, Grooved
PCN 83/F/A/W/T

SWY surface: Asphalt, Grooved
PCN 83/F/A/W/T

512753.82N 0002602.76W
151.0 FT	
THR 76.8 FT
TDZ 77.8 FT

```{r}
# airport-data.com - server down???   source("./R/get_apt_info.R")
## aip_egll <- get_apt_info("EGLL")

apts <- c("EGLL","EDDF","EHAM","KATL","KJFK","KORD","WSSS")

# default back to ourairports.com
# url <- "https://davidmegginson.github.io/ourairports-data/airports.csv"
# aip_apts <- readr::read_csv(url)
# aip_apts <- aip_apts |> filter(ident %in% apts)
```

~REF, ~LAT, ~LON, ~ELEV, ~SRC
  "EGLL", "OSN-center", 51.4706, -0.461941, 83, "https://opensky-network.org/airport-profile?icao=EGLL"

```{r}
aip_apts <- tribble(
  ~ICAO,   ~REF,         ~LAT,    ~LON,     ~ELEV, ~SRC 
  ,"EGLL", "OSN-center", 51.4706, -0.461941, 83, "https://opensky-network.org/airport-profile?icao=EGLL"
  ,"EGLL", "ARP",        51.4775, -0.461389, NA, "AIP"
  ,"EGLL", "THR-09L",    51.4775, -0.484992, 79 ,"AIP - 512839.00N 0002905.97W THR 78.6 FT"
  ,"EGKK", "OSN-center", 51.148102,-0.190277, 202,"https://opensky-network.org/airport-profile?icao=EGKK"
  ,"EDDF", "OSN-center", 50.033333,8.570556, 364,"https://opensky-network.org/airport-profile?icao=EDDF"
  ,"EHAM", "OSN-center", 52.308601,4.763889, -11, "https://opensky-network.org/airport-profile?icao=EHAM"
  ,"KATL", "OSN-center", 33.6366996,-84.4281005,1026,"https://opensky-network.org/airport-profile?icao=KATL"
  ,"KORD", "OSN-center", 41.9785996,-87.9048004,672, "https://opensky-network.org/airport-profile?icao=KORD"
  ,"KJFK", "OSN-center", 40.6398010,-73.7789002, 13, "https://opensky-network.org/airport-profile?icao=KJFK"
)
write_csv(aip_apts, file = "./data/aip.csv")
aip_apts
```

Check for our study airport whether flights arrive (or depart)

* trajectory of LAT/LON within a certain distance around an airport and below a certain altitude

### 2023-02-27 paper.qmd
starting data crunching pipeline

---
# template taken from https://github.com/gael-close/ieeetran
#----------------------------------------------------------
title: Arrival Management with Open Data
format:
  ieeetran-pdf:
    keep-tex: true

author:
  - name: Rainer Koelle
    orcid: 0000-0000-0000-0000
    email: rainer.koelle@eurocontrol.int
    note: 1  # workaround i/o linked affiliation
  - name: Second Author
    orcid: 0000-0000-0000-0000
    email: second@example.org
    note: 1
  - name: Third Author
    email: third@example.org
    note: 2
#date: 01-Jan-2022.
affiliations:
  - id: 1
    name: Performance Review Unit, EUROCONTROL
    city: Brussels
    country: Belgium
  - id: 2
    name: Some corporation
    city: New York
    country: USA

abstract: |
  This document is a sample illustrating the Quarto `ieeetran` template.
  It includes the key elements of a scientific articles 
  (references, equations, figures, tables, code, cross references).
  The template enables the generation of IEEE-formatted article from a Jupypter notebook.
keywords: [template, demo]
reference-section-title: References
bibliography: bibliography.bib
note: Extra footnote.
# filters:
#    - latex-environment
environments: [mytable, mywidetable]
---

# Header 1

## header 2

some cool text.

Lorem Ipsum is simply dummy text of the printing and typesetting industry. Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book. It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged. It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.

### 2023-02-27 paper.tex
starting data crunching pipeline

\documentclass[letterpaper, 10 pt, journal, twoside]{IEEEtran}
\IEEEoverridecommandlockouts

% Quarto basics minus two problematic packages
\RequirePackage{scrlfile}
\PreventPackageFromLoading{sidenotes}
\PreventPackageFromLoading{marginnotes}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother

% Common packages
\usepackage[T1]{fontenc}
\usepackage{lipsum}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{physics}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage[hyphens]{url}
\usepackage{threeparttable}
\usepackage{xcolor}
\usepackage{float}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{orcidlink}
% monofont
\usepackage[scaled=0.8]{inconsolata}

%styled tabled generated from pandas
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{stfloats} % https://tex.stackexchange.com/a/324358

% Fix table in 2-column format and enable wide table
% https://tex.stackexchange.com/a/224096
\makeatletter
% My tables
\newenvironment{mytable}[1][htbp]{
    \begin{figure}[#1]
    \footnotesize
    \onecolumn
    \begin{minipage}{0.5\textwidth}}
{
    \end{minipage}
    \twocolumn
    \end{figure}}

\newenvironment{mywidetable}[1][tbp]{
    \begin{figure*}[#1]
    \footnotesize
    \onecolumn
    \begin{minipage}{1.0\textwidth}}
{
    \end{minipage}
    \twocolumn
    \end{figure*}}

\renewenvironment{table*}{\begin{mywidetable}}{\end{mywidetable}\ignorespacesafterend}
\renewenvironment{table}{\begin{mytable}}{\end{mytable}\ignorespacesafterend}    
\makeatother

% Muted text (for filler text)
\newcommand\muted[1]{%
\bgroup
\hskip0pt\color{black!40!}%
#1%
\egroup
}

% color links
\usepackage{hyperref}
\hypersetup{ colorlinks, citecolor=teal, linkcolor=teal, urlcolor=teal}

\begin{document}

% author blocks
\author{
        Rainer Koelle\(^1\)\orcidlink{0000-0000-0000-0000},     Second
Author\(^1\)\orcidlink{0000-0000-0000-0000},     Third Author\(^2\)
    \thanks{\(^1\)Performance Review Unit, EUROCONTROL. \(^2\)Some
corporation.}
    \thanks{Released on: }
    \thanks{Extra footnote..}
}

% title
\title{Arrival Management with Open Data}
\maketitle

% abstract
\begin{abstract}
    This document is a sample illustrating the Quarto \texttt{ieeetran}
    template. It includes the key elements of a scientific articles
    (references, equations, figures, tables, code, cross references).
    The template enables the generation of IEEE-formatted article from a
    Jupypter notebook.
\end{abstract}

% body
\hypertarget{header-1}{%
\section{Header 1}\label{header-1}}

\hypertarget{header-2}{%
\subsection{header 2}\label{header-2}}

some cool text.

Lorem Ipsum is simply dummy text of the printing and typesetting
industry. Lorem Ipsum has been the industry's standard dummy text ever
since the 1500s, when an unknown printer took a galley of type and
scrambled it to make a type specimen book. It has survived not only five
centuries, but also the leap into electronic typesetting, remaining
essentially unchanged. It was popularised in the 1960s with the release
of Letraset sheets containing Lorem Ipsum passages, and more recently
with desktop publishing software like Aldus PageMaker including versions
of Lorem Ipsum.

% done
\end{document}

### 2023-02-27 x_osm-extract-rwys.qmd
starting data crunching pipeline

---
title: "Extracting OSM data"
format: html
---

```{r setup}
# setup
library(tidyverse)
library(osmdata)
library(sf)
```

```{r}
# OSM query
bb_lonlat <- c(-0.223915,51.141733,-0.155851,51.165738)
apt_icao  <- "EGKK"
apt_name  <- "London Gatwick"

q <- opq(bbox = bb_lonlat) %>% 
  add_osm_feature(
    key = "aeroway"
    ,value =c("aerodrome", "apron", "control_tower", "gate", "hangar"
              ,"helipad", "runway", "taxiway", "terminal") ) %>% 
  osmdata_sf() 
# %>%
#   unique_osmdata()
```

```{r}
g <- ggplot() +
  geom_sf(data = q$osm_polygons
          ,inherit.aes = FALSE
          ,color = "lightblue"
          #,fill  = "lightblue"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway != "runway")
          , color = "grey"
          ) +
  geom_sf(data = q$osm_lines %>% filter(aeroway == "runway"),
          inherit.aes = FALSE,
          color = "black",
          size = 2 #.4
          ,alpha = .8) +
  theme_void() +
  labs(title = "EGKK")
g
```

Extract runway and endpoints

```{r}
rwys_ls <- q$osm_lines |> filter(aeroway == "runway", is.na(runway))

ggplot() + geom_sf(data = rwys_ls)
```

```{r}
rwys_pts <- st_cast(rwys_ls, "POINT")

ggplot() + 
  geom_sf(data = rwys_ls) +
  geom_sf(data = rwys_pts, color = "red")
```

Go for start and endpoints of linestring

```{r}
thrs_pts1 <- lwgeom::st_startpoint(rwys_ls)
thrs_pts2 <- lwgeom::st_endpoint(rwys_ls)

ggplot() + 
  geom_sf(data = rwys_ls) +
  geom_sf(data = thrs_pts1, size = 3, fill = NA) +
  geom_sf(data = thrs_pts2, size = 3, color = "lightgreen" , fill = NA) +
  geom_sf(data = rwys_pts, color = "red")

```

```{r}
my_pts1 <- lwgeom::st_startpoint(rwys_ls[1,])
my_pts2 <- lwgeom::st_endpoint(rwys_ls[1,])

ggplot() + 
  geom_sf(data = rwys_ls[1,]) +
  geom_sf(data = my_pts1, color = "red") + 
  geom_sf(data = my_pts2, color = "blue")
```

```{r}
extract_rwy_linestrings <- function(.osm_sf){
  rwys_ls <- .osm_sf$osm_lines |> filter(aeroway == "runway", is.na(runway))
  message("number of runways: ", nrow(rwys_ls))
  rwys_ls
}

pt_to_latlon <- function(.pt_sf){
  .pt_sf |> 
    sf::st_coordinates() |> 
    dplyr::as_tibble() |> 
    dplyr::rename(LON = X, LAT = Y) 
}

extract_thresholds <- function(.rwy_ls){
  thr1 <- lwgeom::st_startpoint(.rwy_ls) |> pt_to_latlon()
  thr2 <- lwgeom::st_endpoint(  .rwy_ls) |> pt_to_latlon()

  bind_rows(thr1, thr2)
}
```

```{r}
rwys_ls <- extract_rwy_linestrings(q)
extract_thresholds(rwys_ls)
```

### 2023-02-27 R/get_apt_info.R

@@ -1,0 +1,30 @@
+#' Retrieve airport information from airport-data.com via API
+#'
+#' For successful API requests, make sure that the API page and/or your access
+#' is configured. This is particularly important when accessing the internet
+#' through a proxy server (e.g. at work).
+#'
+#' @param .icao ICAO location indicator
+#'
+#' @return tibble of airport information
+#' @export
+#'
+#' @examples
+#' \dontrun{
+#' egll <- get_apt_info("EGLL")
+#' egll
+#' }
+get_apt_info <- function(.icao){
+  # check for ICAO location indicator
+  if(!is.character(.icao)) stop("Variable .icao must be a character!")
+  if(is.character(.icao) & nchar(.icao) != 4) stop("Variable .icao must be a 4-letter ICAO indicator!")
+
+  # send GET request to web API
+  request <- httr::GET("https://www.airport-data.com/api/ap_info.json", query = list(icao=.icao))
+
+  # extract content from json response
+  payload <- jsonlite::fromJSON(rawToChar(request$content))
+  payload <- tibble::as_tibble(payload)
+
+  return(payload)
+}

### 2023-02-27 R/osn_dumps.R

@@ -1,0 +1,81 @@
+#' Utility functions to discover and download one day (Monday) data from Opensky Network
+#'
+#' @return table of available dump dates
+#' @export
+#'
+#' @examples
+#' \dontrun{}
+osn_list_dumps <- function(){
+
+  url <- "https://opensky-network.org/datasets/states/"
+
+  # scrape page
+  dump_page <- rvest::read_html(url)
+
+  # extract available / prepared dumps
+  dumps <- dump_page %>%
+    rvest::html_element("tbody") %>%
+    rvest::html_table()
+
+  # filer Monday dumps
+  dumps <- dumps %>%
+    dplyr::filter(grepl(pattern = "^2[0-9]{3}", x = X1))    # filter 2020-xx-yy etc
+
+  return(dumps)
+}
+
+
+construct_dump_download_links <- function(.date){
+
+  url <- "https://opensky-network.org/datasets/states/"
+  url_date <- paste0(
+     url
+    ,.date
+  )
+
+  # scrape page
+  hour_chunks <-
+    rvest::read_html(url_date) %>%
+    rvest::html_element("tbody") %>%
+    rvest::html_table() %>%
+    dplyr::filter(grepl(pattern = "[0-9]{2}", x = X1))
+
+  # links
+  urls <- paste0(url_date, "/", dplyr::pull(hour_chunks, X1))
+
+  return(urls)
+}
+
+
+osn_download_hour_dump <- function(.url_date_hr, .dest_path = NULL){
+
+  # check for tar-file exists
+  target <-
+    rvest::read_html(.url_date_hr) %>%
+    rvest::html_element("tbody") %>%
+    rvest::html_table() %>%
+    dplyr::filter(grepl(pattern = ".csv.tar", x = X1))
+
+  if (nrow(target) != 1) {
+    message(paste0("No download for ", .url_date_hr))
+    # df <- tibble::tibble(COMMENT = "No data dump or multiple target files.")
+  } else{
+    message(paste0("Download tar exists."))
+
+    download_url = paste0(.url_date_hr, "/", target$X1)
+
+    if (is.null(.dest_path)) {
+      destination <- basename(download_url)
+    } else{
+      destination <- paste0(.dest_path, "/", basename(download_url))
+    }
+
+    download.file(
+        url      = download_url
+      , destfile = destination
+      , method   = "libcurl"
+   ## https://stackoverflow.com/questions/48927391/trouble-downloading-tar-in-r
[…]

### 2023-02-27 R/osn_make_nice_names.R

@@ -1,0 +1,20 @@
+#' Standardise Opensky Network naming conventions
+#'
+#' @param .df OSN data tibble
+#'
+#' @return
+#' @export
+#'
+#' @examples
+make_nice_names_osn <- function(.df){
+  df <- .df %>%
+    dplyr::rename(
+      TIME   = time
+      ,ICAO24 = icao24
+      ,   LAT = lat, LON = lon
+      , FLTID = callsign
+      , ALT_B = baroaltitude
+      , ALT_G = geoaltitude
+    )
+  return(df)
+}

### 2023-02-27 R/osn_read_tardump.R

@@ -1,0 +1,28 @@
+#' Read downloaded tar files from Opensky Network
+#'
+#' @param .fn filename
+#' @param .exdir temporary directory to extract tar files
+#' @param .remove_exdir remove the temporary directory
+#'
+#' @return tibble of extracted OSN tars
+#' @export
+#'
+#' @examples
+#' \dontrun{}
+read_osn_tar_file <- function(.fn, .exdir="tmp_tar", .remove_exdir=TRUE){
+  # inspired by: https://stackoverflow.com/questions/7151145/unzip-a-tar-gz-file
+  
+  # identify gz file name from tarball 
+  content <- untar(tarfile = .fn, list = TRUE)
+  payload <- content[grepl(pattern = ".csv.gz", x = content)]
+  
+  # extract payload data file
+  untar(tarfile = .fn, files = payload, exdir = .exdir)
+  
+  df <- readr::read_csv(paste0("./", .exdir, "/", payload))
+  
+  if(.remove_exdir == TRUE){  # --- clean-up by removing temp dir ----------
+    unlink(.exdir, recursive = TRUE)
+  }
+  return(df)
+}

### 2023-02-28 data-analytics.qmd
data prep step: extract runway information for study from OSM

@@
 )
-write_csv(aip_apts, file = "./data/aip.csv")
+#write_csv(aip_apts, file = "./data/aip.csv")
 aip_apts
@@
 * trajectory of LAT/LON within a certain distance around an airport and below a certain altitude
+* we start with Heathrow, filtering for some arrivals
+
+> ds |> filter(grepl(pattern = "^BAW", x = FLTID ), ALT_B < 1000) |> distinct(FLTID)
+# A tibble: 761 × 1
+   FLTID  
+   <chr>  
+ 1 BAW112B
+ 2 BAW32  
+ 3 BAW237 
+ 4 BAW236 
+ 5 BAW16  
+ 6 BAW28  
+ 7 BAW34  
+ 8 BAW56  
+ 9 BAW262 
+10 BAW74
+
+```{r}
+baws <- ds |> filter(grepl(pattern = "^BAW", x = FLTID )) |> 
+  group_by(FLTID) |> 
+  mutate(
+    MST = case_when(
+          TIME == min(TIME) ~ "START"
+          ,TIME == max(TIME) ~ "END")
+    )
+```
+```{r}
+picks <- c("BAW112B", "BAW236","BAW56","BAW262","BAW74")
+
+ggplot() +
+  geom_point(data = aip_apts |> filter(ICAO == "EGLL")
+             , aes(x = LON, y = LAT), size = 3, color = "blue") +
+  geom_point(data = baws |> filter(FLTID %in% picks)
+             , aes(x = LON, y = LAT, color = MST))
+```

### 2023-02-28 x_osm-extract-rwys.qmd
data prep step: extract runway information for study from OSM

@@
 ---
+
+Development version
+
+Data preparatory step: extract runway information from Openstreet Map (OSM).
 
@@
 extract_rwy_linestrings <- function(.osm_sf){
-  rwys_ls <- .osm_sf$osm_lines |> filter(aeroway == "runway", is.na(runway))
-  message("number of runways: ", nrow(rwys_ls))
-  rwys_ls
+
+ osm_values <- c("osm_id", "aeroway", "ref", "runway", "geometry")
+
+  this_rwys_ls <- .osm_sf$osm_lines |>
+    dplyr::select(any_of(osm_values)) 
+  message("missing values: ", setdiff(osm_values, names(this_rwys_ls)))
+
+  # fix EDDF no runway-tag!
+  if("runway" %in% setdiff(osm_values, names(this_rwys_ls))){
+    this_rwys_ls <- this_rwys_ls |> mutate(runway = NA_character_)
+  }
+
+  this_rwys_ls <- this_rwys_ls |> 
+    dplyr::filter(aeroway == "runway", is.na(runway))
+  message("number of runways: ", nrow(this_rwys_ls))
+  this_rwys_ls
 }
@@
 
-extract_thresholds <- function(.rwy_ls){
-  thr1 <- lwgeom::st_startpoint(.rwy_ls) |> pt_to_latlon()
-  thr2 <- lwgeom::st_endpoint(  .rwy_ls) |> pt_to_latlon()
+extract_thresholds <- function(.rwys_ls, .drop_geometry = FALSE){
+  rwys <- .rwys_ls |> select(aeroway, ref, geometry)
+  thr1 <- lwgeom::st_startpoint(.rwys_ls) |> pt_to_latlon() |> rename(THR1_LAT = LAT, THR1_LON = LON)
+  thr2 <- lwgeom::st_endpoint(  .rwys_ls) |> pt_to_latlon() |> rename(THR2_LAT = LAT, THR2_LON = LON)
 
-  bind_rows(thr1, thr2)
+  bind_cols(rwys, thr1, thr2)  # todo: add drop_geometry
 }
@@
 ```
+
+get thresholds
+
+* sfnetworks has edge_azimuth() function, coerce bearing into degrees
+* calculate (rough) offset between bearing and runway identifier & select
+
+```{r}
+rq <-  extract_thresholds(rwys_ls)
+
+rq2 <- rq |> 
+  sfnetworks::as_sfnetwork() |> 
+  sfnetworks::activate("edges") |> 
+  dplyr::mutate(  
+      BRG = sfnetworks::edge_azimuth()
+    , BRG = units::set_units(BRG, "degrees")
+    , BRG = (as.numeric(BRG) + 360) %% 360
+
+    , REF1 = stringr::str_split_i(ref, pattern = "/",1)
+    , REF2 = stringr::str_split_i(ref, pattern = "/",2)
+
+    , RWY_1 = if_else( abs( (parse_number(REF1) * 10) - BRG ) < 30, REF1, REF2)
+    , RWY_2 = if_else( abs( (parse_number(REF2) * 10) - BRG ) > 30, REF2, REF1)
+         )
+rq2
+```
+
+Package output
+
+```{r}
+apt <- "EGKK"
+
+# pull out edges/runways 
+rq3 <- rq2 |> sfnetworks::activate("edges") |> st_as_sf() |> sf::st_drop_geometry()
+
+rwy_ends <-
+  bind_cols(rq3 |> select(REF = RWY_1, LAT = THR1_LAT, LON = THR1_LON, NAME = ref)) |> 
+  bind_rows(rq3 |> select(REF = RWY_2, LAT = THR2_LAT, LON = THR2_LON, NAME = ref)) |> 
+  mutate(ICAO = apt) |> select(ICAO, everything())
+rwy_ends
+```
+
+```{r}
+ggplot() + 
+  geom_sf(data = rwys_ls) +
+  geom_point(data = rwy_ends, aes(x = LON, y = LAT)) +
+  geom_text(data = rwy_ends, aes(x = LON, y = LAT, label = REF), vjust = -0.5)
+```
+
+Let's go for Heathrow
+
+```{}
+# fails from github! ### source("./R/xhelper_read_excel_from_url.R")
+url   <- "https://github.com/euctrl-pru/airport-dashboard/blob/master/data/APT_BBOX.xlsx"
+# github urls follow the pattern: https://raw.github.com/user/repository/branch/file.name
+github_download_url <- "https://raw.github.com/euctrl-pru/airport-dashboard/master/data/APT_BBOX.xlsx"
+
+## bboxs <- read_excel_from_url(url)
+download.file(github_download_url, destfile = "./data/airport-bbox.xlsx", mode = "wb")
+bboxs <- readxl::read_excel("./data/airport-bbox.xlsx")
+```
+
+```{r}
+query_osm_bbox <- function(.bb_lonlat){
+  q <- opq(bbox = .bb_lonlat) %>% 
+    add_osm_feature(
+      key = "aeroway"
+      ,value =c("aerodrome", "apron", "control_tower", "gate", "hangar"
+              ,"helipad", "runway", "taxiway", "terminal") ) %>% 
+    osmdata_sf() 
+}
+
+plot_airport <- function(.osm_apt, .apt=NULL){
+  g <- ggplot() +
+  geom_sf(data = .osm_apt$osm_polygons
+          ,inherit.aes = FALSE
+          ,color = "lightblue"
+          #,fill  = "lightblue"
+          ) +
+  geom_sf(data = .osm_apt$osm_lines %>% filter(aeroway != "runway")
+          , color = "grey"
+          ) +
+  geom_sf(data = .osm_apt$osm_lines %>% filter(aeroway == "runway"),
+          inherit.aes = FALSE,
+          color = "black",
+          size = 2 #.4
+          ,alpha = .8) +
+  theme_void() +
+  labs(title = .apt)
+ return(g)
+}
+
+coerce_bbox_string_to_numeric_vector <- function(.bbox_string){
+  .bbox_string |> 
+    str_split_1(pattern = ",") |> 
+    as.numeric()
+}
+```
+
+```{r}
+egll_bb_lonlat <- bboxs |> 
+  filter(AIRPORT == "EGLL") |> pull(BBOX) |> 
+  coerce_bbox_string_to_numeric_vector()
+
+osm_apt <- query_osm_bbox(egll_bb_lonlat)
+
+plot_airport(osm_apt)
+```
+
+Check for runway ends
+
+```{r}
+rwys_ls <- extract_rwy_linestrings(osm_apt)
+extract_thresholds(rwys_ls)
+```
+
+```{r}
+associate_rwy_brg <- function(.thrs_ls){
+  .thrs_ls |> 
+  sfnetworks::as_sfnetwork() |> 
+  sfnetworks::activate("edges") |> 
+  dplyr::mutate(  
+      BRG = sfnetworks::edge_azimuth()
+    , BRG = units::set_units(BRG, "degrees")
+    , BRG = (as.numeric(BRG) + 360) %% 360
+
+    , REF1 = stringr::str_split_i(ref, pattern = "/",1)
+    , REF2 = stringr::str_split_i(ref, pattern = "/",2)
+
+    , RWY_1 = if_else( abs( (parse_number(REF1) * 10) - BRG ) < 30, REF1, REF2)
+    , RWY_2 = if_else( RWY_1 == REF1, REF2, REF1)
+         )
+}
+# rq2 <- rq |> associate_rwy_brg()
+# rq2
+
+pull_out_runway_info <- function(.thrs_brgs, .apt = apt){
+# pull out edges/runways 
+ tmp <- .thrs_brgs |> 
+   sfnetworks::activate("edges") |> 
+   sf::st_as_sf() |> 
+   sf::st_drop_geometry()
+
+ rwy_ends <-
+  bind_cols(tmp |> select(REF = RWY_1, LAT = THR1_LAT, LON = THR1_LON, NAME = ref)) |> 
+  bind_rows(tmp |> select(REF = RWY_2, LAT = THR2_LAT, LON = THR2_LON, NAME = ref)) |> 
+  mutate(ICAO = .apt) |> select(ICAO, everything())
+ rwy_ends
+}
+```
+
+Executed for
+
+* EGLL
+* EHAM
+* EGKK
+* EDDF
+* WSSS
+* KATL, KORD, KJFK
+* SBGR
+
+```{r}
+apt <- "SBGR"
+
+# European airports
+bboxs <- readxl::read_excel("./data/airport-bbox.xlsx")
+# international airports
+bboxs_int <- readxl::read_excel("./data/airport-intnl-bbox.xlsx")
+bboxs <- bind_rows(bboxs, bboxs_int)
+
+bb_lonlat <- bboxs |> 
+  filter(AIRPORT == apt) |> pull(BBOX) |> 
+  coerce_bbox_string_to_numeric_vector()
+
+osm_apt <- query_osm_bbox(bb_lonlat)
+
+rwys_ls <- extract_rwy_linestrings(osm_apt)
+
+if(apt == "EDDF"){
+  rwys_ls$ref <- gsub(pattern = "^18$", replacement = "18/36", x = rwys_ls$ref)
+}
+
+rwy_ends <-  rwys_ls |> 
+  extract_thresholds() |> 
+  associate_rwy_brg() |> 
+  pull_out_runway_info()
+
+write_csv(rwy_ends, file = paste0("./data/rwys-",apt,".csv"))
+rwy_ends
+```
+
+Visual check
+
+```{r}
+plot_airport(osm_apt) +
+  geom_point(data = rwy_ends, aes(x = LON, y = LAT)) +
+  geom_text( data = rwy_ends, aes(x = LON, y = LAT, label = REF), vjust = -0.3)
+```

### 2023-02-28 R/xhelper_read_excel_from_url.R

@@ -1,0 +1,22 @@
+## fails miserably when trying to download excel sheet from github
+## need to recode github download urls!
+
+#' Downlad an excel sheet
+#'
+#' @param .url 
+#' @param .sheet 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+#' #' \dontrun{
+#' fn <- "https://ansperformance.eu/download/xls/Taxi-Out_Additional_Time.xlsx"
+#' ds <- download_excel_from_url(fn, .sheet = "DATA)
+#' egll
+#' }
+read_excel_from_url <- function(.url, .sheet=NULL){
+  httr::GET(.url, httr::write_disk(tf <- tempfile(  #tmpdir = "./tmp", 
+    fileext = ".xlsx")))
+  ds <-  readxl::read_excel(tf, sheet = .sheet)
+}

### 2023-02-28 paper.qmd
initial stab at Quarto paper

@@
 
-# Header 1
+# Introduction
+
+The contribution of this paper comprise:
+
+-   conceptualisation of sequencing separation for arrival management and development of an open data and open software based implementation of the approach; and
+
+-   use-case application of the developed approach on a subset of airports within the European <!-- and international context -->
+
+# Trajectory-Based Operations - Arrival Management
 
@@
 It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+# Data and Conceptual Approach
+
+## Approach
+
+data preparation
+
+-   trajectory data - Opensky Network, weekly downloads
+-   airport information - Openstreet Map
+
+data downloaded & script development, data cleaning
+
+## Open Data for Operational Performance Monnitoring
+
+## Arrival Sequencing
+
+sequence separation
+
+# Case Study - Results
+
+## Data Sampling
+
+At the time of writing no global open flight table exists.
+For this study, we validated the sample with reference data available to the Performance Review Unit.
+Under the EUROCONTROL Performance Review System, airport operators report movement data on a monthly basis [@apdf_v1_2019].
+
+# Conclusions
+
+<!-- what did this paper show -->
+
+This paper aimed at exploring a data-driven approach to measuring arrival management based on open data.
+
+<!-- strength and weaknesses -->
+
+<!-- application -->
+
+<!-- further research -->
+
+<!-- final appraisal -->
+
+# Reproducibility  {.unnumbered}
+
+This paper has been built with the R/RStudio ecosystem.
+The draft manuscript and its supporting data preparatory steps are archived at https://github.com/rainer-rq-koelle/paper-2023-ICNS.
+
+The script to download the weekly global datasets are included.
+The cleaned trajectory data is stored at: \<\<tbd\>\>.
+
+# Acknowledgment and Disclaimer {.unnumbered}
+
+The authors thank the comments by XX YY ZZ.
+
+The views expressed are the authors' own and do not represent a policy or position of EUROCONTROL <!-- and co-authors organisations -->.

### 2023-02-28 paper.tex
initial stab at Quarto paper

@@
 \makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 % body
-\hypertarget{header-1}{%
-\section{Header 1}\label{header-1}}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+The contribution of this paper comprise:
+
+\begin{itemize}
+\item
+  conceptualisation of sequencing separation for arrival management and
+  development of an open data and open software based implementation of
+  the approach; and
+\item
+  use-case application of the developed approach on a subset of airports
+  within the European
+\end{itemize}
+
+\hypertarget{trajectory-based-operations---arrival-management}{%
+\section{Trajectory-Based Operations - Arrival
+Management}\label{trajectory-based-operations---arrival-management}}
 
@@
 
+\hypertarget{data-and-conceptual-approach}{%
+\section{Data and Conceptual
+Approach}\label{data-and-conceptual-approach}}
+
+\hypertarget{approach}{%
+\subsection{Approach}\label{approach}}
+
+data preparation
+
+\begin{itemize}
+\tightlist
+\item
+  trajectory data - Opensky Network, weekly downloads
+\item
+  airport information - Openstreet Map
+\end{itemize}
+
+data downloaded \& script development, data cleaning
+
+\hypertarget{open-data-for-operational-performance-monnitoring}{%
+\subsection{Open Data for Operational Performance
+Monnitoring}\label{open-data-for-operational-performance-monnitoring}}
+
+\hypertarget{arrival-sequencing}{%
+\subsection{Arrival Sequencing}\label{arrival-sequencing}}
+
+sequence separation
+
+\hypertarget{case-study---results}{%
+\section{Case Study - Results}\label{case-study---results}}
+
+\hypertarget{data-sampling}{%
+\subsection{Data Sampling}\label{data-sampling}}
+
+At the time of writing no global open flight table exists. For this
+study, we validated the sample with reference data available to the
+Performance Review Unit. Under the EUROCONTROL Performance Review
+System, airport operators report movement data on a monthly basis
+\protect\hyperlink{ref-apdf_v1_2019}{{[}1{]}}.
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
+
+This paper aimed at exploring a data-driven approach to measuring
+arrival management based on open data.
+
+\hypertarget{reproducibility}{%
+\section*{Reproducibility}\label{reproducibility}}
+\addcontentsline{toc}{section}{Reproducibility}
+
+This paper has been built with the R/RStudio ecosystem. The draft
+manuscript and its supporting data preparatory steps are archived at
+https://github.com/rainer-rq-koelle/paper-2023-ICNS.
+
+The script to download the weekly global datasets are included. The
+cleaned trajectory data is stored at:
+\textless\textless tbd\textgreater\textgreater.
+
+\hypertarget{acknowledgment-and-disclaimer}{%
+\section{Acknowledgment and
+Disclaimer}\label{acknowledgment-and-disclaimer}}
+
+The authors thank the comments by XX YY ZZ.
+
+The views expressed are the authors' own and do not represent a policy
+or position of EUROCONTROL .
+
+\hypertarget{bibliography}{%
+\section*{References}\label{bibliography}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{0}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-apdf_v1_2019}{}}%
+\CSLLeftMargin{{[}1{]} }%
+\CSLRightInline{EUROCONTROL, {``Eurocontrol {Specification} for
+{Operational ANS Performance Monitoring} - {Airport Operator Data
+Flow}.''} 2019 {[}Online{]}. Available:
+\url{https://www.eurocontrol.int/publication/eurocontrol-specification-operational-ans-performance-monitoring}}
+
+\end{CSLReferences}
+
 % done

### 2023-02-28 data-analytics.qmd
associated landing runway (first stab)

@@
 library(tidyverse)
+
+ggplot2::theme_set(ggplot2::theme_minimal())
 ```
@@
 
-> ds |> filter(grepl(pattern = "^BAW", x = FLTID ), ALT_B < 1000) |> distinct(FLTID)
-# A tibble: 761 × 1
-   FLTID  
-   <chr>  
- 1 BAW112B
- 2 BAW32  
- 3 BAW237 
- 4 BAW236 
- 5 BAW16  
- 6 BAW28  
- 7 BAW34  
- 8 BAW56  
- 9 BAW262 
-10 BAW74
-
 ```{r}
@@
 ```
+
+show temporal and vertical profile
+
+```{r}
+ggplot(data = baws |> filter(FLTID %in% picks)) +
+  geom_point(aes(x = TIME, y = ALT_B, color = FLTID), size = 0.3)
+```
+
+save out some test trajectories
+
+write_csv(baws |> filter(FLTID %in% picks), file = "./data/tst-egll-baw.csv.gz")
+
+airport arrival box
+
+```{r}
+
+```

### 2023-02-28 paper.qmd
associated landing runway (first stab)

@@
 
-## Open Data for Operational Performance Monnitoring
+The identification of the landing direction is based on a simple geospatial heuristic.
+Conceptually, aircraft are aligned with the runway (centerline) before landing.
+An aircraft is assigned to a landing runway based on the closeness of its pre-landing positions to the extended runway centerline.
+
+## Open Data for Operational Performance Monitoring
 

### 2023-02-28 x_trj-apt-landing-association.qmd
associated landing runway (first stab)

---
title: "Trajectory Landing Runway Assignment"
format: html
---

Development version

Data preparatory step: associated a trajectory to a landing runway, including determining characteristic timestamps (e.g. gate/6NM time, landing time).

```{r setup}
# setup
library(tidyverse)
library(sf)

# set default ggplot theme
ggplot2::theme_set(ggplot2::theme_minimal())
```

The identification of the landing direction is based on a simple geospatial heuristic.    
Conceptually, aircraft are aligned with the runway (centerline) before landing.
An aircraft is assigned to a landing runway based on the closeness of its pre-landing positions to the extended runway centerline.

Load test data set (BAWs @ EGLL)

```{r}
tst_fn  <- "tst-egll-baw.csv.gz"
tst_pth <- here::here("data", tst_fn)
baws <- read_csv(tst_pth, show_col_types = FALSE)
```

Data preparation

```{r}
source(here::here("R","coerce_unixepoch_to_datetime.R"))
source(here::here("R", "cast_sf_utils.R"))

# load EGLL environment
rwys_egll <- read_csv("./data/rwys-EGLL.csv", show_col_types = FALSE)
arp_egll  <- read_csv("./data/aip.csv", show_col_types = FALSE) |>
  filter(ICAO == "EGLL", REF == "OSN-center")

baws   <- baws |> coerce_unixepoch_to_datetime()

baw236 <- baws |> filter(FLTID == "BAW236") |> 
  trrrj::filter_positions_at_range(c(arp_egll$LON, arp_egll$LAT), dm = 1, dM = 205, lon = LON, lat = LAT)

baw236_ls <- baw236 |> cast_latlon_to_ls(.drop_coord = FALSE, .group_var = ICAO24)
```

```{r}
ggplot() +
  geom_sf(data = arp_egll |> cast_latlon_to_pts(), size = 3, color = "blue") +
  geom_sf(data = rwys_egll |> cast_latlon_to_pts(), color = "red") +
  geom_sf(data = baw236_ls) +
  labs(title = "Overall trajectory - arrival path")
```

Zoom in on airport area

```{r}
baw236_landing <- baw236 |> 
  trrrj::filter_positions_at_range(c(arp_egll$LON, arp_egll$LAT), 0, 20, lon = LON, lat = LAT)
baw236_landing_ls <- baw236_landing |> cast_latlon_to_ls(.drop_coord = FALSE, .group_var = ICAO24)

ggplot() +
  geom_sf(data = arp_egll |> cast_latlon_to_pts(), size = 3, color = "blue") +
  geom_sf(data = rwys_egll |> cast_latlon_to_pts(), color = "red") +
  geom_sf(data = baw236_landing_ls) +
  labs(title = "Arrival portion")
```

focus on runway

```{r}
# setup sf objects
arp_pt   <- arp_egll |> cast_latlon_to_pts()
rwys_pts <- rwys_egll |> cast_latlon_to_pts()
rwys_ls  <- rwys_egll |> 
  group_by(NAME) |> 
  group_modify(.f = ~ cast_latlon_to_ls(.) ) |> 
  st_as_sf()

# utility function to create centerline
rwy_ctr_line <- function(.rwy_df = aip, .ctrl_length = 10000){
  df <- .rwy_df %>% 
    filter(REF != "ARP") %>% 
    select(REF, NAME, LAT, LON) %>% 
    group_by(NAME) %>% 
    mutate( LAT2 = case_when(!is.na(lag(LAT))  ~ lag(LAT)
                            ,!is.na(lead(LAT)) ~ lead(LAT))
           ,LON2 = case_when(!is.na(lag(LON))  ~ lag(LON)
                            ,!is.na(lead(LON)) ~ lead(LON))
          ) %>% 
  # calculate "reverse" runway bearing with geosphere::bearingRhumb
  mutate( RBRG  = geosphere::bearingRhumb(p1 = cbind(LON2, LAT2), p2 = cbind(LON, LAT))
         )

  # determine "endpoint" of extended centerline at d = 10000 meters
  tmp <- with(df, geosphere::destPointRhumb(cbind(LON, LAT), b= RBRG, d = .ctrl_length)) %>% 
    as_tibble() %>% rename(LON3 = lon, LAT3 = lat)

  # combine and return
  df <- df %>% bind_cols(tmp)
  return(df)
}

# utility function to coerce extension points
rwy_ctr_line_pts <- function(.rwy_ctr_line, .debug = FALSE){
  # could not get pivot_longer work with multiple cols
  tmp1 <- .rwy_ctr_line %>% select("REF":"LON")

  # include opposite runway threshold
  tmp2 <- .rwy_ctr_line %>% select("REF","NAME", "LAT2","LON2") %>%
    rename(LAT = LAT2, LON = LON2)

  # centerline end point determined beforehand
  tmp3 <- .rwy_ctr_line %>% select("REF","NAME", "LAT3","LON3") %>%
    rename(LAT = LAT3, LON = LON3)

  if(.debug == TRUE){ 
    df <- bind_rows(tmp1, tmp2, tmp3)
  }else{
    df <- bind_rows(tmp1, tmp3)
  }

  df <- df %>% arrange(REF, NAME)
  return(df)  
}

cast_rwy_ctr_line_ls <- function(.rwys_df){
  .rwys_df |> 
  rwy_ctr_line() |> 
  rwy_ctr_line_pts() |> 
  cast_latlon_to_pts() |> 
  cast_pts_to_ls(.group_var = REF)
}
rwy_centerline_ls <- cast_rwy_ctr_line_ls(rwys_egll)
```

```{r}
ggplot() +
  geom_sf(data = arp_pt, size = 3, color = "blue") +
  geom_sf(data = rwys_ls) +
  geom_sf(data = rwys_pts, color = "red") +
  geom_sf(data = rwy_centerline_ls, color = "green") +
  geom_sf_text(data = rwy_centerline_ls, aes(label = REF)) +
  geom_sf(data = baw236_landing_ls)
```

check proximity on final

```{r}
# spatial join (aka "overlay" with "within")
# to identify most likely ~ closest runway (extended centerline-string)
estimate_rwy <- function(.trj_pts, .rwy_ctrl_ls, .ctrl_offset = 1000){
  # stop if not
  tmp <- st_join( .rwy_ctrl_ls, .trj_pts
                 , join = st_is_within_distance, dist = .ctrl_offset) %>%
  st_drop_geometry() %>%        # keep just the "spatially joined" dataframe
  na.omit() %>%                 # eliminate no hits
  group_by(REF) %>%
  summarise(N = n(), .groups = "drop") %>%
  mutate( TOT   = sum(N, na.rm = TRUE)       # na.rm should not be necessary
         ,TRUST = N / TOT)                   # share in case several runways

return(tmp)
}
```

```{r}
baw236_final_approach <- baw236 |> 
  trrrj::filter_positions_at_range(c(arp_egll$LON, arp_egll$LAT), 0, 10, lon = LON, lat = LAT)
# baw236_landing_ls <- baw236_landing |> cast_latlon_to_ls(.drop_coord = FALSE, .group_var = ICAO24)
ggplot() +
  geom_sf(data = arp_pt, size = 3, color = "blue") +
  geom_sf(data = rwys_ls) +
  geom_sf(data = rwys_pts, color = "red") +
  geom_sf(data = rwy_centerline_ls, color = "green") +
  geom_sf_text(data = rwy_centerline_ls, aes(label = REF)) +
  geom_sf(data = baw236_final_approach |> cast_latlon_to_ls())
```

```{r}
estimate_rwy(rwy_centerline_ls, baw236_final_approach |> cast_latlon_to_pts())
```

Try to understand why / how the closeness is calculated.
Current implementation shows N=1 ?!? while ok, it should be multiple hits.

### 2023-02-28 R/cast_sf_network.R

@@ -1,0 +1,15 @@
+#' Cast a dataframe/tibble of LAT/LON positions to a unidirected sf-network
+#'
+#' @param .df dataframe/tibble of LAT/LON positions (of typically a trajectory/flight)
+#' @param .crs coordinate reference system
+#' @param .drop_coord
+#'
+#' @return sfnetwork object of trajectory
+#' @export
+#'
+#' @examples
+cast_latLon_to_sfnet <- function(.df, .crs = 4326, .drop_coord = FALSE){
+  df <- .df %>%
+    cast_latlon_to_pts(.crs = .crs, .drop_coord = .drop_coord) %>%
+    sfnetworks::as_sfnetwork()
+}

### 2023-02-28 R/cast_sf_steps.R

@@ -1,0 +1,45 @@
+# convert sequence of points to step linestrings
+# inspiration from SO
+# https://stackoverflow.com/questions/64594460/create-line-segments-from-gps-points-in-r-sf
+# source_file <- 'INSERT_FILENAME_HERE'
+#
+# empty <- st_as_sfc("POINT(EMPTY)", crs = 4326)
+#
+# sf::st_read(source_file) %>%
+#   # great circle distances
+#   st_set_crs(4326) %>%
+#   mutate(
+#     geometry_lagged = lag(geometry, default = empty)
+#   ) %>%
+#   # drop the NA row created by lagging
+#   slice(-1) %>%
+#   mutate(
+#     line = st_sfc(purrr::map2(
+#       .x = geometry,
+#       .y = geometry_lagged,
+#       .f = ~{st_union(c(.x, .y)) %>% st_cast("LINESTRING")}
+#     ))) -> track_lines
+
+cast_latlon_to_step_ls <- function(.df, .crs = 4326){
+
+  df <- .df %>%
+    cast_latlon_to_pts(.drop_coord = FALSE)
+
+  empty <- sf::st_as_sfc("POINT(EMPTY)", crs = .crs)
+
+  steps <- df %>%
+    dplyr::mutate(prev_geometry = dplyr::lag(geometry, default = empty)) %>%
+    # drop the NA row created by lagging
+    dplyr::slice(-1) %>%
+    dplyr::mutate(
+      line = st_sfc(purrr::map2(
+              .x = geometry,
+              .y = prev_geometry,
+              .f = ~{
+                  sf::st_union(c(.x, .y)) %>%
+                  sf::st_cast("LINESTRING")
+                }
+              )
+              )
+    )
+}

### 2023-02-28 R/cast_sf_utils.R

@@ -1,0 +1,65 @@
+#' Utility functions to cast lat/lon data frame/tibble to sf-points and/or linestring
+#'
+#' @return
+NULL
+
+#' @rdname cast_latlon_to_pts
+#'
+#' @param .df dataframe/tibble of lat/lon positions
+#' @param .crs coordinate reference system (default: 4326 := WGS84)
+#' @param .drop_coord remove (or keep) lat/lon columns (default: TRUE := remove, FALSE := keep)
+#'
+#' @return
+#' @export
+#'
+#' @examples
+#' \donotrun{
+#' cast_latlon_to_pts(adsb_df)
+#' }
+cast_latlon_to_pts <- function(.df, .crs = 4326, .drop_coord = TRUE){
+  pts_sf <- .df %>%
+    sf::st_as_sf(coords = c("LON","LAT"), crs = .crs, remove = .drop_coord)
+  return(pts_sf)
+}
+
+#' @rdname cast_pts_to_ls
+#'
+#' @return sf linestring
+#'
+#' @export
+cast_pts_to_ls <- function(.pts_sf, .group_var){
+  ls_sf <- .pts_sf %>%
+    dplyr::group_by({{ .group_var}}) %>%
+    dplyr::summarise(do_union = FALSE, .groups = "drop") %>%
+    sf::st_cast("LINESTRING")
+  return(ls_sf)
+}
+
+#' @rdname cast_latlon_to_ls
+#'
+#' @return sf linestring
+#' @export
+cast_latlon_to_ls <- function(.df, .crs = 4326, .drop_coord = TRUE, ...){
+  pts_sf <- cast_latlon_to_pts(.df, .crs, .drop_coord)
+  ls_sf  <- cast_pts_to_ls(pts_sf, .group_var = NULL)
+}
+
+#' @rdname cast_point_to_latlon
+#'
+#' @param .crs
+#' @param ...
+#' @param .df_pts
+#' @param .drop_geometry
+#'
+#' @return df with LAT/LON coordinates
+#' @export
+cast_pts_to_latlon <- function(.df_pts, lat_var=LAT, lon_var=LON, .crs = 4326, .drop_geometry = TRUE, ...){
+  df <- .df_pts %>%
+    dplyr::mutate({{lon_var}} := sf::st_coordinates(df)[,1],
+                  {{lat_var}} := sf::st_coordinates(df)[,2])
+
+  if(isTRUE(.drop_geometry)){
+    df <- df %>% sf::st_drop_geometry()
+  }
+
+}

### 2023-02-28 R/coerce_unixepoch_to_datetime.R

@@ -1,0 +1,13 @@
+#' Title
+#'
+#' @param .df 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+coerce_unixepoch_to_datetime <- function(.df, .time_var=TIME){
+  .df |> dplyr::mutate(
+    {{.time_var}} := lubridate::as_datetime({{.time_var}})
+  )
+}

### 2023-03-01 x_trj-apt-landing-association.qmd
more landing runway associated -- spotted wrong assignment due to ground coverage / not filtered out trajectory points ==> trajectory phase model!

@@
 ```{r}
-ggplot() +
+apt_arrival_plot <- ggplot() +
   geom_sf(data = arp_pt, size = 3, color = "blue") +
@@
   geom_sf(data = rwy_centerline_ls, color = "green") +
-  geom_sf_text(data = rwy_centerline_ls, aes(label = REF)) +
+  geom_sf_text(data = rwy_centerline_ls, aes(label = REF))
+
+apt_arrival_plot +
   geom_sf(data = baw236_landing_ls)
@@
 ```{r}
-# spatial join (aka "overlay" with "within")
+#' (Gu)estimate the arrival runway based on alignment of landing trajectory points and extended runway centerline
+#' 
+#'spatial join (aka "overlay" with "within")
 # to identify most likely ~ closest runway (extended centerline-string)
+#'
+#' @param .trj_pts 
+#' @param .rwy_ctrl_ls 
+#' @param .ctrl_offset 
+#'
+#' @return
+#' @export
+#'
+#' @examples
 estimate_rwy <- function(.trj_pts, .rwy_ctrl_ls, .ctrl_offset = 1000){
@@
   st_drop_geometry() %>%        # keep just the "spatially joined" dataframe
-  na.omit() %>%                 # eliminate no hits
+ # na.omit() %>%                 # eliminate no hits
   group_by(REF) %>%
@@
 }
+
+#' Assign most likely arrival landing
+#'
+#' @param .rwy_hit_score 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+#' #' \dontrun{
+#' }
+arrival_runway_assignment <- function(.rwy_hit_score){
+  .rwy_hit_score |> 
+    dplyr::arrange(desc(N)) |> 
+    dplyr::mutate(
+      COMMENT = paste0("landing runway: ", REF," trust-Level (hits: ", N, "): ", TRUST)) |> 
+    dplyr::filter(row_number() == 1)
+}
 ```
@@
 ```{r}
-baw236_final_approach <- baw236 |> 
+final_approach_trj <- baw236 |> 
   trrrj::filter_positions_at_range(c(arp_egll$LON, arp_egll$LAT), 0, 10, lon = LON, lat = LAT)
-# baw236_landing_ls <- baw236_landing |> cast_latlon_to_ls(.drop_coord = FALSE, .group_var = ICAO24)
-ggplot() +
-  geom_sf(data = arp_pt, size = 3, color = "blue") +
-  geom_sf(data = rwys_ls) +
-  geom_sf(data = rwys_pts, color = "red") +
-  geom_sf(data = rwy_centerline_ls, color = "green") +
-  geom_sf_text(data = rwy_centerline_ls, aes(label = REF)) +
-  geom_sf(data = baw236_final_approach |> cast_latlon_to_ls())
+
+final_approach_trj_pts <- final_approach_trj |> cast_latlon_to_pts()
+
+landing_runway <- estimate_rwy(final_approach_trj_pts, rwy_centerline_ls) |> 
+  arrival_runway_assignment()
+landing_runway <- landing_runway |> mutate(LAT = 51.5, LON = -0.45)
+
+apt_arrival_plot +
+  geom_sf(data = final_approach_trj_pts) +
+  labs(subtitle = landing_runway$COMMENT)
+```
+production pipeline
+
+TO----------------DOOOOOOOOOOOOOOOOOOOOOOOOOOOOO
+
+* multiple "landing trajectories"
+* iterate over landing trajectories
+
+```{r}
+final_trjs <- baws |> 
+  trrrj::filter_positions_at_range(c(arp_egll$LON, arp_egll$LAT), 0, 10, lon = LON, lat = LAT)
+
+final_trjs_pts <- final_trjs |> cast_latlon_to_pts()
+
+ldg_rwy <- final_trjs_pts |> group_by(ICAO24) |> 
+  group_modify(.f = ~ estimate_rwy(.x, rwy_centerline_ls) |> arrival_runway_assignment())
+ldg_rwy
+```
+
+WRONG RUNWAY ASSIGNMENT as filtering does not assure LAT/LONs on final!
+
+```{r}
+picks <-  c("400615")
+
+apt_arrival_plot +
+  geom_sf(data = final_trjs_pts |> filter(ICAO24 %in% picks))
 ```
@@
 ```{r}
-estimate_rwy(rwy_centerline_ls, baw236_final_approach |> cast_latlon_to_pts())
+baw74_landing <- baws |> filter(ICAO24 == "400615") |>  
+  trrrj::filter_positions_at_range(c(arp_egll$LON, arp_egll$LAT), 0, 20, lon = LON, lat = LAT)
+baw74_landing_ls <- baw74_landing |> cast_latlon_to_ls(.drop_coord = FALSE, .group_var = ICAO24)
+
+apt_arrival_plot +
+  geom_sf(data = baw74_landing_ls)
 ```
 
-Try to understand why / how the closeness is calculated.
-Current implementation shows N=1 ?!? while ok, it should be multiple hits.
+NEED FOR TRAJECTORY SEGMENTATION / PHASE ASSOCIATION
+
+* trajectory = (time-)ordered set of 4D positions
+* check
+  + runway-end crossing  <-- arrival or departure
+  + airport box          <-- arrival or departure
+* other milestone
+  + gate/6NM final
+  + extended runway box (e.g. 15NM)
+  + other characteristic milestones
+* phase points
+  + top of decent (/ climb)
+  + ...

### 2023-03-02 paper.tex
trajectory classification

@@
 
-\hypertarget{open-data-for-operational-performance-monnitoring}{%
+\hypertarget{trajectory-flight-phase-segmentation-milestone}{%
+\subsubsection{Trajectory Flight Phase Segmentation \&
+Milestone}\label{trajectory-flight-phase-segmentation-milestone}}
+
+Different approaches exists to detect and describe aircraft flight
+phases, e.g.~recent machine learning algorithm
+\protect\hyperlink{ref-sun2017flightphase}{{[}1{]}}. This paper
+implements a heuristic approach with a focus on the detection of arrival
+traffic at the study airports.
+
+\hypertarget{landing-runway-identification}{%
+\subsubsection{Landing Runway
+Identification}\label{landing-runway-identification}}
+
+The identification of the landing direction is based on a simple
+geospatial heuristic. Conceptually, aircraft are aligned with the runway
+(centerline) before landing. An aircraft is assigned to a landing runway
+based on the closeness of its pre-landing positions to the extended
+runway centerline.
+
+\hypertarget{open-data-for-operational-performance-monitoring}{%
 \subsection{Open Data for Operational Performance
-Monnitoring}\label{open-data-for-operational-performance-monnitoring}}
+Monitoring}\label{open-data-for-operational-performance-monitoring}}
 
@@
 System, airport operators report movement data on a monthly basis
-\protect\hyperlink{ref-apdf_v1_2019}{{[}1{]}}.
+\protect\hyperlink{ref-apdf_v1_2019}{{[}2{]}}.
 
@@
 \hypertarget{acknowledgment-and-disclaimer}{%
-\section{Acknowledgment and
+\section*{Acknowledgment and
 Disclaimer}\label{acknowledgment-and-disclaimer}}
+\addcontentsline{toc}{section}{Acknowledgment and Disclaimer}
 
@@
 \begin{CSLReferences}{0}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-sun2017flightphase}{}}%
+\CSLLeftMargin{{[}1{]} }%
+\CSLRightInline{J. Sun, J. Ellerbroek, and J. Hoekstra, {``Flight
+extraction and phase identification for large automatic dependent
+surveillance\textendash broadcast datasets,''} \emph{Journal of
+Aerospace Information Systems}, vol. 14, no. 10, pp. 566--572, 2017. }
+
 \leavevmode\vadjust pre{\hypertarget{ref-apdf_v1_2019}{}}%
-\CSLLeftMargin{{[}1{]} }%
+\CSLLeftMargin{{[}2{]} }%
 \CSLRightInline{EUROCONTROL, {``Eurocontrol {Specification} for

### 2023-03-02 x_trj-classification-segmentation.qmd
trajectory classification

---
title: "Trajectory Classification"
format: html
---

```{r setup}
# setup
library(tidyverse)
library(sf)

source(here::here("R", "level_segments_check.r"))

# set default ggplot theme
ggplot2::theme_set(ggplot2::theme_minimal())
```

DEVELOPMENT VERION

# Overview

We conceptualise

* a **trajectory** as the time-ordered sequence of 4D positions of a flight;
* a **flight** is an air transport movement from a departure aerodrome to its destination aerodrome;
* flights could include segments with specific operations (e.g. military flights: military exercise area, refueling, all flights: flight maneuvers, holding, low approach/touch-and-go)
* outside of these specific operations, flights progress through phases: (i) parking, (ii) taxi-out, including take-off roll, (iii) initial climb-out, (iv) climb-out, (v) en-route, (vi) decent, (vii) final approach, (viii) taxi-in to parking position.

download sample = state vector data (4D positions)

==> split flights into DA(I)O ~ ODA

```{mermaid}
%%| fig-width: 6
flowchart LR
  OSN[(Opensky \nNetwork)] 
  OSN --> A[Data \nExtraction] 
  A --> FP(Flight Phases)
  A --> B(Arrival \nTrajectories)
  FP -.- B
```

# Model Development

Load test data set (BAWs @ EGLL)

```{r}
tst_fn  <- "tst-egll-baw.csv.gz"
tst_pth <- here::here("data", tst_fn)
baws <- read_csv(tst_pth, show_col_types = FALSE)

if(class(baws$TIME) == "numeric") baws <- baws |> coerce_unixepoch_to_datetime()

# clean trajectory
baws <- baws |> Alt_Outlier_filter(.alt_var = ALT_B)
```

TODO = outlier filter works well when airborne, ground jumps not picked up

```{r}
baws |> 
ggplot() +
  geom_point(aes(x = TIME, y = ALT_B, color = FLTID))
```

## Flight Table and Milestone Table

A flight can be characterised by a series of milestones.    
Specific milestones (and supporting data) can be extracted in form of a flight table.

Unique flight identification:

An aircraft (identifiable by its ICAO24 address) can perform multiple flights on a given day.
Flights may enter/leave the coverage area.
Conceptually - and within the studied context of flights operating at European airports / night flying restrictions - the associated of the date of flight and ICAO24 address including the FLTID can serve as a unique key.

Milestone table allows for combining flights on day n and n+1, as characteristic milestones and their time offset allow for an association (and subsequent cleaning of the UID - not part of this study).

### Identify "flights"

On the basic level, flights are detected between START (first seen) and END (last seen) of an ICAO24 - typically - using a single FLTID

```{r}
identify_legs <- function(.trjs, .uid = ICAO24, .max_step = 300){
  .trjs |> 
    dplyr::group_by({{.uid}}) |> 
    dplyr::arrange(TIME) |> 
    dplyr::mutate(
       STEP = difftime(TIME, lag(TIME, default = dplyr::first(TIME))) |> as.numeric()
      ,LEG  = cumsum(STEP > .max_step)
      ,N_ID = length(unique(FLTID))
      ) |> 
    ungroup()
} 
make_uid <- function(.trjs_legs){
  .trjs_legs |> 
    dplyr::mutate(
       DOF = lubridate::date(TIME) 
      ,UID = paste(ICAO24, DOF, LEG, sep = "-")
      ) |> 
    dplyr::select(UID, ICAO24, TIME, LAT, LON, ALT_B, FLTID, everything())
}

init_flight_table <- function(.trjs_uid){
  this_flt <- .trjs_uid  |> 
    dplyr::summarise(
       UID = unique(UID)
      ,FLTID = unique(FLTID) 
      ,N   = n()
      ,FIRST = min(TIME), LAST = max(TIME), DUR = difftime(LAST, FIRST, units = "min")
      ,.by = UID
      ) 
}

init_milestone_table <- function(.trj_uid){
  this_mst <- .trj_uid  |> 
    dplyr::slice(c(1,n())) |> 
    dplyr::mutate(MST = c("START","STOP")) |> 
    dplyr::select(ICAO24, FLTID, TIME, LAT, LON, ALT_B, MST)
}
```

```{r}
mst <- baws |> identify_legs() |> make_uid() |> 
  group_by(UID) |> 
  group_modify(.f = ~ init_milestone_table(.x))
```

DETECT LEVEL SEGMENTS
TOC / TOD

```{r}
baws |> filter(FLTID == "BAW236") |> check_level_flight(.alt_var = ALT_B) |> ggplot() + geom_point(aes(x = TIME, y = ALT_B, color = as.factor(LEVEL)))
```

DETECT AIRPORT BOX

load airport environment

```{r}
# load EGLL environment
rwys_egll <- read_csv("./data/rwys-EGLL.csv", show_col_types = FALSE)
arp_egll  <- read_csv("./data/aip.csv", show_col_types = FALSE) |>
  filter(ICAO == "EGLL", REF == "OSN-center")

rwys_egll_pts <- rwys_egll |> cast_latlon_to_pts()
rwys_egll_pts_buf <- rwys_egll_pts |> sf::st_buffer(dist = 500)
rwys_egll_box <- rwys_egll_pts_buf |> sf::st_union() |> sf::st_convex_hull()
```

```{r}
ggplot() +
  geom_sf(data = rwys_egll_pts) +
  geom_sf(data = rwys_egll_pts_buf, color = "blue", fill = NA) +
  geom_sf(data = rwys_egll_box, color = "lightgreen", fill = NA)
```

### 2023-03-02 R/altitude_outlier_filter.R

@@ -1,0 +1,32 @@
+#' Smooth altitude
+#' 
+#' Kudos to Sam
+#'
+#' @param df 
+#' @param KernelSize 
+#' @param Fill 
+#' @param IntResults 
+#' @param .id 
+#' @param .alt_var 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+Alt_Outlier_filter=function(df, KernelSize = 5, Fill = TRUE, IntResults = FALSE, .id = ICAO24, .alt_var=FL_B){
+  df1 = dplyr::group_by(df, {{.id}}) %>%
+    dplyr::mutate(
+      ALT_med  = zoo::rollmedian({{.alt_var}}, KernelSize, fill=NA),
+      sq_eps   = ({{.alt_var}}-ALT_med)^2,
+      sigma    = sqrt(zoo::rollmean(sq_eps, KernelSize, fill=NA)),
+      Outlier   = ifelse(sq_eps>sigma, 1, 0))
+  # Choose if the outliers need to be filled by the median altitude
+  if (Fill==TRUE) {
+    df1 = dplyr::mutate(df1, {{.alt_var}} := ifelse((Outlier==0 | is.na(Outlier)), {{.alt_var}}, ALT_med))
+  }
+  # Choose if the intermediate results need to be returned in the dataframe
+  if (IntResults==FALSE) {
+    df1=dplyr::select(df1, -ALT_med, -sq_eps, -sigma)
+  }
+  return(df1)
+}

### 2023-03-02 R/level_segments_check.r

@@ -1,0 +1,34 @@
+#' Check for level portions of flight
+#'
+#' @param .trj
+#' @param VV_threshold
+#' @param Time_threshold
+#'
+#' @return
+#' @export
+#'
+#' @examples
+check_level_flight <- function(.trj
+                               , VV_threshold = 3/60   # FL/min, min = 60 seconds
+                               , Time_threshold= 20    # seconds := min length of level
+                               , .alt_var=FL
+){
+  Alt_threshold=VV_threshold*Time_threshold #FL
+  ## wrapper for Sam's CDO/CCO level identification
+  df <- .trj
+  df <- df %>%
+    mutate(SEQ_ID = row_number()) %>%
+    rename(FLIGHT_LEVEL = {{.alt_var}}, TIME_OVER = TIME) %>%
+
+    mutate(
+      LEVEL=ifelse(SEQ_ID,
+                   ifelse(abs((FLIGHT_LEVEL-lead(FLIGHT_LEVEL))/as.numeric(difftime(TIME_OVER, lead(TIME_OVER), units="secs")))<=VV_threshold, 1, 0),
+                   ifelse(is.na(ifelse(abs(FLIGHT_LEVEL-lead(FLIGHT_LEVEL))<=Alt_threshold, 1, 0)), 0,
+                          ifelse(abs(FLIGHT_LEVEL-lead(FLIGHT_LEVEL)) <= Alt_threshold, 1, 0))),
+      BEGIN=ifelse(LEVEL==1 & (lag(LEVEL)==0 | TIME_OVER==min(TIME_OVER)), 1, 0),
+      END=ifelse(LEVEL==0 & TIME_OVER!=min(TIME_OVER) & lag(LEVEL)==1, 1, 0)
+    )
+  ## re-wrap
+  df <- df %>% rename({{.alt_var}} := FLIGHT_LEVEL, TIME = TIME_OVER)
+
+}

### 2023-03-08 x_trj-classification-segmentation.qmd
heuristic for ARR or DEP assignmnt

@@
 ```{r}
-mst <- baws |> identify_legs() |> make_uid() |> 
+baws <- baws |> identify_legs() |> make_uid() 
+mst <- baws |> 
   group_by(UID) |> 
@@
 ```
+
+BAW74 - ICAO24 c("400615")
+
+checking first for overhead threshold
+
+```{r}
+fb <- baws |> filter(ICAO24 %in% c("400615")) |> cast_latlon_to_pts(.drop_coord = FALSE)
+
+over_thr <- fb[rwys_egll_pts_buf,]
+
+ggplot() +
+  geom_sf(data = rwys_egll_pts, size = 3) +
+  geom_sf(data = rwys_egll_pts_buf, color = "blue", fill = NA) +
+  geom_sf(data = over_thr, color = "yellow")
+```
+
+trajectory points that fall within buffered threshold polygons
+
+```{r}
+my_trj <- fb
+my_thres_buf <- rwys_egll_pts_buf
+
+check_thr_mst <- function(.trj, .thres_buf){
+  this_mst <- st_join(.trj, .thres_buf["REF"], left = FALSE) |> 
+  select(ICAO24, FLTID, TIME, LAT, LON, ALT_B, REF) |> 
+  mutate(MST = "THR") |> 
+  mutate(
+        LEAD = geometry[row_number() + 1]
+      , DIST = st_distance(geometry, LEAD, by_element = T)
+      , DIST = if_else(is.na(DIST), 0, as.numeric(DIST))   #correct for lead/NA
+      , CUM_DIST = cumsum(DIST)
+      , RLEID = consecutive_id(CUM_DIST)
+    ) |> 
+  sf::st_drop_geometry() |> 
+  filter(TIME == min(TIME), .by = RLEID)
+  return(this_mst)
+}
+
+thr_mst <- check_thr_mst(my_trj, my_thres_buf)
+thr_mst
+```
+
+```{r}
+baws |> filter(ICAO24 == "40073e") |> cast_latlon_to_pts(.drop_coord = FALSE) |> sf::st_crop(xmin = -1, xmax = 0, ymin = 51.2, ymax = 51.6) |>  ggplot() + geom_point(aes(x = LON, y = LAT)) + geom_sf(data = rwys_egll_pts_buf)
+```
+
+```{r}
+baw262 <- baws |> filter(ICAO24 == "40073e")
+baw262_ls <- baw262 |> cast_latlon_to_ls() |> sf::st_crop(xmin = -0.5, xmax = 0, ymin = 51.2, ymax = 51.6)
+
+ggplot() +
+  geom_sf(data = rwys_egll_box) +
+  geom_sf(data = baw262_ls)
+
+```
+
+```{r}
+box_poly_sf  <- rwys_egll_box |> sf::st_as_sf() |> mutate(REF = "EGLL-box") |> rename(geometry = x)
+this_trj_pts <- baw262 |> cast_latlon_to_pts(.drop_coord = FALSE)
+
+# test with one
+baw262 <- sf::st_join(this_trj_pts, box_poly_sf)
+
+key_mst <- baw262 |> filter(MST %in% c("START","END")) |> 
+  select(UID, ICAO24, TIME, FLTID, LAT, LON, ALT_B, MST, REF)
+key_mst
+```

### 2023-03-20 paper-abstract.qmd
get some stuff to github

---
title: "abstract"
format: pdf
---

Partial abstract ... taken from EDAS

The years 2019 through 2022 mark an unprecedented period for air transportation. Following peak congestion phenomena in 2019, the impact of COVID-19 resulted in the grounding of the global fleet in 2020 as governments aimed to contain the further spread of the corona virus. Within Europe, the recovery of air transportation in 2022 was further impacted by the repercussion of the war in Ukraine and the overall economic fallout in terms of cost-of-living inflation and fuel prices.

Across the industry the call to build back is accompanied by the need to address the challenges of a continued growth of aviation. Emerging political goals calls to build back "greener" and ensure a highly efficient air navigation system. Arrival operations at airports form part of a substantial benefit pool by managing planning uncertainty, actual travel times, and limited runway system throughput capacity. This paper considers a data-driven approach to monitor the operational performance of arrival management in Europe.

One of the key issues identified is the balance between utilisation of the runway system capacity and associated traffic separation and synchronisation within the arrival phase. The success of avoiding air traffic control interventions within the terminal airspace diminishes without utilising upstream synchronisation of arrival traffic. Experience from increasing additional times in the terminal airspace during the COVID recovery suggest that there is a benefit pool. However, monitoring of the efficiency of arrival management with the concept of additional time provides not for the required level of transparency in support of tapping into this benefit pool.

This paper introduces a spatial-temporal framework for arrival management by analysing flight trajectories. As assessing the efficiency of arrival management includes the extension of the analysis horizon beyond the terminal airspace. The trajectory-based approach overcomes the limitations of today's operational performance monitoring metrics and allows for the inclusion of meteorological or airspace structure related impacts, and support to evaluate the operated sequencing techniques/concepts.

This paper addresses the challenges of establishing a data-driven trajectory-analytical approach to characterising and monitoring the performance of arrival management techniques. The approach is detailed and studied on the basis of open trajectory data for a subset of airports with different arrival concepts (e.g. runway pressure/time-based separation on final, point-merge, extended arrival management) in Europe and the United States. We describe the concept development and modelling work. The validation of the approach has been tested based on historic open trajectory data. The utility of the measures is further demonstrated on current data showing how the measures can help to evaluate the performance during the recovery from COVID low traffic/demand scenario.

The results obtained indicate the general feasibility of the approach and its application within the European context. The operational performance measures obtained support a richer discussion of efficiency increasing improvements and concepts. The ability to provide spatial and temporal measures based on 4D trajectories is an enabler to assess and compare the benefits of specific arrival management techniques. This allows strategic planers and decision-makers to select from a wider range of such techniques and options. The latter will also impact the continual strive for higher operational efficiency to achieve and further reduce the environmental impact of air navigation.

The work presented serves as a building block to develop a data-driven approach to measuring performance in the arrival phase and addressing discussions on the utility of arrival management techniques at different airports. The results of t

### 2023-03-20 paper.qmd
trajectory classification
get some stuff to github

@@
 
+Operational efficiency is a key element of addressing aviation's contribution to climate change.
+<!-- verbatim copy --> Emissions may increase due to the expected growth in international air traffic until lower emitting technologies and fuels and other mitigating measures are developed and deployed ..
+<!-- end copy --> ICAO adopted at its 41st Assembly a long-term aspirational goal for international aviation [@icao2022].
+In support to the Paris Agreement, the goal is to achieve net-zero carbon emissions by 2050.
+
+Levers for fuel reduction:
+
+\* operational efficiency
+
+\* market-based measures
+
+\* sustainable aviation fuel
+
+\* new aircraft propulsion and airframes
+
+A wider use and pick-up of sustainable aviation fuel, and new aircraft propulsion technologies or aircraft design requires further research.
+
+Despite the introduction of an initial market-based mechanism, immediate action to curb fuel burn and CO2 emissions rests with improvements of operational efficiency.
+
 The contribution of this paper comprise:
@@
 ## header 2
-
-some cool text.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
 
@@
 
+### Trajectory Flight Phase Segmentation & Milestone
+
+Different approaches exists to detect and describe aircraft flight phases, e.g. recent machine learning algorithm [@sun2017flightphase].
+This paper implements a heuristic approach with a focus on the detection of arrival traffic at the study airports.
+
+### Landing Runway Identification
+
 The identification of the landing direction is based on a simple geospatial heuristic.
@@
 
-## Open Data for Operational Performance Monitoring
-
 ## Arrival Sequencing
 
-sequence separation
+Spacing deviation
+
+Let us consider a pair of consecutive landing aircraft denoted leader and trailer, with s their required time spacing1.
+Using the constant time delay principle, the spacing deviation (or spacing error) at time t considers the current position of trailer at time t, and the past position of leader at time t -- s.
+Precisely, it is defined as the difference between the respective minimum times from these two positions (see figure below): spacing deviation (t) = min time (trailer (t)) -- min time (leader (t -- s))
 

### 2023-03-20 x_bra-check.qmd
get some stuff to github

---
title: "Brazil"
format: html
---

```{r}
library(tidyverse)
```

Unfortunately https://aisweb.decea.mil.br/ was down (20.03.).    
Google search offered an AIP pdf from 2017 (hope that no major changes took place - check with DECEA).

Get airport information from AIP

```{r}
knitr::include_graphics("./figures/aip-snapshot-SBGR.png")
```

use measurements package to calculate decimal degrees (rounded to 6 digits)
measurements::conv_unit("046 28 23", "deg_min_sec", "dec_deg")

SBGR: ARP 232608S / 0462823W

bboxfinder - box lng/lat - min/max
bounding box SBGR: -46.500975,-23.460242,-46.435372,-23.416276

```{r}
rwys_sbgr <- tribble(
   ~ICAO  , ~REF , ~LAT      , ~LON      , ~NAME    , ~ELEV
  , "SBGR", "ARP", -23.435555, -46.473056, "ARP"    , 2461
  , "SBGR", "O9L", -23.434281, -46.483361, "09L/27R", 2444
  , "SBGR", "27R", -23.424894, -46.448619, "09L/27R", 2440
  , "SBGR", "09R", -23.438797, -46.487027, "09R/27L", 2450
  , "SBGR", "27L", -23.431189, -46.458855, "09R/27L", 2445
)

#write_csv(rwys_sbgr, "./data/rwys-SBGR.csv")
```

```{r}
#sbgr_arp   <- c(rwys_sbgr$LON, rwys_sbgr$LAT)
sbgr_200NM <- trrrj::polygon_at_distance(c(8.570556, 50.03333), d = 205)
```

## Load OSN data - Monday dump

```{r}
# load helper scripts to download and store OSN data
source("./R/osn_dumps.R")
source("./R/osn_read_tardump.R")
source("./R/osn_make_nice_names.R")
```
```{r}
#| message: false
pth_to_dumps <- "../__DATA/OSN-MondayDumps"
ptn <- "2022-05-09"

# list downloaded dumps
fns <- list.files(path = pth_to_dumps, pattern = ptn, full.names = TRUE)

# load list of chosen dumps
bra  <- fns |> 
  purrr::map_dfr(.f = read_osn_tar_file) |> 
  make_nice_names_osn()
```

### 2023-03-20 x_data-crunching.qmd
get some stuff to github

---
title: "Data Preparation"
format: pdf
---

DEVELOPMENT VERSION

```{r}
#| message: false
library(tidyverse)
library(sf)

# set default ggplot theme
ggplot2::theme_set(ggplot2::theme_minimal())
```

## Overview

This Quarto-markdown document lists the data preparatory steps.   
The following is developed for London Heathrow.

1. Load OSN data
2. load target airport(s)

## Load OSN data

```{r}
source("./R/osn_dumps.R")
source("./R/osn_read_tardump.R")
source("./R/osn_make_nice_names.R")
source("./R/coerce_meter_to_feet.R")
```

```{r}
#| message: false
#| warning: false
pth_to_dumps <- "../__DATA/OSN-dumps"
ptn <- "2019-08-01"

# list downloaded dumps
fns <- list.files(path = pth_to_dumps, pattern = ptn, full.names = TRUE)

# load list of chosen dumps
ds  <- fns |> 
  purrr::map_dfr(.f = read_osn_tar_file) |> 
  make_nice_names_osn() |> 
  mutate(ALT_B = coerce_meter_to_feet(ALT_B))
```

## Load target airport

```{r}
source(here::here("R", "cast_sf_utils.R"))
source(here::here("R", "cast_sf_rwy.R"))

# load EGLL environment
rwys_egll <- read_csv("./data/rwys-EGLL.csv", show_col_types = FALSE)
arp_egll  <- read_csv("./data/aip.csv", show_col_types = FALSE) |>
  filter(ICAO == "EGLL", REF == "OSN-center")

# cast sf
arp_pt    <- arp_egll  |> cast_latlon_to_pts(.drop_coord = FALSE) 
rwys_pts  <- rwys_egll |> cast_latlon_to_pts()
rwys_ls   <- rwys_egll |> cast_rwy_ls()

rwy_centerline_ls <- cast_rwy_ctr_line_ls(rwys_egll)
rwy_box           <- airport_threshold_box(rwys_egll)
```

```{r}
apt_arrival_plot <- ggplot() +
  geom_sf(data = rq) +
  geom_sf(data = rwy_box) +
  geom_sf(data = arp_pt, size = 3, color = "blue") +
  geom_sf(data = rwys_ls) +
  geom_sf(data = rwys_pts, color = "red") +
  geom_sf(data = rwy_centerline_ls, color = "green") +
  geom_sf_text(data = rwy_centerline_ls, aes(label = REF)) 
apt_arrival_plot
```

apt_arrival_plot +
  geom_sf(data = baw236_landing_ls)

### 2023-03-20 x_trj-classification-segmentation.qmd
get some stuff to github

@@
 
-source(here::here("R", "level_segments_check.r"))
+source(here::here("R", "level_segments_check.R"))
+source(here::here("R", "altitude_outlier_filter.R"))
 
@@
 baw262 <- baws |> filter(ICAO24 == "40073e")
-baw262_ls <- baw262 |> cast_latlon_to_ls() |> sf::st_crop(xmin = -0.5, xmax = 0, ymin = 51.2, ymax = 51.6)
+baw262_ls <- baw262 |> 
+  cast_latlon_to_ls() |> 
+  sf::st_crop(xmin = -0.5, xmax = 0, ymin = 51.2, ymax = 51.6)
 
@@
   geom_sf(data = baw262_ls)
-
 ```
@@
 ```
+
+## Start data preparation
+
+work with dataset for a day
+
+```{r}
+source("./R/osn_dumps.R")
+source("./R/osn_read_tardump.R")
+source("./R/osn_make_nice_names.R")
+source("./R/coerce_meter_to_feet.R")
+```
+
+```{r}
+#| message: false
+pth_to_dumps <- "../__DATA/OSN-dumps"
+ptn <- "2019-08-01"
+
+# list downloaded dumps
+fns <- list.files(path = pth_to_dumps, pattern = ptn, full.names = TRUE)
+
+# load list of chosen dumps
+ds  <- fns |> 
+  purrr::map_dfr(.f = read_osn_tar_file) |> 
+  make_nice_names_osn() |> 
+  mutate(ALT_B = coerce_meter_to_feet(ALT_B))
+```
+
+* filter pts within airport box (join per day runs for ~ 5mins)
+
+```{r}
+# EGLL airport box
+box_sf  <- rwys_egll_box |> sf::st_as_sf() |> mutate(REF = "EGLL-box") |> rename(geometry = x)
+
+# spatial subsetting, check alt, get ICAO24
+# command x[y, ] is identical to x[y, , op = st_intersects]
+seen_at_airport <- function(.pts, .poly_sf, .alt_threshold = 1000, .drop_coord = FALSE,...) {
+  ds_pts  <- .pts |> cast_latlon_to_pts(.drop_coord = .drop_coord)
+  # spatial subsetting, check alt, get ICAO24
+  # command x[y, ] is identical to x[y, , op = st_intersects]
+  inbox   <- ds_pts[.poly_sf,] |> 
+    dplyr::filter(ALT_B < .alt_threshold)
+  return(inbox)
+}
+```
+
+```{r}
+at_egll <- ds |> seen_at_airport(box_sf) 
+at_egll <- at_egll |> mutate(SEEN = "AT EGLL") |> 
+  summarise(BOX_MIN)
+```
+
+at egll with is.na(FLTID)
+# A tibble: 214 × 1
+   ICAO24
+   <chr> 
+ 1 4951c7
+ 2 400941
+ 3 4ca13d
+ 4 4006b0
+ 5 3c65cd
+ 6 406a3e
+ 7 4b17fc
+ 8 406ae3
+ 9 4010dd
+10 4009da
+
+Load validation data set
+
+```{}
+egll_fact <- read_csv2("~/RProjects/__DATA/EGLL/EGLL_2019_FACT.csv")
+egll_fact <- egll_fact |> 
+  select(FLTID = AP_C_FLTID, REG = AP_C_REG, ADEP = ADEP_ICAO, ADES = ADES_ICAO
+         ,MVT_TIME = MVT_TIME_UTC, BLOCK_TIME = BLOCK_TIME_UTC, PHASE = SRC_PHASE
+         ,TYPE = ARCTYP, CLASS = AC_CLASS, RWY = AP_C_RWY, STND = AP_C_STND)
+write_csv(egll_fact, "./data/validation-EGLL-2019.csv.gz")
+```
+```{r}
+egll_fact <- read_csv("./data/validation-EGLL-2019.csv.gz", show_col_types = FALSE) |> 
+  mutate(across(.cols = contains("_TIME"), .fns = lubridate::dmy_hms))
+
+egll_20190801 <- egll_fact |> filter(lubridate::date(MVT_TIME) == lubridate::date(ptn)) |> 
+  mutate(SEEN_OSN = FLTID %in% at_egll$FLTID)
+```
+
+```{r}
+egll_20190801 |> summarise(MVTS = n(), .by = PHASE) |> 
+  ggplot() + 
+  geom_col( aes(x = PHASE, y = MVTS)) +
+  geom_text(aes(x = PHASE, y = MVTS, label = MVTS), vjust = 1.5, color = "white") +
+  labs(title = "EGLL - movement count - 20190801", x = NULL, y = NULL)
+```

### 2023-03-20 R/cast_sf_rwy.R

@@ -1,0 +1,81 @@
+# RUNWAY UTILITY FUNCTIONS
+
+# setup sf objects
+#arp_pt   <- arp_egll |> cast_latlon_to_pts()
+#rwys_pts <- rwys_egll |> cast_latlon_to_pts()
+
+# utility function to coerce data frame of runway points to sf linestring
+cast_rwy_ls <- function(.rwys_df, .rwy_name = NAME){
+  rwys_ls  <- .rwys_df |> 
+    dplyr::group_by({{.rwy_name}}) |>
+    dplyr::group_modify(.f = ~ cast_latlon_to_ls(.) ) |> 
+    sf::st_as_sf()
+  return(rwys_ls)
+}
+
+# utility function to create centerline
+rwy_ctr_line <- function(.rwy_df = aip, .ctrl_length = 10000){
+  df <- .rwy_df %>% 
+    filter(REF != "ARP") %>% 
+    select(REF, NAME, LAT, LON) %>% 
+    group_by(NAME) %>% 
+    mutate( LAT2 = case_when(!is.na(lag(LAT))  ~ lag(LAT)
+                             ,!is.na(lead(LAT)) ~ lead(LAT))
+            ,LON2 = case_when(!is.na(lag(LON))  ~ lag(LON)
+                              ,!is.na(lead(LON)) ~ lead(LON))
+    ) %>% 
+    # calculate "reverse" runway bearing with geosphere::bearingRhumb
+    mutate( RBRG  = geosphere::bearingRhumb(p1 = cbind(LON2, LAT2), p2 = cbind(LON, LAT))
+    )
+  
+  # determine "endpoint" of extended centerline at d = 10000 meters
+  tmp <- with(df, geosphere::destPointRhumb(cbind(LON, LAT), b= RBRG, d = .ctrl_length)) %>% 
+    as_tibble() %>% rename(LON3 = lon, LAT3 = lat)
+  
+  # combine and return
+  df <- df %>% bind_cols(tmp)
+  return(df)
+}
+
+# utility function to coerce extension points
+rwy_ctr_line_pts <- function(.rwy_ctr_line, .debug = FALSE){
+  # could not get pivot_longer work with multiple cols
+  tmp1 <- .rwy_ctr_line %>% select("REF":"LON")
+  
+  # include opposite runway threshold
+  tmp2 <- .rwy_ctr_line %>% select("REF","NAME", "LAT2","LON2") %>%
[…]

### 2023-03-20 R/coerce_meter_to_feet.R

@@ -1,0 +1,13 @@
+#' OSN mtrs to feet
+#'
+#' 1m = 100cm/(2.54cm/in)/(12in/ft) = 3.280839895ft
+#' 
+#' @param .altitude_m altitude measured in metres
+#'
+#' @return .altitude_ft altitude in ft
+#' @export
+#'
+#' @examples
+coerce_meter_to_feet <- function(.altitude_m){
+  altitude_ft <- .altitude_m * 3.280839895
+}

### 2023-03-22 x_bra-check.qmd
still playing around with pre-processing

@@
 library(tidyverse)
+library(patchwork)
 ```
@@
 ```{r}
-#sbgr_arp   <- c(rwys_sbgr$LON, rwys_sbgr$LAT)
-sbgr_200NM <- trrrj::polygon_at_distance(c(8.570556, 50.03333), d = 205)
+sbgr_arp   <- rwys_sbgr |> filter(ICAO == "SBGR", REF == "ARP") |> select(LAT, LON)
+sbgr_200NM <- trrrj::polygon_at_distance(c(sbgr_arp$LON, sbgr_arp$LAT), d = 205) |> 
+  sf::st_as_sf() |> 
+  rename(geometry = x) |> 
+  mutate(NAME = "SBGR-200")
+```
+
+```{r}
+sbgr_base_plot <- ggplot() +
+  geom_sf(data = sbgr_200NM) +
+  geom_point(data = sbgr_arp, aes(x = LON, y = LAT))
+sbgr_base_plot
 ```
@@
 ```
+
+## spatial subsetting for SBGR
+
+```{r}
+spatial_subset <- function(.payload, .outer, ...){
+  # check if payload is already a sf object
+  if(!is(.payload,"sf")){
+    .payload <- .payload |> cast_latlon_to_pts(.drop_coord = FALSE)
+  }
+  # spatial subsetting
+  trimmed_data <- .payload[.outer,]
+}
+
+trjs_pts <- spatial_subset(bra |> filter(!is.na(LAT)), sbgr_200NM)
+
+# trjs_pts |> sf::st_drop_geometry() |> 
+#   arrow::write_parquet("./data/bra-sbgr-20220509.parquet")
+```
+
+```{r}
+sbgr_base_plot + geom_sf(data = trjs_pts)
+```
+
+```{r}
+picks <- trjs_pts |> filter(FLTID %in% c("TAM3201", "GLO1612","TAM3172", "DLH506"))
+
+sbgr_base_plot + geom_sf(data = picks, aes(color = ALT_B))
+```
+
+```{r}
+trjs_pts |> filter(FLTID == "DLH506") |> ggplot() + geom_point(aes(x = TIME, y = ALT_B))
+```
+
+## Subsetting for KATL
+
+```{r}
+katl_arp   <- c(-84.4278640,33.6366996)
+katl_200NM <- trrrj::polygon_at_distance(katl_arp, d = 205) |> sf::st_as_sf() |> 
+  rename(geometry = x) |> 
+  mutate(NAME = "KATL-205")
+
+katl_pts <- spatial_subset(bra |> filter(!is.na(LAT)), katl_200NM)
+katl_pts |> sf::st_drop_geometry() |> arrow::write_parquet("./data/usa-katl-20220509.parquet")
+```
+
+```{}
+# just a black ball = fully covered
+ggplot() +
+  geom_sf(data = katl_200NM) +
+  geom_sf(data = katl_pts)
+```
+
+```{r}
+check_flight <- katl_pts |> filter(ICAO24 %in% c("a3b4b5"))
+
+p1 <- check_flight |> 
+  ggplot() +
+  geom_sf(data = katl_200NM) +
+  geom_point(aes(x = LON, y = LAT, color = ALT_B))
+
+p2 <- check_flight |> 
+  ggplot() +
+  geom_point(aes(x = TIME, y = ALT_B, color = ALT_B))
+
+p1 + p2 + plot_layout(guides = 'collect')
+```

### 2023-03-22 x_data-crunching.qmd
still playing around with pre-processing

@@
   make_nice_names_osn() |> 
-  mutate(ALT_B = coerce_meter_to_feet(ALT_B))
+  mutate(across(.cols = c("ALT_B","ALT_G"), .fns = coerce_meter_to_feet))
+```
+
+## some trajectory cleaning
+
+we study airborn part, i.e. from take-off to landing
+
+* departing flights: velocity > 50 m/s
+* landing flights cannot have altitude above 5000ft, velocity < 90
+
+```{r}
+ds  |> 
+  filter(ICAO24 == "3c6498") |> 
+  filter(velocity > 50) |> 
+  Alt_Outlier_filter(.alt_var = ALT_B) |> 
+  ggplot() + 
+    geom_point(aes(x = TIME, y = ALT_B, color = velocity < 90 & ALT_B > 5000)) + 
+    geom_point(aes(x = TIME, y = ALT_G), color = "green") +
+  theme(legend.position = "top")
 ```
@@
 rwy_box           <- airport_threshold_box(rwys_egll)
+airport_box       <- airport_centerline_box(rwys_egll)
 ```
+
+Check some of the functions and shapes
 
@@
 apt_arrival_plot <- ggplot() +
-  geom_sf(data = rq) +
-  geom_sf(data = rwy_box) +
+  geom_sf(data = airport_box) +
   geom_sf(data = arp_pt, size = 3, color = "blue") +
@@
 
-apt_arrival_plot +
-  geom_sf(data = baw236_landing_ls)
+Address[which(LeastNEmployees == max(LeastNEmployees))]
+
+```{r}
+icao24_in_polygon <- function(.ds, .poly, .id = ICAO24){
+  # determine bounding box for polygon
+  this_box <- sf::st_bbox(.poly)
+  # pre-filter with dplyr - cut bounding box
+  icao24s <- .ds |> 
+    dplyr::filter(
+       LAT > this_box$ymin, LAT < this_box$ymax
+      ,LON > this_box$xmin, LON < this_box$xmax, 
+    )
+  # get ids seen
+  icao24s <- icao24s |> 
+    dplyr::summarise(
+      N = n(), N_ID = length(unique(FLTID))
+      , MIN_TIME = min(TIME,na.rm = TRUE)
+      , MIN_TIME_ALT = ALT_B[which(TIME == MIN_TIME)]
+      , MAX_TIME = max(TIME, na.rm = TRUE)
+      , MAX_TIME_ALT = ALT_B[which(TIME == MAX_TIME)]
+      , MIN_ALT = min(ALT_B, na.rm = TRUE)
+      , MED_ALT = median(ALT_B, na.rm = TRUE)
+      , MAX_ALT = max(ALT_B, na.rm = TRUE)
+      , .by = {{.id}})
+}
+```
+
+```{r}
+seen_at_airport <- icao24_in_polygon(ds, airport_box)
+seen_at_airport
+```
+
+```{r}
+source("~/RProjects/paper-2023-ICNS/R/trj_identify_legs.R")
+source("~/RProjects/paper-2023-ICNS/R/coerce_unixepoch_to_datetime.R")
+source("~/RProjects/paper-2023-ICNS/R/trj_fix_fltid.R")
+source("~/RProjects/paper-2023-ICNS/R/add_distance_rowwise.R")
+source("~/RProjects/paper-2023-ICNS/R/trj_fix_altitude.R")
+source("~/RProjects/paper-2023-ICNS/R/heuristic_DAO.R")
+
+# helper function to check missing FLTID or gaps
+check_leg_fltid <- function(.trjs_with_legs, ...){
+  .trjs_with_legs |> 
+    dplyr::summarise(
+        N = n(), N_ID = length(unique(FLTID))
+      , IDs = paste(unique(FLTID), collapse = "-")
+      , NAs = sum(is.na(FLTID))
+      , .by = c("ICAO24","LEG")
+      )
+}
+```
+
+```{r}
+rq <- ds |> 
+  filter(ICAO24 %in% seen_at_airport$ICAO24) |> 
+  coerce_unixepoch_to_datetime() |> 
+  identify_trajectory_legs() |> 
+  Alt_Outlier_filter(.alt_var = ALT_B) |> 
+  fix_altitude_heuristic() |> 
+  mutate(UID = paste(ICAO24, LEG, sep = "-"))
+rq <- rq |> ungroup()
+
+# check one more time for airport
+seen_at_airport2 <- icao24_in_polygon(rq, airport_box, UID) |> 
+  heuristic_DAO()
+seen_at_airport2
+```
+
+```{r}
+viz_check_DAO <- function(.trj, .apt_box = airport_box){
+  plt <- .trj |> 
+    ggplot() + 
+      geom_sf(data = .apt_box) + 
+      geom_point(data = .trj, aes(x = LON, y = LAT, color = TIME)) + 
+      coord_sf(xlim = c(-1, 0), ylim = c(51.3, 51.6), expand = FALSE)
+  return(plt)
+}
+
+picks <- seen_at_airport2$UID[1:9]
+
+rq |> filter(UID %in% picks) |> viz_check_DAO() + facet_wrap(.~UID)
+```

### 2023-03-22 R/add_distance_rowwise.R

@@ -1,0 +1,43 @@
+#' Calculate distance between successive points in a dataframe of LAT/LON positions
+#'
+#' @param .trj
+#' @param unit
+#'
+#' @return
+#' @export
+#'
+#' @examples
+add_flight_distance <- function(.trj, unit = "NM"){
+  df <- .trj
+  
+  #-------------- units
+  # distHaversine defaults to radius of Earth in meters
+  r_NM = 3443.92
+  unit_radius = r_NM
+  #----------------------------------------------------
+  
+  df <- df %>%
+    dplyr::mutate(
+      LAT_PREV = lag(LAT, default = first(LAT))
+      ,LON_PREV = lag(LON, default = first(LON))
+      ,DIST_FLOWN = geosphere::distHaversine( cbind(LON,LAT)
+                                              ,cbind(LON_PREV, LAT_PREV)
+                                              , r = unit_radius)
+      ,DIST_FLOWN = cumsum(DIST_FLOWN)
+    ) %>%
+    # remove additional columns
+    select(-LAT_PREV, - LON_PREV)
+  return(df)
+}
+
+
+# =================== sf based solution ==========================
+# library(sf)
+# library(dplyr)
+#
+# df %>%
+#   group_by(gr) %>%
+#   mutate(
+#     lead = geometry[row_number() + 1],
+#     dist = st_distance(geometry, lead, by_element = T),
+#   )

### 2023-03-22 R/cast_sf_rwy.R

@@ -75,7 +75,17 @@ airport_threshold_box <- function(.rwys_df, .thr_buffer = 500, ...){
   rwys_pts_buf <- .rwys_df |> cast_latlon_to_pts() |> 
     sf::st_buffer(dist = .thr_buffer)
   # combine and add convex hull to encompass all thresholds
-  rwys_box <- rwys_pts_buf |> sf::st_union() |> sf::st_concave_hull()
+  rwys_box <- rwys_pts_buf |> sf::st_union() |> sf::st_convex_hull()
 
   return(rwys_box)
+}
+
+airport_centerline_box <- function(.rwys_df, .thr_buffer = 500, .ctrline_buffer = 500){
+  airport_box  <- airport_threshold_box(.rwys_df, .thr_buffer)
+  ctr_line_buf <- cast_rwy_ctr_line_ls(.rwys_df) |> sf::st_buffer(dist = .ctrline_buffer)
+  
+  combo_box    <- sf::st_union(airport_box, ctr_line_buf) |> 
+    sf::st_convex_hull() |> 
+    sf::st_union()
+  return(combo_box)
 }

### 2023-03-22 R/heuristic_DAO.R

@@ -1,0 +1,20 @@
+#' Title
+#'
+#' @param .seen_at_airport 
+#' @param ... 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+heuristic_DAO <- function(.seen_at_airport, ...){
+  DAO <- .seen_at_airport |> 
+    dplyr::mutate(
+      DAO = case_when(
+         MIN_TIME_ALT < MAX_TIME_ALT ~ "D"
+        ,MIN_TIME_ALT > MAX_TIME_ALT ~ "A"
+        ,MED_ALT > 5000 ~ "O"
+        ,TRUE ~ NA_character_
+      )
+      )
+}

### 2023-03-22 R/osn_dumps.R

@@ -5,10 +5,9 @@
 #'
 #' @examples
 #' \dontrun{}
-osn_list_dumps <- function(){
+osn_list_dumps <- function(.url = "https://opensky-network.org/datasets/states/"){
 
-  url <- "https://opensky-network.org/datasets/states/"
-
+  url <- .url
   # scrape page
   dump_page <- rvest::read_html(url)
 
@@ -25,9 +24,9 @@ osn_list_dumps <- function(){
 }
 
 
-construct_dump_download_links <- function(.date){
+construct_dump_download_links <- function(.date, .url = "https://opensky-network.org/datasets/states/"){
 
-  url <- "https://opensky-network.org/datasets/states/"
+  url <- .url    # todo check for web address, etc
   url_date <- paste0(
      url
     ,.date

### 2023-03-22 R/trj_fix_altitude.R

@@ -1,0 +1,28 @@
+#' Removing absurd altitudes when aircraft on ground
+#'
+#' @param .trj 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+fix_altitude_heuristic <- function(.trj){
+  fixed <- .trj |> 
+    dplyr::mutate(
+      ALT_B = dplyr::case_when(
+        # capture off-sets for landing/rolling/blocking-in aircraft
+          velocity < 90 & ALT_B > 5000 & !is.na(ALT_G) ~ ALT_G
+        # capture off-sets for departing aircraft
+        # or is.na(ALT_G)
+        # TODO
+      )
+    )
+  
+  # landing trajectories have no altitude information towards the end
+  # we need to probably identify standstill / gate-in
+  # heuristic we fill downwards in time
+  fixed <- fixed |> 
+    tidyr::fill(ALT_B, .direction = "down")
+  
+  return(fixed)
+}

### 2023-03-22 R/trj_fix_fltid.R

@@ -1,0 +1,27 @@
+#' Clean weird stuff in FLTIDs of OSN
+#'
+#' @param .trjs 
+#' @param .fltid default FLTID, set as appropriate
+#' @param ... not sure yet ... could be of help later
+#'
+#' @return corrected FLTIDs
+#' @export
+#'
+#' @examples
+fix_invalid_fltid <- function(.trjs, .fltid = FLTID,...){
+  fixed <- .trjs   # to-do ... check what you get
+  # remove fltids with space
+  fixed <- fixed |> 
+    dplyr::mutate(
+       FIXCHECK = stringr::str_count({{.fltid}}, pattern = "\\s+")
+      ,{{.fltid}} := ifelse(FIXCHECK == 1, NA_character_, {{.fltid}})
+      )
+  
+  # fixes should leave only valid FLTIDs and NAs
+  # NAs can be filled
+  fixed <- fixed |> 
+    tidyr::fill({{.fltid}}, .direction = "downup") |> 
+  # remove fixing label
+    dplyr::select(-FIXCHECK)
+}
+

### 2023-03-22 R/trj_identify_legs.R

@@ -1,0 +1,25 @@
+#' Identify legs / gaps in a trajectory
+#'
+#' Label legs (segments) of a trajectory (time ordered sequence of ICAO24)
+#' addresses.
+#'
+#' @param .trjs
+#' @param .grp_var
+#' @param .max_gap in minutes
+#'
+#' @return
+#' @export
+#'
+#' @examples
+identify_trajectory_legs <- function(.trjs, .grp_var=ICAO24, .max_gap=15){
+  df <- .trjs %>%
+    dplyr::group_by({{.grp_var}}) %>%
+    dplyr::arrange(TIME, .by_group = TRUE) %>%
+    dplyr::mutate(
+      SEQ_ID = dplyr::row_number()
+      , STEP = difftime(TIME, dplyr::lag(TIME, default = dplyr::first(TIME)), units = "sec")
+      #-- add grouping for trajectory segments
+      ,  LEG = cumsum( STEP >= .max_gap * 60)    ) %>%
+    ungroup()
+  return(df)
+}

### 2023-03-22 x_data-crunching.qmd
sort of clean trajectory for EGLL/day 1 and arrival identification

@@
   filter(velocity > 50) |> 
+  Alt_Outlier_filter(.alt_var = ALT_B) |> 
+  ggplot() + 
+    geom_point(aes(x = TIME, y = ALT_B, color = velocity < 90 & ALT_B > 5000)) + 
+    geom_point(aes(x = TIME, y = ALT_G), color = "green") +
+  theme(legend.position = "top")
+```
+
+```{r}
+ds  |> 
+  filter(ICAO24 == "3c6498") |> 
+  fix_altitude_heuristic() |> 
   Alt_Outlier_filter(.alt_var = ALT_B) |> 
@@
 rq <- ds |> 
+  # subset for ICAO24s seen at the study airport
   filter(ICAO24 %in% seen_at_airport$ICAO24) |> 
+  # human readable TIME
   coerce_unixepoch_to_datetime() |>
-  identify_trajectory_legs() |> 
+  # clean ALT and trim for airborne part with heuristic fix
   Alt_Outlier_filter(.alt_var = ALT_B) |> 
   fix_altitude_heuristic() |> 
+  # assign leg for each airborne segment
+  identify_trajectory_legs() |> 
+  # force single FLTID here to construct niche UIDs
+  group_by(ICAO24, LEG) |> fix_invalid_fltid() |> ungroup() |> 
+  # construct UID
   mutate(UID = paste(ICAO24, LEG, sep = "-"))
-rq <- rq |> ungroup()
 
@@
 ```
+
+per arrival trajectory (iterate, group modify)
+
+* subset to 200NM
+* initialise mst
+* assign landing runway & landing time
+* calculate characteristics
+  + time-to-go
+  + distance-to-go
+  + crossings (e.g for ASMA)
+
+* save out mst/enriched trajectory

### 2023-03-22 R/trj_fix_altitude.R

@@ -6,23 +6,18 @@
 #' @export
 #'
 #' @examples
-fix_altitude_heuristic <- function(.trj){
+fix_altitude_heuristic <- function(.trj, .v_to = 50){
   fixed <- .trj |> 
-    dplyr::mutate(
-      ALT_B = dplyr::case_when(
-        # capture off-sets for landing/rolling/blocking-in aircraft
-          velocity < 90 & ALT_B > 5000 & !is.na(ALT_G) ~ ALT_G
-        # capture off-sets for departing aircraft
-        # or is.na(ALT_G)
-        # TODO
-      )
-    )
+  # flights are airborne at velocity > 
+    dplyr::filter(velocity >= .v_to) |> 
+  # remove off-set for landing flights
+    dplyr::filter(!(velocity < 90 & ALT_B > 5000))
   
   # landing trajectories have no altitude information towards the end
   # we need to probably identify standstill / gate-in
   # heuristic we fill downwards in time
-  fixed <- fixed |> 
-    tidyr::fill(ALT_B, .direction = "down")
+  # fixed <- fixed |> 
+  #   tidyr::fill(ALT_B, .direction = "down")
   
   return(fixed)
 }

### 2023-03-22 R/trj_fix_fltid.R

@@ -13,6 +13,7 @@ fix_invalid_fltid <- function(.trjs, .fltid = FLTID,...){
   # remove fltids with space
   fixed <- fixed |> 
     dplyr::mutate(
+      # label and remove FLTIDs with empty space(s)
        FIXCHECK = stringr::str_count({{.fltid}}, pattern = "\\s+")
       ,{{.fltid}} := ifelse(FIXCHECK == 1, NA_character_, {{.fltid}})
       )

### 2023-03-22 R/trj_identify_legs.R

@@ -11,7 +11,7 @@
 #' @export
 #'
 #' @examples
-identify_trajectory_legs <- function(.trjs, .grp_var=ICAO24, .max_gap=15){
+identify_trajectory_legs <- function(.trjs, .grp_var=ICAO24, .max_gap=10){
   df <- .trjs %>%
     dplyr::group_by({{.grp_var}}) %>%
     dplyr::arrange(TIME, .by_group = TRUE) %>%

### 2023-03-23 x_data-crunching.qmd
initialise milestione table for arrivals

--- 2023-03-22 x_data-crunching.qmd
@@
 source("./R/coerce_meter_to_feet.R")
+source("./R/altitude_outlier_filter.R")
+source("./R/trj_fix_altitude.R")
+source("./R/spatial_subset_jormungand.R")
 ```
@@
 
+```{r}
+ds  |> 
+  filter(ICAO24 == "3c6498") |> 
+  fix_altitude_heuristic() |> 
+  Alt_Outlier_filter(.alt_var = ALT_B) |> 
+  ggplot() + 
+    geom_point(aes(x = TIME, y = ALT_B, color = velocity < 90 & ALT_B > 5000)) + 
+    geom_point(aes(x = TIME, y = ALT_G), color = "green") +
+  theme(legend.position = "top")
+```
+
 ## Load target airport
@@
 airport_box       <- airport_centerline_box(rwys_egll)
+
+# for later
+jormungand <- trrrj::polygon_at_distance(c(arp_egll$LON, arp_egll$LAT), d = 205) |> 
+  sf::st_as_sf() |> rename(geometry = x) |> mutate(NAME = "EGLL-205")
 ```
@@
 ```
-
-Address[which(LeastNEmployees == max(LeastNEmployees))]
 
@@
 rq <- ds |> 
+  # subset for ICAO24s seen at the study airport
   filter(ICAO24 %in% seen_at_airport$ICAO24) |> 
+  spatial_subset_jormungand(jormungand) |> 
+  # human readable TIME
   coerce_unixepoch_to_datetime() |>
-  identify_trajectory_legs() |> 
+  # clean ALT and trim for airborne part with heuristic fix
   Alt_Outlier_filter(.alt_var = ALT_B) |> 
   fix_altitude_heuristic() |> 
+  # assign leg for each airborne segment
+  identify_trajectory_legs() |> 
+  # force single FLTID here to construct niche UIDs
+  group_by(ICAO24, LEG) |> fix_invalid_fltid() |> ungroup() |> 
+  # construct UID
   mutate(UID = paste(ICAO24, LEG, sep = "-"))
-rq <- rq |> ungroup()
 
 # check one more time for airport
-seen_at_airport2 <- icao24_in_polygon(rq, airport_box, UID) |> 
+seen_at_airport2 <- icao24_in_polygon(rq |> sf::st_drop_geometry(), airport_box, UID) |> 
   heuristic_DAO()
@@
 ```
+
+per arrival trajectory (iterate, group modify)
+
+* subset to 200NM (done - jormungand)
+```{r}
+fig <- ggplot() + geom_sf(data = arr_trjs |> cast_pts_to_ls(.group_var = UID)) +
+  labs(subtitle = paste("EGLL arrivals", ptn, sep = "-")) +
+  theme(plot.title.position = "plot")
+#ggsave(plot = fig, filename = "./figures/EGLL-arrivals-single-day.png", dpi = 300, height = 5, width = 5)
+fig
+```
+
+* initialise mst (done - initialise milestone table function on filtered trajectories)
+* assign landing runway & landing time
+* calculate characteristics
+  + time-to-go
+  + distance-to-go
+  + crossings (e.g for ASMA)
+
+* save out mst/enriched trajectory
+
+```{r}
+source("~/RProjects/paper-2023-ICNS/R/milestone_functions.R")
+arr_trjs <- rq |> filter(UID %in% (seen_at_airport2 |> filter(DAO == "A") |> pull(UID)) )
+mst <- initialise_milestone_table(arr_trjs)
+mst
+```

### 2023-03-23 R/milestone_functions.R

@@ -1,0 +1,24 @@
+#' Set up milestone table for study
+#'
+#' @param .trjs set of trajectories UID & LAT/LON positions
+#' @param ... 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+initialise_milestone_table <- function(.trjs, ...){
+  # ensure time ordering
+  trjs <- .trjs |> 
+    dplyr::arrange(TIME, .by_group = TRUE)
+  # first and last trajectory point
+  start <- trjs |> dplyr::slice(1, .by = UID) |> 
+    dplyr::mutate(MST = "START")
+  stop  <- trjs |> dplyr::slice(n(), .by = UID) |> 
+    dplyr::mutate(MST = "STOP")
+  
+  mst <- dplyr::bind_rows(start, stop) |> 
+    dplyr::select(UID, TIME, FLTID, LAT, LON, ALT_B, MST) |> 
+    dplyr::arrange(UID, TIME)
+  return(mst)
+}

### 2023-03-23 R/spatial_subset_jormungand.R

@@ -1,0 +1,27 @@
+#' Subset trajectory LAT/LON positions
+#'
+#' @param .df_latlon 
+#' @param .outer_poly sf-object of outer rim 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+spatial_subset_jormungand <- function(.df_latlon, .outer_poly, ...){
+  # dplyr cut for bounding box is more efficient than sf one-by-one
+  # determine bounding box for polygon
+  this_box <- sf::st_bbox(.outer_poly)
+  # pre-filter with dplyr - cut bounding box
+  df <- .df_latlon |> 
+    dplyr::filter(
+       LAT > this_box$ymin, LAT < this_box$ymax
+      ,LON > this_box$xmin, LON < this_box$xmax, 
+    )
+  
+  # fine-tune filter with sf; spatial subsetting
+  has_geometry <- is(.df_latlon, "sf")
+  if(isFALSE(has_geometry)){
+    df <- df |> cast_latlon_to_pts(.drop_coord = FALSE)
+  }
+   df <- df[.outer_poly, ]
+}

### 2023-03-23 x_data-crunching.qmd
initial use of sfnetwork for blending thr point in linestring/sequence of lat-long nodes

@@
 library(sf)
+library(patchwork)
 
@@
 
+viz_check_vertical <- function(.trj){
+  plt <- .trj |> 
+    ggplot() +
+    geom_point(aes(x = TIME, y = ALT_B))
+  return(plt)
+}
+
 picks <- seen_at_airport2$UID[1:9]
@@
 ```
+
+landing runway assignment
+
+```{r}
+source("~/RProjects/paper-2023-ICNS/R/trj_landing_runway.R")
+
+# test for one
+arr_rwy <- arr_trjs |> filter(UID == "00b1f9-0") |> assign_landing_runway()
+arr_rwy
+
+arr_rwys <- arr_trjs |> assign_landing_runway()
+arr_rwys
+```
+
+```{r}
+arr_rwys |> ggplot() + geom_histogram(aes(x = TRUST), bindwidth = 2)
+```
+
+# check for some fringe cases
+
+```{r}
+fringe_picks <-  arr_rwys |> filter(between(TRUST, 0.4,0.7))
+
+viz_check_DAO(arr_trjs |> filter(UID %in% fringe_picks$UID)) + facet_wrap(.~ UID)
+```
+
+Potential to remove = no landing with TRUST < 0.7
+
+* assign landing runway & landing time
+
+do for one/some
+
+> arr_rwys
+# A tibble: 733 × 6
+# Groups:   UID [733]
+   UID      REF       N   TOT TRUST COMMENT                                                      
+   <chr>    <chr> <int> <int> <dbl> <chr>                                                        
+ 1 00b1f9-0 27L      13    16 0.812 landing runway: 27L trust-Level (hits: 13): 0.8125           
+ 2 01013c-0 27L      11    14 0.786 landing runway: 27L trust-Level (hits: 11): 0.785714285714286
+ 3 0101bb-0 27R      12    15 0.8   landing runway: 27R trust-Level (hits: 12): 0.8              
+ 4 0200cb-0 27R      12    15 0.8   landing runway: 27R trust-Level (hits: 12): 0.8              
+ 5 04014c-0 27L      12    15 0.8   landing runway: 27L trust-Level (hits: 12): 0.8              
+ 6 04c15b-0 27R      11    14 0.786 landing runway: 27R trust-Level (hits: 11): 0.785714285714286
+ 7 06a0e6-0 27R      12    15 0.8   landing runway: 27R trust-Level (hits: 12): 0.8              
+ 8 06a11f-0 27R      12    15 0.8   landing runway: 27R trust-Level (hits: 12): 0.8              
+ 9 06a144-0 27L      13    16 0.812 landing runway: 27L trust-Level (hits: 13): 0.8125           
+10 06a149-0 27L      13    16 0.812 landing runway: 27L trust-Level (hits: 13): 0.8125
+
+```{r}
+arr_trjs |> filter(UID %in% c("00b1f9-0","01013c-0","0101bb-0","0200cb-0")) |> 
+  viz_check_DAO() + facet_wrap(.~UID)
+```
+
+```{r}
+arr_trjs |> filter(UID %in% c("00b1f9-0")) |> 
+  ggplot(
+    aes(  x = TIME
+        , y = ALT_B
+        , color = between(ALT_B, arp_egll$ELEV + 100, arp_egll$ELEV + 1000)
+        )
+    ) + 
+  geom_point() + labs(color = "Approach")
+```
+
+```{r}
+source("~/RProjects/paper-2023-ICNS/R/cast_sf_network.R")
+
+arp  <- arp_egll
+rwys <- rwys_egll
+
+final <- arr_trjs |> filter(UID %in% c("00b1f9-0")) |>
+  filter(ALT_B < arp$ELEV + 2000)
+
+ldg_rwy <- final |> assign_landing_runway(.arp = arp) |> 
+  inner_join(rwys, by = "REF")
+my_thr  <- ldg_rwy |> select(UID, REF, LAT, LON, COMMENT) |> cast_latlon_to_pts(.drop_coord = FALSE)
+
+final_sfn <- final |> cast_latLon_to_sfnet()
+# blend crossing in network
+# c.f. https://luukvdmeer.github.io/sfnetworks/reference/st_network_blend.html
+# blend points with a tolerance.
+# tol = units::set_units(50, "m")
+final_sfn <- final_sfn %>%
+    sfnetworks::st_network_blend(my_thr |> select(REF, geometry)) #, tolerance = tol)
+
+viz_check_vertical(final) + viz_check_DAO(final)
+```
+
+check for blended point
+final_sfn |> activate("nodes") |> filter(is.na(TIME)) |> select(TIME, ICAO24, LAT, LON, ALT_B, geometry)
+
+point is injected, need to fill details, e.g. TIME
+then extract nodes ~ enriched sequence of final points (maybe use sequence number)
+
+NEXT-NEXT
+* calculate characteristics
+  + time-to-go
+  + distance-to-go
+  + crossings (e.g for ASMA)
+
+* save out mst/enriched trajectory

### 2023-03-23 R/trj_landing_runway.R

@@ -1,0 +1,72 @@
+#' (Gu)estimate the arrival runway based on alignment of landing trajectory points and extended runway centerline
+#' 
+#'spatial join (aka "overlay" with "within")
+# to identify most likely ~ closest runway (extended centerline-string)
+#'
+#' @param .trj_pts 
+#' @param .rwy_ctrl_ls 
+#' @param .ctrl_offset 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+estimate_rwy <- function(.trj_pts, .rwy_ctrl_ls, .ctrl_offset = 1000){
+  # stop if not
+  tmp <- st_join( .rwy_ctrl_ls, .trj_pts
+                  , join = st_is_within_distance, dist = .ctrl_offset) %>%
+    st_drop_geometry() %>%        # keep just the "spatially joined" dataframe
+    # na.omit() %>%                 # eliminate no hits
+    group_by(REF) %>%
+    summarise(N = n(), .groups = "drop") %>%
+    mutate( TOT   = sum(N, na.rm = TRUE)       # na.rm should not be necessary
+            ,TRUST = N / TOT)                   # share in case several runways
+  
+  return(tmp)
+}
+
+
+#' Assign most likely arrival landing
+#'
+#' @param .rwy_hit_score 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+#' #' \dontrun{
+#' }
+arrival_runway_assignment <- function(.rwy_hit_score){
+  .rwy_hit_score |> 
+    dplyr::arrange(desc(N)) |> 
+    dplyr::mutate(
+      COMMENT = paste0("landing runway: ", REF," trust-Level (hits: ", N, "): ", TRUST)) |> 
+    dplyr::filter(row_number() == 1)
+}
+
+
+#' Assign landing runway
+#'
+#' @param .arr_trjs 
+#' @param .arp 
+#' @param .group_var 
+#' @param ... 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+assign_landing_runway <- function(.arr_trjs, .arp = arp_egll , .group_var = UID ,...){
+  final_trjs <- .arr_trjs |>
+    # omit flights close to ARP ~ remove potential surface hits -> min = 2!
+    # find a better way to filter final 
+    trrrj::filter_positions_at_range(c(.arp$LON, .arp$LAT), 2, 10, lon = LON, lat = LAT)
+  
[…]

### 2023-03-24 x_data-crunching.qmd
playing around with landing time estimation

@@
 
+viz_check_vertical(final)
+```
+
+```{r}
+source("~/RProjects/paper-2023-ICNS/R/identify_threshold_overhead.R")
+source("~/RProjects/paper-2023-ICNS/R/add_bearing_to_target.R")
+
 ldg_rwy <- final |> assign_landing_runway(.arp = arp) |> 
   inner_join(rwys, by = "REF")
-my_thr  <- ldg_rwy |> select(UID, REF, LAT, LON, COMMENT) |> cast_latlon_to_pts(.drop_coord = FALSE)
+my_thr  <- ldg_rwy |> select(UID, REF, LAT, LON, COMMENT) |> 
+  cast_latlon_to_pts(.drop_coord = FALSE)
 
-final_sfn <- final |> cast_latLon_to_sfnet()
-# blend crossing in network
-# c.f. https://luukvdmeer.github.io/sfnetworks/reference/st_network_blend.html
-# blend points with a tolerance.
-# tol = units::set_units(50, "m")
-final_sfn <- final_sfn %>%
-    sfnetworks::st_network_blend(my_thr |> select(REF, geometry)) #, tolerance = tol)
+# check if we have a threshold overflight
+ovr_thr <- identify_threshold_overhead(final, my_thr)
+if(nrow(ovr_thr) > 0){
+  message("THRESHOLD")
+  eldt  <- ovr_thr |> filter(DIST_THR == min(DIST_THR))
+  final <- final |> mutate(MST_LABEL = "ALDT") 
+}else{
+  message("NEED TO WORK ON LANDING TIME")
+}
 
 viz_check_vertical(final) + viz_check_DAO(final)
+```
+
+Lets' work this out.
+
+```{r}
+# we have identified an arrival trajectory
+trj_arr <- arr_trjs |> filter(UID %in% "0200cb-0") |> 
+#trj_arr <- arr_trjs |> filter(UID %in% c("00b1f9-0")) |> 
+# trim payload and add sequence id
+  select(UID, TIME, ICAO24, LAT, LON, ALT_B)  |> 
+  rowid_to_column("NODE_ID") |> 
+  mutate(MST = NA_character_)
+
+ldg_rwy <- trj_arr |> assign_landing_runway(.arp = arp) |> 
+  inner_join(rwys, by = "REF")
+my_thr  <- ldg_rwy |> select(UID, REF, LAT, LON, COMMENT) |> 
+  cast_latlon_to_pts(.drop_coord = FALSE)
+
+# check if we have a threshold overflight
+ovr_thr <- identify_threshold_overhead(trj_arr, my_thr)
+if(nrow(ovr_thr) > 0){
+  message("THRESHOLD")
+  eldt    <- ovr_thr |> filter(DIST_THR == min(DIST_THR)) 
+  trj_arr[eldt$NODE_ID,"MST"] <- "ALDT"
+}else{
+  message("NEED TO WORK ON LANDING TIME")
+}
+```
+
+look at function ahead ... started adding threshold --> complement/fill columns & calc landing time
+
+```{r}
+viz_check_vertical(trj_arr)
 ```

### 2023-03-24 R/add_bearing_to_target.R

@@ -1,0 +1,30 @@
+#' Title
+#'
+#' @param .trj 
+#' @param target 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+trj_add_target_bearing <- function(.trj_latlon_sf, .target_latlon_sf ){
+  # define default NA for sf empty points when using lag/led
+  empty <- sf::st_as_sfc("POINT(EMPTY)", crs = 4326)
+
+  final <- .trj_latlon_sf |> 
+ 
+  # now calculate the data we need  
+    mutate(
+        THR      = sf::st_geometry(.target_latlon_sf)
+      , DIST_THR = sf::st_distance(geometry, THR, by_element = TRUE)
+      , distance_to_next = sf::st_distance(
+                geometry
+              , dplyr::lead(geometry, default = empty)
+              , by_element = TRUE)
+      , BRG = geosphere::bearingRhumb(
+                cbind(LON, LAT)
+              , cbind(.target_latlon_sf$LON, .target_latlon_sf$LAT)
+              )
+      )
+  return(final)
+}

### 2023-03-24 R/identify_threshold_overhead.R

@@ -1,0 +1,19 @@
+identify_threshold_overhead <- function(.final, .my_thr ,...){
+  trj <- .final |> 
+    trj_add_target_bearing(.my_thr) |> 
+    dplyr::mutate(
+       OVERHEAD = between(as.numeric(DIST_THR), -100, 100)
+      , BRG_CHG = ! abs(lag(BRG) - BRG) < 25
+      , OVR_THR = OVERHEAD & BRG_CHG) |> 
+    dplyr::filter(OVR_THR)
+}
+
+identify_threshold_ahead <- function(.final, .my_thr ,...){
+  trj <- .final |> 
+    dplyr::filter(ALT_B > arp$ELEV + 500) |> 
+    trj_add_target_bearing(.my_thr)
+  
+  # add threshold
+  thr <- select(.my_thr, LAT, LON)
+  trj <- bind_rows(trj, thr)
+}

### 2023-03-25 x_data-crunching.qmd
getting towards completing arrival trajectory extraction

@@
 ```{r}
+source("~/RProjects/paper-2023-ICNS/R/assign_threshold_time.R")
+
 # we have identified an arrival trajectory
@@
 #trj_arr <- arr_trjs |> filter(UID %in% c("00b1f9-0")) |> 
-# trim payload and add sequence id
-  select(UID, TIME, ICAO24, LAT, LON, ALT_B)  |> 
-  rowid_to_column("NODE_ID") |> 
-  mutate(MST = NA_character_)
-
-ldg_rwy <- trj_arr |> assign_landing_runway(.arp = arp) |> 
-  inner_join(rwys, by = "REF")
-my_thr  <- ldg_rwy |> select(UID, REF, LAT, LON, COMMENT) |> 
-  cast_latlon_to_pts(.drop_coord = FALSE)
-
-# check if we have a threshold overflight
-ovr_thr <- identify_threshold_overhead(trj_arr, my_thr)
-if(nrow(ovr_thr) > 0){
-  message("THRESHOLD")
-  eldt    <- ovr_thr |> filter(DIST_THR == min(DIST_THR)) 
-  trj_arr[eldt$NODE_ID,"MST"] <- "ALDT"
-}else{
-  message("NEED TO WORK ON LANDING TIME")
-}
+  assign_threshold_time(my_thr)
 ```
@@
 
-check for blended point
-final_sfn |> activate("nodes") |> filter(is.na(TIME)) |> select(TIME, ICAO24, LAT, LON, ALT_B, geometry)
+PICK SOME ARRIVALS FOR GETTING CODE IN PLACE
 
-point is injected, need to fill details, e.g. TIME
-then extract nodes ~ enriched sequence of final points (maybe use sequence number)
+```{r}
+# derived from validation data set, binned per 15 mins per arrival runway
+# 27L 06:30 - 0645
+peak_wave <- c("BAW377", "EWG6VP", "BAW11J", "ETD89G", "AIC161", "AFR82FE", "BAW236","DLH3WU","VIR138M","EWG30N")
+
+peak_trjs <- arr_trjs |> filter(FLTID %in% peak_wave)
+
+viz_check_DAO(peak_trjs) + viz_check_vertical(peak_trjs |> filter(TIME > lubridate::ymd_hm("2019-08-01 06:20")))
+```
+
+```{r}
+# run loop with group_modify()
+# create a new UID for the looping as group_modify cannot accept UID in payload
+# TODO - write functions independent of UID or handle UID outside calculation.
+study_trjs <- peak_trjs |> 
+  mutate(LUID = UID) |> group_by(LUID) |> 
+  group_modify(.f = ~assign_threshold_time(.x,.my_thr = my_thr)) |> 
+  ungroup() |> select(-LUID)
+```
+
+```{r}
+ggplot(data = study_trjs) +
+  geom_point(aes(x = TIME, y = ALT_B)) +
+  geom_vline(xintercept = study_trjs$TIME[which(study_trjs$MST == "ALDT")])
+```
+
+first pair
+
+```{r}
+# "BAW377","EWG6VP","BAW11J"
+inter_arrival <- study_trjs |> filter(MST == "ALDT") |> group_by(MST_REF) |> arrange(TIME, .by_group = TRUE) |> 
+  mutate(
+     FOLLOWER = lead(UID, default = NA_character_)
+    ,SEP_TIME = difftime(lead(TIME, default = NA), TIME, unit = "sec")
+    )
+inter_arrival
+```
+
+```{r}
+p1 <- inter_arrival |> ggplot() +
+  geom_step(aes(x = TIME, y = SEP_TIME))
+p2 <- inter_arrival |> ggplot() +
+  geom_boxplot(aes(x = MST_REF, y = SEP_TIME))
+p1 + p2
+```
+
+OK - FUCKED UP HERE 
+NEED to catch improper trajectories that throw error ==> improve cleaning
+wrap look in safely/possibly ... to take care of any other uncaught exception
+
+PRESS ON WITH PIPELINE ... data cleaning is on track ... should work for paper
+
+Do this for all of 1. August
+
+```{r}
+fn <- here::here("data","validation-EGLL-APDF-20190801.csv.gz")
+egll_arrs_apdf <- read_csv(fn, show_col_types = FALSE)
+
+# safe loop
+safe_assign_threshold_time <- possibly(.f = assign_threshold_time, otherwise = "Error")
+
+study_trjs <- arr_trjs |> 
+  filter(FLTID %in% egll_arrs_apdf$FLTID) |> 
+  mutate(LUID = UID) |> group_by(LUID) |> 
+  group_modify(.f = ~safe_assign_threshold_time(.x,.my_thr = my_thr)) |> 
+  ungroup() |> select(-LUID)
+```
 

### 2023-03-25 x_validation-EGLL.qmd
getting towards completing arrival trajectory extraction

---
title: "Untitled"
format: html
---

Check for landing sequence sample from APDF

```{r}
library(tidyverse)

# test sample 
ptn <- "2019-08-01"

apdf <- read_csv("./data/validation-EGLL-2019.csv.gz", show_col_types = FALSE)

arrs <- apdf |> filter(PHASE == "ARR") |> 
  mutate(across(.cols = contains("TIME"), .fns = lubridate::dmy_hms)) |> 
  filter(grepl(pattern = ptn, x = MVT_TIME)) |> 
  mutate(BIN = lubridate::floor_date(MVT_TIME, unit = "15 min"))
```
Runway usage

```{r}
arrs |> ggplot() +
  geom_point(aes(x = MVT_TIME, y = RWY), alpha = 0.3)
```

```{r}
tmp <- arrs |> 
  group_by(BIN, RWY) |> summarise(N = n())

tmp |> 
  ggplot() + geom_step(aes(x = BIN, y = N))
```

peak sessions 10 per 15 mins

```{r}
peak_bins <- tmp |> filter(N >=10)
peak_arrs <- arrs |> filter(BIN %in% peak_bins$BIN) |> arrange(RWY, MVT_TIME)
peak_arrs
```

peak_wave = peak_arrs$FLTID[1:10]

```{r}
fn <- here::here("data","validation-EGLL-APDF-201908-01.csv.gz")
write_csv(arrs, file = fn)
```

### 2023-03-25 R/assign_threshold_time.R

@@ -1,0 +1,49 @@
+#' Determine threshold time for arrival trajectories
+#'
+#' Work out homing and coasting of arrival trajectories. Identify overhead
+#' threshold or ahead of threshold in case of low coverage.
+#' Estimate time over threshold based on most recent position updates (and
+#' derived approach speed).
+#'
+#' @param .trj_latlon_sf 
+#' @param .my_thr 
+#' @param ... 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+assign_threshold_time <- function(.trj_latlon_sf, .my_thr, ...){
+  trj_arr <- .trj_latlon_sf
+  my_thr  <- .my_thr
+  # trim payload and add sequence id
+  trj_arr <- trj_arr |> select(UID, TIME, ICAO24, LAT, LON, ALT_B)  |> 
+    rowid_to_column("NODE_ID") |> 
+    mutate(MST = NA_character_)
+  
+  ldg_rwy <- trj_arr |> assign_landing_runway(.arp = arp) |> 
+    inner_join(rwys, by = "REF")
+  my_thr  <- ldg_rwy |> select(UID, REF, LAT, LON, COMMENT) |> 
+    cast_latlon_to_pts(.drop_coord = FALSE) 
+  
+  # check if we have a threshold overflight
+  ovr_thr <- identify_threshold_overhead(trj_arr, my_thr)
+  if(nrow(ovr_thr) > 0){
+    message("THRESHOLD")
+    thr_row <- ovr_thr |> filter(DIST_THR == min(DIST_THR)) 
+    trj_arr[thr_row$NODE_ID,"MST"] <- "ALDT"
+    trj_arr[thr_row$NODE_ID,"MST_REF"] <- my_thr$REF
+  }else{
+    message("NEED TO WORK ON LANDING TIME")
+    thr_row <- trj_arr |> 
+      identify_threshold_ahead(my_thr) |> 
+      estimate_threshold_time(my_thr)
+    thr_row$NODE_ID <- thr_row$NODE_ID +1       # increase node counter
+    thr_row$MST     <- "ALDT"
+    thr_row$MST_REF <- my_thr$REF
+    trj_arr <- trj_arr |> 
+      filter(NODE_ID < thr_row$NODE_ID)  |>     # truncate
+      bind_rows(thr_row)             
+  }
+  return(trj_arr <- trj_arr |> select(NODE_ID:MST, MST_REF))
+}

### 2023-03-25 R/estimate_threshold_time.R

@@ -1,0 +1,48 @@
+#' Estimation of time over threshold based on last positions before threshold
+#'
+#' @param .trjs_latlon_sf 
+#' @param .my_thr 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+estimate_threshold_time <- function(.trjs_latlon_sf, .my_thr){
+  
+  # decode runway identifier
+  RWY <- .my_thr$REF
+  rwy_to_number <- function(.rwy){
+    rwy_dir <- gsub("\\D","", .rwy) # strip non-digets and replace with empty
+    rwy_dir <- ((as.numeric(rwy_dir) * 10) + 360) %% 360
+  }
+  
+  last_pos <- .trjs_latlon_sf %>%
+    mutate(#DIST_THR = distHaversine(cbind(LON, LAT), cbind(LON_THR,LAT_THR))
+           #, BRG_THR = bearingRhumb( cbind(LON, LAT), cbind(LON_THR, LAT_THR))
+           #,
+             BEFORE  = abs(BRG - rwy_to_number(RWY)) < 30
+           , COASTG  = DIST_THR == lag(DIST_THR)
+           , V_GND   = (lag(DIST_THR) - DIST_THR) / as.numeric(TIME - lag(TIME))
+           , V_GND_SM= zoo::rollmean(V_GND, k = 9, fill = NA, align = "right")
+    ) 
+  
+   any_costing <- sum(last_pos$COASTG == TRUE, na.rm = TRUE) #; message(paste0("ONLY AHEAD: "), any_costing) 
+   # no costing
+   if(any_costing > 1){
+     last_pos <-  last_pos %>%
+    # cut-off any position after the start of coasting
+    # or if there are positions post threshold ~ BRG <> landing direction
+     filter(COASTG == FALSE & BEFORE == TRUE )
+   }
+  
+  # check length / duration of last positions
+  last_15 <- last_pos %>% tail(n = 15)
+  avg_final_app_speed <- mean(last_15$V_GND_SM, na.rm = TRUE) |> as.numeric()
+  avg_final_app_speed <- avg_final_app_speed * 0.9   # assume 10% for flare
+  
+  thr_row <- last_15 |> dplyr::slice_tail(n = 1) |> 
+    mutate( LAT = .my_thr$LAT, LON = .my_thr$LON
+           ,TIME = TIME + ceiling((DIST_THR |> as.numeric()) / avg_final_app_speed)  # round up for complete secs
+           ,DIST_THR = 0                                           # @ threshold!
+    )
+}

### 2023-03-25 R/identify_threshold_overhead.R

@@ -14,6 +14,9 @@ identify_threshold_ahead <- function(.final, .my_thr ,...){
     trj_add_target_bearing(.my_thr)
   
   # add threshold
-  thr <- select(.my_thr, LAT, LON)
-  trj <- bind_rows(trj, thr)
+  # thr <- select(.my_thr, LAT, LON)
+  # trj <- bind_rows(trj, thr) |>
+  #   dplyr::mutate(NODE_ID = dplyr::lag(NODE_ID) + 1) |> 
+  #   tidyr::fill(UID, ICAO24) |> 
+  #   trj_add_target_bearing(.my_thr)
 }

### 2023-03-25 paper.tex
clean up project folder; update gitignore

@@
 
+Operational efficiency is a key element of addressing aviation's
+contribution to climate change. Emissions may increase due to the
+expected growth in international air traffic until lower emitting
+technologies and fuels and other mitigating measures are developed and
+deployed .. ICAO adopted at its 41st Assembly a long-term aspirational
+goal for international aviation
+\protect\hyperlink{ref-icao2022}{{[}1{]}}. In support to the Paris
+Agreement, the goal is to achieve net-zero carbon emissions by 2050.
+
+Levers for fuel reduction:
+
+* operational efficiency
+
+* market-based measures
+
+* sustainable aviation fuel
+
+* new aircraft propulsion and airframes
+
+A wider use and pick-up of sustainable aviation fuel, and new aircraft
+propulsion technologies or aircraft design requires further research.
+
+Despite the introduction of an initial market-based mechanism, immediate
+action to curb fuel burn and CO2 emissions rests with improvements of
+operational efficiency.
+
 The contribution of this paper comprise:
@@
 \subsection{header 2}\label{header-2}}
-
-some cool text.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting
-industry. Lorem Ipsum has been the industry's standard dummy text ever
-since the 1500s, when an unknown printer took a galley of type and
-scrambled it to make a type specimen book. It has survived not only five
-centuries, but also the leap into electronic typesetting, remaining
-essentially unchanged. It was popularised in the 1960s with the release
-of Letraset sheets containing Lorem Ipsum passages, and more recently
-with desktop publishing software like Aldus PageMaker including versions
-of Lorem Ipsum.
 
@@
 phases, e.g.~recent machine learning algorithm
-\protect\hyperlink{ref-sun2017flightphase}{{[}1{]}}. This paper
+\protect\hyperlink{ref-sun2017flightphase}{{[}2{]}}. This paper
 implements a heuristic approach with a focus on the detection of arrival
-traffic at the study airports.
+traffic at the study airports. Figure Figure~\ref{fig-EGLL-arrivals}
+shows the detected arrival flights for a single day.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{./figures/EGLL-arrivals-single-day.png}
+
+}
+
+\caption{\label{fig-EGLL-arrivals}Arrivals at London Heathrow (EGLL) on
+sample day}
+
+\end{figure}
 
@@
 
-\hypertarget{open-data-for-operational-performance-monitoring}{%
-\subsection{Open Data for Operational Performance
-Monitoring}\label{open-data-for-operational-performance-monitoring}}
-
 \hypertarget{arrival-sequencing}{%
@@
 
-sequence separation
+Spacing deviation
+
+Let us consider a pair of consecutive landing aircraft denoted leader
+and trailer, with s their required time spacing1. Using the constant
+time delay principle, the spacing deviation (or spacing error) at time t
+considers the current position of trailer at time t, and the past
+position of leader at time t -- s. Precisely, it is defined as the
+difference between the respective minimum times from these two positions
+(see figure below): spacing deviation (t) = min time (trailer (t)) --
+min time (leader (t -- s))
 
@@
 System, airport operators report movement data on a monthly basis
-\protect\hyperlink{ref-apdf_v1_2019}{{[}2{]}}.
+\protect\hyperlink{ref-apdf_v1_2019}{{[}3{]}}.
 
@@
 \begin{CSLReferences}{0}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-icao2022}{}}%
+\CSLLeftMargin{{[}1{]} }%
+\CSLRightInline{ICAO, {``Resolution A41-21: Consolidated statement of
+continuing ICAO policies and practices related to environmental
+protection {\textemdash} climate change,''} 2022 {[}Online{]}.
+Available:
+\url{https://www.icao.int/environmental-protection/Documents/Assembly/Resolution_A41-21_Climate_change.pdf}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-sun2017flightphase}{}}%
-\CSLLeftMargin{{[}1{]} }%
+\CSLLeftMargin{{[}2{]} }%
 \CSLRightInline{J. Sun, J. Ellerbroek, and J. Hoekstra, {``Flight
@@
 \leavevmode\vadjust pre{\hypertarget{ref-apdf_v1_2019}{}}%
-\CSLLeftMargin{{[}2{]} }%
+\CSLLeftMargin{{[}3{]} }%
 \CSLRightInline{EUROCONTROL, {``Eurocontrol {Specification} for

### 2023-03-27 paper.qmd
initialise milestione table for arrivals
extracting KATL single day data

@@
 note: Extra footnote.
+
+execute: 
+  echo: false
+
 # filters:
@@
 Despite the introduction of an initial market-based mechanism, immediate action to curb fuel burn and CO2 emissions rests with improvements of operational efficiency.
+<!-- verbatim -->
+ICAO has introduced the trajectory-based operations concept, which promises operational and environmental benefits to aviation. 
+where the flight trajectory of an aircraft is flown as close as possible to the userpreferred route, with as little disruptions as possible through collaborative decision-making mechanism.
+This includes reducing potential conflicts and resolving demand/capacity imbalances earlier and more efficiently.
+TBO can therefore bring significant operational and environmental benefits to aviation.
+<!-- end verbatim copy from ICAO paper -->
 
@@
 This paper implements a heuristic approach with a focus on the detection of arrival traffic at the study airports.
+Figure @fig-EGLL-arrivals shows the detected arrival flights for a single day.
+
+```{r}
+#| label: fig-EGLL-arrivals
+#| fig-cap: Arrivals at London Heathrow (EGLL) on sample day
+#| out-width: 45%     # 2 column half-size
+knitr::include_graphics("./figures/EGLL-arrivals-single-day.png")
+```
 
@@
 An aircraft is assigned to a landing runway based on the closeness of its pre-landing positions to the extended runway centerline.
+
+```{r}
+#| label: fig-KATL-arrivals
+#| fig-cap: Arrivals at Atlanta Hartsfiled (KATL) on sample day
+#| out-width: 45%     # 2 column half-size
+knitr::include_graphics("./figures/KATL-traffic-single-day.png")
+```
 

### 2023-03-27 x_KATL.qmd
extracting KATL single day data

---
title: "Untitled"
format: html
---

working with ATLANTA

```{r}
rwys_katl <- read_csv("./data/rwys-KATL.csv", show_col_types = FALSE)
arp_katl  <- tibble(LAT = 33.6366996,LON = -84.4278640, REF = "ARP", "ELEV" = 1026)

katl_ctrl_ls <- cast_rwy_ctr_line_ls(rwys_katl)
katl_rwy_box <- airport_threshold_box(rwys_katl)
katl_apt_box <- airport_centerline_box(rwys_katl)

fn <- here::here("data", "usa-katl-20220509.parquet")
katl_trjs <- arrow::read_parquet(fn)

# |> 
#   coerce_unixepoch_to_datetime() |> 
#   identify_trajectory_legs() |> 
#   fix_invalid_fltid() |> 
#   make_uid("2022-05-09")
```

```{r}
ggplot() +
  geom_sf(data = katl_apt_box) +
  geom_sf(data = katl_rwy_box, fill = "lightblue") +
  geom_sf(data = katl_ctrl_ls) +
  geom_point(data = katl_trjs |> filter(ALT_B <= arp_katl$ELEV + 500), aes(x = LON, y = LAT), size = 0.2, alpha = 0.2) +
  coord_sf(xlim = c(-84.8, -84.1), ylim = c(33.5,33.7) ) +
  labs(x = NULL, y = NULL, subtitle = "KATL - Atlanta Hartsfield") +
  theme(plot.title.position = "plot")
#ggsave("./figures/KATL-traffic-single-day.png", dpi = 300, width = 8, height = 5)
```

identify traffic arriving at Hartsfiled

```{r}
icaos_seen <- katl_trjs |> filter(ALT_B <= arp_katl$ELEV + 1500) |> icao24_in_polygon(katl_apt_box)

study_trjs <- katl_trjs |> filter(ICAO24 %in% icaos_seen$ICAO24) |> 
  coerce_unixepoch_to_datetime() |> 
  identify_trajectory_legs() |> 
  mutate(TID = paste(ICAO24, LEG, sep = "-"))

icaos_seen2 <- study_trjs |> 
  filter(ALT_B <= arp_katl$ELEV + 1500) |> 
  icao24_in_polygon(katl_apt_box, .id = TID) |> 
  heuristic_DAO()

heuristic_arrs <- icaos_seen2 |> filter(DAO == "A")

study_trjs <- study_trjs |> filter(TID %in% heuristic_arrs$TID)  
```

### 2023-03-27 x_bra-airport-geoserver.qmd
extracting KATL single day data

---
title: "BRA airport data"
format: html
---

BRA airport geo data

based on coordination

* go to geo server https://geoaisweb.decea.mil.br/geoserver/web/wicket/bookmarkable/org.geoserver.web.demo.MapPreviewPage?0
* search for 
  + airport
  rwydirection
  + runway_v2

select file type to download files.

Stored in data/bra.

```{r}
library(tidyverse)

bra_apt_arp <- readr::read_csv(here::here("data","bra","airport.csv"), show_col_types = FALSE)
bra_rwys    <- readr::read_csv(here::here("data","bra","runway_v2.csv"), show_col_types = FALSE)
bra_thrs    <- readr::read_csv(here::here("data","bra","rwydirection.csv"), show_col_types = FALSE)

bra_study_apts <- bra_apt_arp |> 
  filter(localidade_id %in% c("SBGR", "SBSP","SBRJ")) |> 
  select(ICAO = localidade_id, APT_NAME = nome, LAT = latitude_dec, LON = longitude_dec, ELEV = elevacao, airport_pk) |> mutate(ELEV = coerce_meter_to_feet(ELEV))
```

```{r}
check_bra_rwys <- bra_rwys |> filter(airport_pk %in% bra_study_apts$airport_pk) |> glimpse()
```

```{r}
check_bra_thrs <- check_bra_rwys |> filter(runway_pk %in% check_bra_rwys$runway_pk) |> glimpse()
```

### 2023-03-27 x_bra-check.qmd
extracting KATL single day data

@@
   , "SBGR", "ARP", -23.435555, -46.473056, "ARP"    , 2461
-  , "SBGR", "O9L", -23.434281, -46.483361, "09L/27R", 2444
+  , "SBGR", "09L", -23.434281, -46.483361, "09L/27R", 2444
   , "SBGR", "27R", -23.424894, -46.448619, "09L/27R", 2440
@@
 
-#write_csv(rwys_sbgr, "./data/rwys-SBGR.csv")
+# write_csv(rwys_sbgr, "./data/rwys-SBGR.csv")
 ```
@@
 ```{r}
+rwys_sbgr  <- read_csv("./data/rwys-SBGR.csv", show_col_types = FALSE)
 sbgr_arp   <- rwys_sbgr |> filter(ICAO == "SBGR", REF == "ARP") |> select(LAT, LON)
@@
   mutate(NAME = "SBGR-200")
+
+# cast sf
+arp_pt    <- sbgr_arp  |> cast_latlon_to_pts(.drop_coord = FALSE) 
+rwys_pts  <- rwys_sbgr |> filter(REF != "ARP") |> cast_latlon_to_pts(.drop_coord = FALSE)
+rwys_ls   <- rwys_sbgr |> cast_rwy_ls()
+
+sbgr_centerline_ls <- cast_rwy_ctr_line_ls(rwys_sbgr |> filter(REF != "ARP"))
+sbgr_rwy_box       <- airport_threshold_box(rwys_sbgr |> filter(REF != "ARP"))
+sbgr_box           <- airport_centerline_box(rwys_sbgr |> filter(REF != "ARP"), .ctrl_length = 15000)
 ```
@@
   geom_sf(data = sbgr_200NM) +
+  geom_sf(data = sbgr_box) +
   geom_point(data = sbgr_arp, aes(x = LON, y = LAT))
@@
 
-```{r}
+```{}
 # load helper scripts to download and store OSN data
@@
 ```
-```{r}
+```{}
 #| message: false
@@
 
-```{r}
+```{}
 spatial_subset <- function(.payload, .outer, ...){
   # check if payload is already a sf object
+  # if not, coerce to sf
   if(!is(.payload,"sf")){
@@
   }
+  # spatial subsetting on global scale is calc intensive for sf
+  # check for bounding box of .outer
+  bbox <- sf::st_bbox(.outer)
+
+  # simple pre-filter based on LAT/LON ~ bounding box
+  trimmed_data <- .payload |> 
+    dplyr::filter(LAT >= bbox$ymin, LAT <= bbox$ymax) |> 
+    dplyr::filter(LON >= bbox$xmin, LON <= bbox$xmax)
+
   # spatial subsetting
-  trimmed_data <- .payload[.outer,]
+  trimmed_data <- trimmed_data[.outer,]
 }
@@
 ```{r}
+trjs_pts <- arrow::read_parquet("./data/bra-sbgr-20220509.parquet") |> 
+  coerce_unixepoch_to_datetime() |> 
+  cast_latlon_to_pts(.drop_coord = FALSE)
+
 sbgr_base_plot + geom_sf(data = trjs_pts)
+```
+
+Check for coverage close to SBGR
+
+```{r}
+arp_sbgr <- rwys_sbgr |> filter(REF == "ARP") |> select(LAT, LON)
+apt_tfc  <- trjs_pts |> 
+  trrrj::filter_positions_at_range(
+    c(arp_sbgr$LON, arp_sbgr$LAT), 0, 15, lon = LON, lat = LAT
+    ) |> 
+  filter(ALT_B < 4000)
+
+ggplot() + 
+  geom_sf(   data = sbgr_rwys_ls, col = "blue" ) +
+  geom_sf(   data = sbgr_box, fill = NA) +
+  geom_point(data = apt_tfc, aes(x = LON, y = LAT), size = 0.5)
+```
+
+Which ids do we see at SBGR?
+
+```{r}
+sbgr_uids <- apt_tfc[sbgr_box,] |> distinct(ICAO24, FLTID)
+viz_check_vertical(apt_tfc |> filter(ICAO24 %in% sbgr_uids$ICAO24))
+```
+
+```{r}
+sbgr_trjs <- trjs_pts |> 
+  filter(ICAO24 %in% sbgr_uids$ICAO24) |> 
+  identify_trajectory_legs() |> 
+  fix_invalid_fltid() |> 
+  make_uid("2022-05-09")
+
+ggplot() + geom_sf(data = sbgr_trjs)
+```
+
+```{r}
+picks <- trjs_pts |> filter(FLTID %in% c("TAM3201", "GLO1612","TAM3172", "DLH506"))
+viz_check_DAO(picks |> filter(ALT_B < 8000), sbgr_rwys_ls |> sf::st_buffer(dist = 10000)) + geom_sf(data = sbgr_rwys_ls)
+
+viz_check_vertical(picks)
 ```

### 2023-03-27 R/cast_sf_rwy.R

@@ -60,7 +60,7 @@ rwy_ctr_line_pts <- function(.rwy_ctr_line, .debug = FALSE){
   return(df)  
 }
 
-cast_rwy_ctr_line_ls <- function(.rwys_df){
+cast_rwy_ctr_line_ls <- function(.rwys_df, ...){
   .rwys_df |> 
     rwy_ctr_line() |> 
     rwy_ctr_line_pts() |> 
@@ -80,9 +80,10 @@ airport_threshold_box <- function(.rwys_df, .thr_buffer = 500, ...){
   return(rwys_box)
 }
 
-airport_centerline_box <- function(.rwys_df, .thr_buffer = 500, .ctrline_buffer = 500){
+airport_centerline_box <- function(.rwys_df, .thr_buffer = 500, .ctrline_buffer = 500, ...){
   airport_box  <- airport_threshold_box(.rwys_df, .thr_buffer)
-  ctr_line_buf <- cast_rwy_ctr_line_ls(.rwys_df) |> sf::st_buffer(dist = .ctrline_buffer)
+  ctr_line_buf <- cast_rwy_ctr_line_ls(.rwys_df) |> 
+    sf::st_buffer(dist = .ctrline_buffer)
   
   combo_box    <- sf::st_union(airport_box, ctr_line_buf) |> 
     sf::st_convex_hull() |> 

### 2023-03-27 R/osn_clean_statevector.R

@@ -1,0 +1,13 @@
+#' Title
+#'
+#' @param .df 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+clean_osn_statevectors <- function(.df){
+  clean_states <- .df |> 
+    make_nice_names_osn() |> 
+    mutate(across(.cols = c("ALT_B","ALT_G"), .fns = coerce_meter_to_feet))
+}

### 2023-03-27 R/trj_make_uid.R

@@ -1,0 +1,17 @@
+#' assign an UID for the study by attributing ICAO24, FLTID, DOF, and LEG
+#'
+#' @param .trj_with_legs 
+#' @param .dof 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+make_uid <- function(.trj_with_legs, .dof){
+  # strip off hyphons from date
+  dof <- .dof |> gsub(pattern = "-", replacement = "")
+  my_trjs <- .trj_with_legs |> 
+    dplyr::mutate(UID = paste(ICAO24, dof, LEG, FLTID, sep = "-")) |> 
+    dplyr::select(UID, dplyr::everything())
+  return(my_trjs)
+}

### 2023-03-27 xhelper_quick_viz.R

@@ -1,0 +1,18 @@
+viz_check_DAO <- function(.trj, .apt_box = airport_box){
+  # check dim of box
+  bbox = sf::st_bbox(.apt_box)
+  
+  plt <- .trj |> 
+    ggplot() + 
+    geom_sf(data = .apt_box, fill = NA, color = "blue") + 
+    geom_point(data = .trj, aes(x = LON, y = LAT, color = TIME)) + 
+    coord_sf(xlim = c(bbox$xmin, bbox$xmax), ylim = c(bbox$ymin, bbox$ymax), expand = FALSE)
+  return(plt)
+}
+
+viz_check_vertical <- function(.trj){
+  plt <- .trj |> 
+    ggplot() +
+    geom_point(aes(x = TIME, y = ALT_B))
+  return(plt)
+}

### 2023-03-30 x_KATL.qmd
building study data loop

@@
 ---
-title: "Untitled"
+title: "Spacing Deviation at US airports"
 format: html
 ---
+
+```{r}
+library(tidyverse)
+
+files_to_sources = list.files(path = "./R", full.names = TRUE)
+purrr::walk(files_to_sources, .f = source)
+```
 
@@
 katl_apt_box <- airport_centerline_box(rwys_katl)
+
+katl_200NM <- trrrj::polygon_at_distance(c(arp_katl$LON, arp_katl$LAT), d = 200) |> 
+  sf::st_as_sf() |> mutate(REF = "KATL-200NM")
 
@@
   identify_trajectory_legs() |> 
+ # fix_invalid_fltid() |> 
   mutate(TID = paste(ICAO24, LEG, sep = "-"))
@@
 study_trjs <- study_trjs |> filter(TID %in% heuristic_arrs$TID)
+
+arrivals <- study_trjs |> 
+  assign_landing_runway(
+      .arp = arp_katl
+    , .rwy_centerline_ls = katl_ctrl_ls 
+    , .upper_bound = 10
+    , .group_var = TID
+    ) |> 
+  filter(N > 5)    # for 10 sec update crossers have typically low number of hits
 ```
+```{r}
+itr <- arrivals[1,] |> select(UID = TID, UID_REF = REF)
+
+make_study_arrival_trajectories <- function(.itr, .trjs, .rwys){
+  message("\n", itr$UID)
+  app_trj <- .trjs |> dplyr::filter(UID == itr$UID)
+  app_thr <- .rwys |> dplyr::filter(REF == itr$UID_REF)
+
+  # estimate threshold (aka landing time)
+  thr_row <- estimate_threshold_time(app_trj, app_thr) |> 
+    select(UID, ICAO24, FLTID, TIME, LAT, LON ) |> 
+    mutate(MST = "THR", MST_LABEL = itr$UID_REF)
+
+  # append threshold row and trim trajectory
+  app_trj <- app_trj |> bind_rows(thr_row) |> 
+    select(UID, ICAO24, FLTID, TIME, LAT, LON, ALT_B, MST, MST_LABEL) |> 
+    arrange(TIME) |> 
+    rowid_to_column("SEQ_ID")
+
+  # add "distance-to-go" and "time-to-fly" 
+  app_trj <- app_trj |> add_flight_distance()
+  return(app_trj)
+}
+
+bingo <- itr |> make_study_arrival_trajectories(study_trjs |> rename(UID = TID), rwys_katl)
+```
+
+run for many, function re-written to add RWY to trjs
+
+```{r}
+study_payload <- study_trjs |> 
+  select(TID, ICAO24, FLTID, TIME, LAT, LON, ALT_B) |> 
+  left_join(arrivals |> select(TID, RWY = REF), by = join_by(TID)) |> 
+  rename(UID = TID)
+
+rqq <- study_payload |> 
+  mutate(LUID = UID) |> 
+  group_by(LUID) |> 
+  group_map(.f = ~ safer_make_study_arrival_trajectories2( ., .rwys = rwys_katl))
+
+# check for errors
+names(rqq) <- seq_along(rqq)
+keep(rqq, ~ ! is.data.frame(.x))
+```
+
+```{r}
+next_payload <- keep(rqq, ~is.data.frame(.x)) |> bind_rows()
+```

### 2023-03-30 x_bra-check.qmd
building study data loop

@@
 ```
-```{}
+```{r}
 #| message: false
@@
 katl_arp   <- c(-84.4278640,33.6366996)
-katl_200NM <- trrrj::polygon_at_distance(katl_arp, d = 205) |> sf::st_as_sf() |> 
-  rename(geometry = x) |> 
-  mutate(NAME = "KATL-205")
+arp_katl   <- tribble(
+  ~LAT,   ~LON,~REF,~ELEV
+  ,33.6, -84.4,"ARP", 1026
+)
+
+# SPOILER =========
+# trrrj polygon at uses "ireland crs"! 
+# katl_200NM <- trrrj::polygon_at_distance(katl_arp, d = 205) |> sf::st_as_sf() |> 
+#   rename(geometry = x) |> 
+#   mutate(NAME = "KATL-205")
+katl_200NM <- arp_katl |> select(LAT, LON) |> 
+  cast_latlon_to_pts() |> sf::st_buffer(205 * 1852) |> 
+  mutate(REF = "KATL-205NM")
 

### 2023-03-30 x_data-crunching.qmd
building study data loop

@@
 ```{r}
-icao24_in_polygon <- function(.ds, .poly, .id = ICAO24){
-  # determine bounding box for polygon
-  this_box <- sf::st_bbox(.poly)
-  # pre-filter with dplyr - cut bounding box
-  icao24s <- .ds |> 
-    dplyr::filter(
-       LAT > this_box$ymin, LAT < this_box$ymax
-      ,LON > this_box$xmin, LON < this_box$xmax, 
-    )
-  # get ids seen
-  icao24s <- icao24s |> 
-    dplyr::summarise(
-      N = n(), N_ID = length(unique(FLTID))
-      , MIN_TIME = min(TIME,na.rm = TRUE)
-      , MIN_TIME_ALT = ALT_B[which(TIME == MIN_TIME)]
-      , MAX_TIME = max(TIME, na.rm = TRUE)
-      , MAX_TIME_ALT = ALT_B[which(TIME == MAX_TIME)]
-      , MIN_ALT = min(ALT_B, na.rm = TRUE)
-      , MED_ALT = median(ALT_B, na.rm = TRUE)
-      , MAX_ALT = max(ALT_B, na.rm = TRUE)
-      , .by = {{.id}})
-}
+source("~/RProjects/paper-2023-ICNS/R/trj_icao24_in_polygon.R")
+# MOVED TO R
+# ==================
+# icao24_in_polygon <- function(.ds, .poly, .id = ICAO24){
+#   # determine bounding box for polygon
+#   this_box <- sf::st_bbox(.poly)
+#   # pre-filter with dplyr - cut bounding box
+#   icao24s <- .ds |> 
+#     dplyr::filter(
+#        LAT > this_box$ymin, LAT < this_box$ymax
+#       ,LON > this_box$xmin, LON < this_box$xmax, 
+#     )
+#   # get ids seen
+#   icao24s <- icao24s |> 
+#     dplyr::summarise(
+#       N = n(), N_ID = length(unique(FLTID))
+#       , MIN_TIME = min(TIME,na.rm = TRUE)
+#       , MIN_TIME_ALT = ALT_B[which(TIME == MIN_TIME)]
+#       , MAX_TIME = max(TIME, na.rm = TRUE)
+#       , MAX_TIME_ALT = ALT_B[which(TIME == MAX_TIME)]
+#       , MIN_ALT = min(ALT_B, na.rm = TRUE)
+#       , MED_ALT = median(ALT_B, na.rm = TRUE)
+#       , MAX_ALT = max(ALT_B, na.rm = TRUE)
+#       , .by = {{.id}})
+# }
 ```

### 2023-03-30 x_replicate.qmd
building study data loop

---
title: "Replicate earlier study"
format: html
---

check what we have done in the past

```{r}
library(tidyverse)

# read and load study data
source("~/RProjects/paper-2023-ICNS/R/osn_read_tardump.R")
source("~/RProjects/paper-2023-ICNS/R/osn_make_nice_names.R")
source("~/RProjects/paper-2023-ICNS/R/coerce_meter_to_feet.R")
source("~/RProjects/paper-2023-ICNS/R/coerce_unixepoch_to_datetime.R")
# subset study data ~ jormungand
source("~/RProjects/paper-2023-ICNS/R/spatial_subset_jormungand.R")
source("~/RProjects/paper-2023-ICNS/R/cast_sf_utils.R")

# loop and trajectory processing
source("~/RProjects/paper-2023-ICNS/R/trj_extract_4d_nest.R")
source("~/RProjects/paper-2023-ICNS/R/trj_icao24_in_polygon.R")
source("~/RProjects/paper-2023-ICNS/R/trj_confirm_arrival_at.R")

# runway 
source("~/RProjects/paper-2023-ICNS/R/cast_sf_rwy.R")
```

Load one day of OSN dump

```{}
#| message: false
pth_to_dumps <- "../__DATA/OSN-dumps"
ptn <- "2019-08-01"

# list downloaded dumps
fns <- list.files(path = pth_to_dumps, pattern = ptn, full.names = TRUE)

# load list of chosen dumps
ds  <- fns |> 
  purrr::map_dfr(.f = read_osn_tar_file) |> 
  make_nice_names_osn()
ds <- ds |> 
  mutate(across(.cols = c("ALT_B","ALT_G"), .fns = coerce_meter_to_feet)) 
ds <- ds |> coerce_unixepoch_to_datetime()
```

Subset for study airports
EGLL, EHAM, LSZH

```{}
aip <- read_csv("./data/aip-arp-rwys.csv", show_col_types = FALSE)

apt <- "LSZH"
arp <- filter(aip, ICAO == apt, REF == "ARP") |> rename(ELEV = GEOM_ALT)
jormungand <- trrrj::polygon_at_distance(c(arp$LON, arp$LAT), d = 205)

svs <- spatial_subset_jormungand(ds, jormungand)

arrow::write_parquet(svs |> sf::st_drop_geometry(), paste0("./data/", apt, "-200NM-20190801.feather"))
```

## Load Heathrow

```{r}
apt <- "EGLL"
fn  <- list.files(path = "./data", pattern = paste0(apt,".*\\.feather"), full.names = TRUE)
```

```{r}
study_loop <- function(.apt, .fn, .debug, ...){

  #----------- read airport data -------------------------
  aip <- readr::read_csv("./data/aip-arp-rwys.csv", show_col_types = FALSE) |> 
    dplyr::filter(ICAO == .apt)
  if("GEOM_ALT" %in% names(aip)){
    aip <- aip |> dplyr::rename(ELEV = GEOM_ALT)
    }

  arp     <- aip |> dplyr::filter(ICAO == .apt,   REF == "ARP")
  rwys    <- aip |> dplyr::filter(ICAO == .apt, ! REF == "ARP")
  apt_box <- airport_centerline_box(rwys, ...)

  # ---------------------- load trajectory data --------------------------------
  warning(paste(.apt, " - reading - ", .fn))
  ds <- arrow::read_feather(.fn) %>% 
    tibble::as_tibble() |> 
    #---- enforce above threshold ~ ELEV arp
    dplyr::filter(ALT_B >= arp$ELEV + 300)

  #---------------------- check for local traffic ----------------------------
  # subset icao24s inside airport box
  seen_at_airport <- ds |> 
    dplyr::filter(ALT_B <= arp$ELEV + 4000) |> 
    spatial_subset_jormungand(apt_box, ...) |> 
    dplyr::distinct(ICAO24)

  ds <- ds |> 
    dplyr::filter(ICAO24 %in% seen_at_airport$ICAO24) |> 
    identify_trajectory_legs(.max_gap = 5) |> 
    mutate(UID = paste(ICAO24, LEG, sep = "-"))

  phase <- ds |> extract_4d_nest() |> confirm_arrival_at(apt_box)
}
```

```{r}
gotcha <- study_loop(apt, fn)
```

rq |> 
  # subset for ICAO24s seen at the study airport
  filter(ICAO24 %in% seen_at_airport$ICAO24) |> 
  spatial_subset_jormungand(jormungand) |> 
  # human readable TIME
  coerce_unixepoch_to_datetime() |>
  # clean ALT and trim for airborne part with heuristic fix
  Alt_Outlier_filter(.alt_var = ALT_B) |> 
  fix_altitude_heuristic() |> 
  # assign leg for each airborne segment
  identify_trajectory_legs() |> 
  # force single FLTID here to construct niche UIDs
  group_by(ICAO24, LEG) |> fix_invalid_fltid() |> ungroup() |> 
  # construct UID
  mutate(UID = paste(ICAO24, LEG, sep = "-"))

### 2023-03-30 R/add_distance_rowwise.R

@@ -18,15 +18,15 @@ add_flight_distance <- function(.trj, unit = "NM"){
   
   df <- df %>%
     dplyr::mutate(
-      LAT_PREV = lag(LAT, default = first(LAT))
-      ,LON_PREV = lag(LON, default = first(LON))
-      ,DIST_FLOWN = geosphere::distHaversine( cbind(LON,LAT)
+      LAT_PREV =  dplyr::lag(LAT, default = dplyr::first(LAT))
+      ,LON_PREV = dplyr::lag(LON, default = dplyr::first(LON))
+      ,DIST_NEXT = geosphere::distHaversine(  cbind(LON,LAT)
                                               ,cbind(LON_PREV, LAT_PREV)
                                               , r = unit_radius)
-      ,DIST_FLOWN = cumsum(DIST_FLOWN)
+      ,DIST_FLOWN = cumsum(DIST_NEXT)
     ) %>%
     # remove additional columns
-    select(-LAT_PREV, - LON_PREV)
+    dplyr::select(-LAT_PREV, - LON_PREV)
   return(df)
 }
 

### 2023-03-30 R/cast_sf_rwy.R

@@ -62,7 +62,7 @@ rwy_ctr_line_pts <- function(.rwy_ctr_line, .debug = FALSE){
 
 cast_rwy_ctr_line_ls <- function(.rwys_df, ...){
   .rwys_df |> 
-    rwy_ctr_line() |> 
+    rwy_ctr_line(...) |> 
     rwy_ctr_line_pts() |> 
     cast_latlon_to_pts() |> 
     cast_pts_to_ls(.group_var = REF)

### 2023-03-30 R/estimate_threshold_time.R

@@ -8,6 +8,7 @@
 #'
 #' @examples
 estimate_threshold_time <- function(.trjs_latlon_sf, .my_thr){
+  this_trj <- .trjs_latlon_sf
   
   # decode runway identifier
   RWY <- .my_thr$REF
@@ -15,12 +16,25 @@ estimate_threshold_time <- function(.trjs_latlon_sf, .my_thr){
     rwy_dir <- gsub("\\D","", .rwy) # strip non-digets and replace with empty
     rwy_dir <- ((as.numeric(rwy_dir) * 10) + 360) %% 360
   }
+  # check if BEARING and DISTANCE to threshold are already calculated
+  if(!"DIST_THR" %in% colnames(this_trj)){
+    this_trj <- this_trj |> 
+      dplyr::mutate(
+        DIST_THR = geosphere::distHaversine(
+                cbind(LON, LAT)
+              , cbind(.my_thr$LON,.my_thr$LAT)
+              )
+        )
+  }
+  if(!"BRG_THR" %in% colnames(this_trj)){
+    this_trj <- this_trj |> 
+      dplyr::mutate(
+        BRG_THR = geosphere::bearingRhumb(cbind(LON, LAT), cbind(.my_thr$LON,.my_thr$LAT)))
+  }
   
-  last_pos <- .trjs_latlon_sf %>%
-    mutate(#DIST_THR = distHaversine(cbind(LON, LAT), cbind(LON_THR,LAT_THR))
-           #, BRG_THR = bearingRhumb( cbind(LON, LAT), cbind(LON_THR, LAT_THR))
-           #,
-             BEFORE  = abs(BRG - rwy_to_number(RWY)) < 30
+  last_pos <- this_trj %>%
+    mutate(
+             BEFORE  = abs(BRG_THR - rwy_to_number(RWY)) < 30
            , COASTG  = DIST_THR == lag(DIST_THR)
            , V_GND   = (lag(DIST_THR) - DIST_THR) / as.numeric(TIME - lag(TIME))
            , V_GND_SM= zoo::rollmean(V_GND, k = 9, fill = NA, align = "right")

### 2023-03-30 R/make_study_arrival_trajectories.R

@@ -1,0 +1,34 @@
+make_study_arrival_trajectories2 <- function(.trjs, .rwys){
+  this_itr <- .trjs |> dplyr::distinct(UID, RWY)
+  message("\n", this_itr$UID)
+  app_trj <- .trjs |> dplyr::filter(UID == this_itr$UID)
+  app_thr <- .rwys |> dplyr::filter(REF == this_itr$RWY)
+  
+  # estimate threshold (aka landing time)
+  thr_row <- estimate_threshold_time(app_trj, app_thr) |> 
+    dplyr::select(UID, ICAO24, FLTID, TIME, LAT, LON ) |> 
+    dplyr::mutate(MST = "THR", MST_LABEL = this_itr$RWY)
+  
+  # append threshold row and trim trajectory
+  app_trj <- app_trj |> 
+    dplyr::bind_rows(thr_row) |> 
+    dplyr::select(UID, ICAO24, FLTID, TIME, LAT, LON, ALT_B, MST, MST_LABEL) |> 
+    dplyr::arrange(TIME) |> 
+    dplyr::rowid_to_column("SEQ_ID") |> 
+    # trim to threshold
+    dplyr::filter(TIME <= thr_row$TIME)
+  
+  # add "distance-to-go" and "time-to-fly" 
+  app_trj <- app_trj |> 
+    add_flight_distance() 
+  # |> 
+  #   dplyr::mutate(
+  #      DIST_TO_THR = thr_row$DIST_FLOWN - DIST_FLOWN
+  #     ,TIME_TO_THR = difftime(thr_row$TIME, TIME, unit = "sec")
+  #   )
+  return(app_trj)
+}
+
+safe_make_study_arrival_trajectories2 <- purrr::safely(.f = make_study_arrival_trajectories2, otherwise = "ERROR" )
+
+safer_make_study_arrival_trajectories2 <- purrr::possibly(.f = make_study_arrival_trajectories2, otherwise = "ERROR")

### 2023-03-30 R/trj_confirm_arrival_at.R

@@ -1,0 +1,30 @@
+#' Utility function to confirm arrival phase at an airport.
+#'
+#' The time-ordered set of trajectories is spatially intersected with a polygon
+#' around the airport (aka airport box).
+#'
+#' @param .df
+#' @param .apt_box
+#'
+#' @return
+#' @export
+#'
+#' @examples
+confirm_arrival_at <- function(.df, .apt_box, .time_window = 500){
+  df <- .df %>%
+    # extract last (aka oldest) pos reports
+    dplyr::mutate(last_4d = map(
+      .x = data
+      ,.f = ~ filter(.x, TIME >= max(TIME) - .time_window)
+    ) # pam
+    ) %>%
+    
+    dplyr::mutate(ARR = map(
+      .x = last_4d
+      ,.f = ~ cast_latlon_to_ls(.x) %>%
+        st_intersects(.apt_box, sparse = FALSE))
+    ) %>%
+    tidyr::unnest(ARR)
+  df <- df %>% dplyr::mutate(ARR = as.logical(ARR))
+  return(df)
+}

### 2023-03-30 R/trj_extract_4d_nest.R

@@ -1,0 +1,14 @@
+#' Extract and nest 4D positions
+#'
+#' @param .trjs_df
+#'
+#' @return nested dataframe/tibble
+#' @export
+#'
+#' @examples
+extract_4d_nest <- function(.trjs_df){
+  df <- .trjs_df %>%
+    dplyr::select(UID, TIME, LAT, LON, ALT_B) %>%
+    dplyr::group_by(UID) %>%
+    tidyr::nest()
+}

### 2023-03-30 R/trj_icao24_in_polygon.R

@@ -1,0 +1,22 @@
+icao24_in_polygon <- function(.ds, .poly, .id = ICAO24){
+  # determine bounding box for polygon
+  this_box <- sf::st_bbox(.poly)
+  # pre-filter with dplyr - cut bounding box
+  icao24s <- .ds |> 
+    dplyr::filter(
+      LAT > this_box$ymin, LAT < this_box$ymax
+      ,LON > this_box$xmin, LON < this_box$xmax, 
+    )
+  # get ids seen
+  icao24s <- icao24s |> 
+    dplyr::summarise(
+      N = n(), N_ID = length(unique(FLTID))
+      , MIN_TIME = min(TIME,na.rm = TRUE)
+      , MIN_TIME_ALT = ALT_B[which(TIME == MIN_TIME)]
+      , MAX_TIME = max(TIME, na.rm = TRUE)
+      , MAX_TIME_ALT = ALT_B[which(TIME == MAX_TIME)]
+      , MIN_ALT = min(ALT_B, na.rm = TRUE)
+      , MED_ALT = median(ALT_B, na.rm = TRUE)
+      , MAX_ALT = max(ALT_B, na.rm = TRUE)
+      , .by = {{.id}})
+}

### 2023-03-30 R/trj_landing_runway.R

@@ -13,13 +13,13 @@
 #' @examples
 estimate_rwy <- function(.trj_pts, .rwy_ctrl_ls, .ctrl_offset = 1000){
   # stop if not
-  tmp <- st_join( .rwy_ctrl_ls, .trj_pts
-                  , join = st_is_within_distance, dist = .ctrl_offset) %>%
-    st_drop_geometry() %>%        # keep just the "spatially joined" dataframe
+  tmp <- sf::st_join( .rwy_ctrl_ls, .trj_pts
+                  , join = sf::st_is_within_distance, dist = .ctrl_offset) %>%
+    sf::st_drop_geometry() %>%        # keep just the "spatially joined" dataframe
     # na.omit() %>%                 # eliminate no hits
-    group_by(REF) %>%
-    summarise(N = n(), .groups = "drop") %>%
-    mutate( TOT   = sum(N, na.rm = TRUE)       # na.rm should not be necessary
+    dplyr::group_by(REF) %>%
+    dplyr::summarise(N = dplyr::n(), .groups = "drop") %>%
+    dplyr::mutate( TOT   = sum(N, na.rm = TRUE)       # na.rm should not be necessary
             ,TRUST = N / TOT)                   # share in case several runways
   
   return(tmp)
@@ -56,17 +56,17 @@ arrival_runway_assignment <- function(.rwy_hit_score){
 #' @export
 #'
 #' @examples
-assign_landing_runway <- function(.arr_trjs, .arp = arp_egll , .group_var = UID ,...){
+assign_landing_runway <- function(.arr_trjs, .arp = arp_egll, .rwy_centerline_ls, .upper_bound = 12 , .group_var = UID ,...){
   final_trjs <- .arr_trjs |>
     # omit flights close to ARP ~ remove potential surface hits -> min = 2!
     # find a better way to filter final 
-    trrrj::filter_positions_at_range(c(.arp$LON, .arp$LAT), 2, 10, lon = LON, lat = LAT)
+    trrrj::filter_positions_at_range(c(.arp$LON, .arp$LAT), 2, .upper_bound, lon = LON, lat = LAT)
   
   final_trjs_pts <- final_trjs |> cast_latlon_to_pts()
   
   ldg_rwy <- final_trjs_pts |> 
     group_by({{.group_var}}) |> 
-    group_modify(.f = ~ estimate_rwy(.x, rwy_centerline_ls) |> arrival_runway_assignment())
[…]

### 2023-03-31 x_replicate.qmd
nest stab at new study loop

@@
   rwys    <- aip |> dplyr::filter(ICAO == .apt, ! REF == "ARP")
+  rwy_centerline_ls <- cast_rwy_ctr_line_ls(rwys)
   apt_box <- airport_centerline_box(rwys, ...)
@@
 
-  phase <- ds |> extract_4d_nest() |> confirm_arrival_at(apt_box)
+  phase <- ds |> 
+    extract_4d_nest() |> 
+    confirm_arrival_at(apt_box) 
 }
@@
 ```
+
+part of data cleaning 
+- remove stray orphans inside aerodrome
+
+gotcha |> ungroup() |> filter(ARR) |> mutate(GOTCHA = RWY_TRUST >= quantile(RWY_TRUST, probs = 0.1)) |> filter(GOTCHA, N_LAST < 15) |> filter(UID == "3c66b1-1") |> unnest(last_4d) |> ggplot() + geom_point(aes(x = LON, y = LAT)) + geom_point(data = arp, aes(x = LON, y = LAT), color = "red", size = 3) + geom_sf(data = my_apt_box, fill = NA)
 

### 2023-03-31 R/trj_confirm_arrival_at.R

@@ -10,21 +10,42 @@
 #' @export
 #'
 #' @examples
-confirm_arrival_at <- function(.df, .apt_box, .time_window = 500){
+confirm_arrival_at <- function(.df, .apt_box, .time_window = 500, ...){
   df <- .df %>%
     # extract last (aka oldest) pos reports
     dplyr::mutate(last_4d = map(
       .x = data
       ,.f = ~ filter(.x, TIME >= max(TIME) - .time_window)
-    ) # pam
-    ) %>%
-    
+        ) # pam
+    )  |> 
     dplyr::mutate(ARR = map(
       .x = last_4d
       ,.f = ~ cast_latlon_to_ls(.x) %>%
         st_intersects(.apt_box, sparse = FALSE))
     ) %>%
-    tidyr::unnest(ARR)
-  df <- df %>% dplyr::mutate(ARR = as.logical(ARR))
+    tidyr::unnest(ARR) |> 
+    dplyr::mutate(N_LAST = map(last_4d, nrow)) |> 
+    tidyr::unnest(N_LAST) |> 
+    ##------------ add runway estimation
+    dplyr::mutate(
+      RWY_SCORE = map(
+            .x = last_4d
+          , .f = ~ estimate_rwy(
+                    .x |> cast_latlon_to_pts()
+                  , rwy_centerline_ls
+                  ) |> 
+            arrival_runway_assignment()
+          )
+      ) |> mutate(RWY = map(.x = RWY_SCORE, .f = ~ .x$REF)) |> tidyr::unnest(RWY) |> 
+    mutate(
+      RWY_TRUST = map(
+                .x = RWY_SCORE
+              , .f = ~ .x$TRUST)
+      ) |> tidyr::unnest(RWY_TRUST)
+  
+  df <- df %>% 
+    dplyr::mutate(
+      ARR = as.logical(ARR)
+      )
   return(df)
 }

### 2023-03-31 R/trj_landing_runway.R

@@ -66,7 +66,10 @@ assign_landing_runway <- function(.arr_trjs, .arp = arp_egll, .rwy_centerline_ls
   
   ldg_rwy <- final_trjs_pts |> 
     group_by({{.group_var}}) |> 
-    group_modify(.f = ~ estimate_rwy(.x, .rwy_centerline_ls) |> arrival_runway_assignment())
+    group_modify(
+      .f = ~ estimate_rwy(.x, .rwy_centerline_ls) |> 
+             arrival_runway_assignment()
+      ) |> ungroup()
   
   return(ldg_rwy)
 }

### 2023-04-01 x_replicate.qmd
added orphan filter for cleaning arrivals

@@
 source("~/RProjects/paper-2023-ICNS/R/trj_confirm_arrival_at.R")
+source("~/RProjects/paper-2023-ICNS/R/trj_check_orphans.R")
+
+# trajectory cleaning
+source("~/RProjects/paper-2023-ICNS/R/trj_fix_altitude.R")
+source("~/RProjects/paper-2023-ICNS/R/altitude_outlier_filter.R")
 
@@
     #---- enforce above threshold ~ ELEV arp
-    dplyr::filter(ALT_B >= arp$ELEV + 300)
+    dplyr::filter(ALT_B >= arp$ELEV + 300) |>
+     fix_altitude_heuristic() |>
+     Alt_Outlier_filter(.alt_var = ALT_B)
 
@@
     extract_4d_nest() |> 
-    confirm_arrival_at(apt_box) 
+    confirm_arrival_at(apt_box) |> 
+    check_arrival_orphan()
 }

### 2023-04-01 R/trj_check_orphans.R

@@ -1,0 +1,22 @@
+#' Check for orphans not captured during data cleaning
+#' 
+#' Orphans are characterised by own UID and thus sample size equalling 
+#' "last points" size
+#'
+#' @param .df_nested 
+#'
+#' @return
+#' @export
+#'
+#' @examples \dontrun{}
+check_arrival_orphan <-  function(.df_nested){
+  .df_nested |> 
+    dplyr::mutate(
+      N_DATA = map(
+              .x = data
+            , .f = ~ nrow(.x))
+      ) |> 
+    tidyr::unnest(N_DATA) |>
+    # check whether "last points" and data are of same size
+    dplyr::mutate(ORPHAN = N_DATA == N_LAST)
+}

### 2023-04-01 x_replicate.qmd
getting data prep in place

@@
 source("~/RProjects/paper-2023-ICNS/R/cast_sf_rwy.R")
+
+source("~/RProjects/paper-2023-ICNS/R/assign_threshold_time2.R")
+source("~/RProjects/paper-2023-ICNS/R/identify_threshold_overhead.R")
+source("~/RProjects/paper-2023-ICNS/R/add_bearing_to_target.R")
+source("~/RProjects/paper-2023-ICNS/R/estimate_threshold_time.R")
+source("~/RProjects/paper-2023-ICNS/R/identify_threshold_overhead.R")
+source("~/RProjects/paper-2023-ICNS/R/add_landing_time_to_nested_data.R")
 ```
@@
     check_arrival_orphan()
+
+  # # add landing time
+  # with_ldg_time <-  phase |> 
+  #   dplyr::mutate(
+  #     study_trj = 
+  #     case_when(
+  #       ARR & ! ORPHAN ~ pmap(
+  #             .l = list(data, UID, RWY)
+  #           , .f = ~ add_landing_time_to_nested_data(..1, ..2, ..3, rwys))) 
+  #     , TRUE ~ "Departure"
+  #   )
 }
@@
 
-part of data cleaning 
-- remove stray orphans inside aerodrome
+```{r}
+test_landing_time <- function(phase, rwys = rwys_egll){
 
-gotcha |> ungroup() |> filter(ARR) |> mutate(GOTCHA = RWY_TRUST >= quantile(RWY_TRUST, probs = 0.1)) |> filter(GOTCHA, N_LAST < 15) |> filter(UID == "3c66b1-1") |> unnest(last_4d) |> ggplot() + geom_point(aes(x = LON, y = LAT)) + geom_point(data = arp, aes(x = LON, y = LAT), color = "red", size = 3) + geom_sf(data = my_apt_box, fill = NA)
+  arrs <- phase |> filter(ARR & !ORPHAN)
+  rest <- phase |> filter(! UID %in% arrs$UID )
 
-rq |> 
-  # subset for ICAO24s seen at the study airport
-  filter(ICAO24 %in% seen_at_airport$ICAO24) |> 
-  spatial_subset_jormungand(jormungand) |> 
-  # human readable TIME
-  coerce_unixepoch_to_datetime() |>
-  # clean ALT and trim for airborne part with heuristic fix
-  Alt_Outlier_filter(.alt_var = ALT_B) |> 
-  fix_altitude_heuristic() |> 
-  # assign leg for each airborne segment
-  identify_trajectory_legs() |> 
-  # force single FLTID here to construct niche UIDs
-  group_by(ICAO24, LEG) |> fix_invalid_fltid() |> ungroup() |> 
-  # construct UID
-  mutate(UID = paste(ICAO24, LEG, sep = "-"))
+  with_ldg_time <-  arrs |>
+    dplyr::mutate(
+      study_trj = pmap(
+              .l = list(data, UID, RWY)
+            , .f = ~ add_landing_time_to_nested_data(..1, ..2, ..3, rwys)
+            )
+      ) |> 
+   # dplyr::mutate(study_trj = map(.x = study_trj, .f = ~ interpolate_lin_4D(.x, .source = "OSN"))) |> 
+    dplyr::mutate(ALDT = map(.x = study_trj, .f = ~ filter(.x, MST == "ALDT") |> pull(TIME))) |> tidyr::unnest(ALDT)
+
+  return(bind_rows(with_ldg_time, rest))
+}
+
+(
+  mickey <- gotcha |> test_landing_time()
+)
+```
+
+```{r}
+study_sequences <- mickey |> filter(ARR, !ORPHAN) |> select(UID, RWY, ALDT) |> 
+  arrange(RWY, ALDT, UID) |> 
+  group_by(RWY) |> 
+  mutate(LEAD_DELTA = difftime(ALDT, lag(ALDT, default = first(ALDT)), units = "sec"))
+study_sequences
+```

### 2023-04-01 R/add_landing_time_to_nested_data.R

@@ -1,0 +1,10 @@
+add_landing_time_to_nested_data <- function(.trj_arr, .uid, .thr, .rwys){
+#add_landing_time_to_nested_data <- function(.input_list, .rwys){
+  message(paste0("processing ", .uid , " with ", .thr))
+  this_trj <- .trj_arr |>
+    dplyr::mutate(LUID = .uid, RWY = .thr) |>
+    dplyr::group_by(LUID) |>
+    dplyr::group_modify(.f = ~assign_threshold_time2(.x, .rwys) ) |>
+    dplyr::ungroup() |>
+    dplyr::select(-LUID)
+}

### 2023-04-01 R/assign_threshold_time2.R

@@ -1,0 +1,68 @@
+#' Determine threshold time for arrival trajectories
+#' REVISED VERSION TO WORK WITH REPLICATE WORKFLOW
+#'
+#' Work out homing and coasting of arrival trajectories. Identify overhead
+#' threshold or ahead of threshold in case of low coverage.
+#' Estimate time over threshold based on most recent position updates (and
+#' derived approach speed).
+#'
+#' @param .trj_latlon_sf 
+#' @param .my_thr 
+#' @param ... 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+assign_threshold_time2 <- function(.trj_latlon_sf, .rwys, ...){
+  trj_arr <- .trj_latlon_sf
+  my_thr  <- .rwys |> dplyr::filter(REF == unique(trj_arr$RWY))
+  # trim payload and add sequence id
+  trj_arr <- trj_arr |> # select(UID, TIME, ICAO24, LAT, LON, ALT_B)  |> 
+    rowid_to_column("NODE_ID") |> 
+    mutate(MST = NA_character_) 
+  
+  if(!is(trj_arr, "sf")){
+  # stop("Y has to be a spatial object ")
+    trj_arr <- trj_arr |> cast_latlon_to_pts(.drop_coord = FALSE)
+  }  
+  
+  # ldg_rwy <- trj_arr |> 
+  #   assign_landing_runway(.arp = arp) |> 
+  #   inner_join(rwys, by = "REF")
+  # my_thr  <- ldg_rwy |> select(UID, REF, LAT, LON, COMMENT) |>
+  my_thr <- my_thr |> 
+    cast_latlon_to_pts(.drop_coord = FALSE)
+
+  # # check if we have a threshold overflight
+  ovr_thr <- identify_threshold_overhead(trj_arr, my_thr)
+  if(nrow(ovr_thr) > 0){
+    message("THRESHOLD")
+    thr_row <- ovr_thr |> filter(DIST_THR == min(DIST_THR))
+    trj_arr[thr_row$NODE_ID,"MST"] <- "ALDT"
+    trj_arr[thr_row$NODE_ID,"MST_REF"] <- my_thr$REF
+  }else{
+     message("NEED TO WORK ON LANDING TIME")
+    thr_row <- trj_arr |>
+      identify_threshold_ahead2(my_thr) |>
+      estimate_threshold_time(my_thr)
+    
+    if(nrow(thr_row) < 1) { # ============= catch error ----------------
+      message("NO THR")
[…]

### 2023-04-01 R/identify_threshold_overhead.R

@@ -20,3 +20,9 @@ identify_threshold_ahead <- function(.final, .my_thr ,...){
   #   tidyr::fill(UID, ICAO24) |> 
   #   trj_add_target_bearing(.my_thr)
 }
+
+identify_threshold_ahead2 <- function(.final, .my_thr ,...){
+  trj <- .final |> 
+    dplyr::filter(ALT_B > .my_thr$ELEV + 500) |> 
+    trj_add_target_bearing(.my_thr)
+}

### 2023-04-02 x_replicate.qmd
added tesselation of study data with H3

@@
 source("~/RProjects/paper-2023-ICNS/R/add_landing_time_to_nested_data.R")
+
+# set ggplot theme
+ggplot2::theme_set(ggplot2::theme_minimal())
 ```
@@
   with_ldg_time <-  arrs |>
+    # estimate and assign landing time to arrival trajectories
     dplyr::mutate(
@@
       ) |> 
-   # dplyr::mutate(study_trj = map(.x = study_trj, .f = ~ interpolate_lin_4D(.x, .source = "OSN"))) |> 
-    dplyr::mutate(ALDT = map(.x = study_trj, .f = ~ filter(.x, MST == "ALDT") |> pull(TIME))) |> tidyr::unnest(ALDT)
+    # pullout ALDT as separate column
+    dplyr::mutate(
+      ALDT = map(.x = study_trj
+               , .f = ~ filter(.x, MST == "ALDT") |> pull(TIME)
+               )
+      ) |> tidyr::unnest(ALDT) |> 
+    # interpolate to 5sec
+    dplyr::mutate(
+      study_trj = map(.x = study_trj
+                    , .f = ~ interpolate_lin_4D(.x, .source = "OSN") |> 
+                             add_time_and_distance_to_go() |> 
+                             dplyr::mutate(H3_BIN = h3::geo_to_h3(c(LAT, LON), res = 8))
+                    )
+      )
 
@@
 ```
+
+```{r}
+study_sequences |> filter(LEAD_DELTA < 500) |> 
+  ggplot() +
+  geom_boxplot(aes(x = RWY, y = LEAD_DELTA))
+```
+
+```{r}
+study_sequences <- mickey |> filter(ARR, !ORPHAN) |> select(UID, RWY, ALDT) |> 
+  arrange(RWY, ALDT, UID) |> 
+  group_by(RWY) |> 
+  mutate(
+     LEAD_DELTA = difftime(ALDT, lag(ALDT, default = first(ALDT)), units = "sec")
+    ,INTERACT   = LEAD_DELTA < 240
+    ,LEADER     = ifelse(INTERACT, lag(UID), NA)) |> 
+  ungroup()
+study_sequences
+```
+
+write-out tesselation data for reference times
+
+```{r}
+tessel_fn <- here::here("data", "EGLL-tessel-db.csv.gz")
+
+extract_tessel_info <- function(.study_trj, ...){
+  df <- .study_trj
+  df <- df |> dplyr::select(H3_BIN, TIME_2GO, DIST_2GO )
+}
+
+make_tessel_db_entry <- function(.study_trj, .ptn = ptn, ...){
+  df <- .study_trj |> 
+    dplyr::select(UID, RWY, study_trj) |> 
+    dplyr::mutate(UID = paste(UID, gsub("-", "", .ptn), sep = "-")) |> 
+    dplyr::mutate(
+      info = purrr::map(
+                .x = study_trj
+              , .f = ~ extract_tessel_info(.x)
+              )
+      ) |> 
+    tidyr::unnest(info)
+  return(df)
+}
+```
+
+```{r}
+tessel_info <- make_tessel_db_entry(mickey |> filter(ARR, !ORPHAN))
+tessel_info
+```
+
+```{r}
+ti <- tessel_info |> ungroup() |> 
+  select(-study_trj) |> 
+  group_by(H3_BIN) |> summarise(N = n(), .groups = "drop") |> 
+  mutate(
+    # add geometry column without naming column (easier than fixing it afterwards)
+    h3::h3_to_geo_boundary_sf(H3_BIN)
+    ) |> 
+  sf::st_as_sf()
+
+p1 <- ggplot() +
+  geom_sf(data = ti, aes(fill = N), color = NA) + 
+  scale_fill_distiller(palette = "Spectral")
+
+leftright <- 0.3
+p2 <- p1 +
+  coord_sf( xlim = c(arp_egll$LON - leftright, arp_egll$LON + leftright)
+           ,ylim = c(arp_egll$LAT - leftright, arp_egll$LAT + leftright)
+           )
+
+library(patchwork)
+p1 + p2 +
+  plot_layout(guides = 'collect') &
+  theme(legend.position = "top")
+```

### 2023-04-02 R/add_distance_rowwise.R

@@ -30,6 +30,18 @@ add_flight_distance <- function(.trj, unit = "NM"){
   return(df)
 }
 
+add_time_and_distance_to_go <- function(.trj, .mst = "ALDT", .time_unit = "sec"){
+  df <- .trj
+  
+  df <- df |> 
+    add_flight_distance() |> 
+    dplyr::mutate(
+      TIME_2GO = difftime(
+          TIME[which(MST == .mst)], TIME, units = .time_unit)
+    , DIST_2GO = DIST_FLOWN[which(MST == .mst) - DIST_FLOWN])
+  
+  return(df)
+}
 
 # =================== sf based solution ==========================
 # library(sf)

### 2023-04-02 R/interpolate_lin_4D.R

@@ -1,0 +1,35 @@
+#' Simple linear interpolation of 4D positions (e.g. CPF/CPR)
+#'
+#' @param .cpf
+#' @param .source
+#'
+#' @return
+#' @export
+#'
+#' @examples
+interpolate_lin_4D <- function(.cpf, .time_step = "5 sec", .source = "CPF"){
+  df <- .cpf  # to-do check and test
+
+  # #----------- set UID, if not existing
+  # uid_exists <- colnames(df) %in% "UID" %>% any()
+  # if(!uid_exists) df <- df %>% dplyr::mutate(UID = SAM_ID)
+
+  min_max_time = df %>%
+    dplyr::filter(dplyr::row_number() %in% c(1, n())) %>%
+    dplyr::summarise(start = dplyr::first(TIME), end = dplyr::last(TIME)) %>%
+    dplyr::transmute(TIME = purrr::map2(start, end, seq, by = .time_step)) %>%
+    tidyr::unnest(cols = TIME)
+
+  df2 <- min_max_time %>%
+    dplyr::full_join(df %>% dplyr::mutate(SOURCE = .source), by = "TIME") %>%
+    dplyr::arrange(TIME) %>%
+    #tidyr::fill(UID, .direction = "down") %>%
+    # mutate(across(.cols = c(ALT), .fns = ~zoo::na.approx, seq, na.rm=FALSE))
+    dplyr::mutate(
+       ALT_B  = zoo::na.approx(ALT_B, x = TIME, na.rm = FALSE)
+      ,LAT = zoo::na.approx(LAT, x = TIME, na.rm = FALSE)
+      ,LON = zoo::na.approx(LON, x = TIME, na.rm = FALSE)
+      ,SOURCE = if_else(is.na(SOURCE), "INP", SOURCE)
+    )
+  return(df2)
+}

### 2023-04-02 R/trj_confirm_arrival_at.R

@@ -21,7 +21,7 @@ confirm_arrival_at <- function(.df, .apt_box, .time_window = 500, ...){
     dplyr::mutate(ARR = map(
       .x = last_4d
       ,.f = ~ cast_latlon_to_ls(.x) %>%
-        st_intersects(.apt_box, sparse = FALSE))
+        sf::st_intersects(.apt_box, sparse = FALSE))
     ) %>%
     tidyr::unnest(ARR) |> 
     dplyr::mutate(N_LAST = map(last_4d, nrow)) |> 

### 2023-04-02 x_replicate.qmd
moving towards spacing deviation

@@
 tessel_info <- make_tessel_db_entry(mickey |> filter(ARR, !ORPHAN))
+
 tessel_info
@@
 ```
+
+```{r}
+if(file.exists(tessel_fn)) { message("Open existing tessel database")
+}else{
+  message("Create tessel database")
+    tessel_info |> 
+    dplyr::select(H3_BIN, RWY, TIME_2GO, DIST_2GO, UID ) |> 
+    dplyr::ungroup() |> 
+    readr::write_csv(tessel_fn)
+  }
+```
+
+t-digest (for paper to DASC?)
+https://git.sr.ht/~hrbrmstr/tdigest
+
+```{r}
+tessel_lookup <- readr::read_csv(tessel_fn, show_col_types = FALSE) |> 
+  group_by(H3_BIN, RWY) |> 
+  summarise( TIME_MIN = min(TIME_2GO), TIME_Q5 = quantile(TIME_2GO, probs = 0.05)
+            , DIST_MIN= min(DIST_2GO), DIST_Q5 = quantile(DIST_2GO, probs = 0.05)
+            , SAMPLE_N = n())
+```
+
+```{r}
+study_sequences <- mickey |> filter(ARR, !ORPHAN) |> select(UID, RWY, ALDT) |> 
+  arrange(RWY, ALDT, UID) |> 
+  group_by(RWY) |> 
+  mutate(LEAD_DELTA = difftime(ALDT, lag(ALDT, default = first(ALDT)), units = "sec"))
+study_sequences
+```
+
+```{r}
+tmp <- mickey |> filter(UID %in% study_sequences$UID[1:4]) |> 
+  select(UID, RWY2 = RWY, ALDT, study_trj) |> 
+  unnest(study_trj)
+
+ggplot(data = tmp) +
+  geom_point(aes(x = TIME, y = ALT_B))
+```
+
+spacing deviation (t) = min time (trailer (t)) –    min time (leader (t – s))
+
+```{r}
+add_bin_mins <- function(.trj, .bins){
+  trj <- .trj %>%
+    left_join(.bins %>% select(H3_BIN, RWY, MIN_DIST_2GO, MIN_TIME_2GO)
+              , by = c("RWY", "H3_BIN"))
+  return(trj)
+}
+
+calc_pairwise_spacing <- function(.uid, .seq_uid, .t_trail , .ds, .bins = bins){
+
+  leader <- .ds %>% filter(UID == .uid) %>% 
+    add_bin_mins(.bins) %>%
+    select(UID, TIME, RWY, DIST_2GO, TIME_2GO, MIN_DIST_2GO, MIN_TIME_2GO)
+
+  trailer<- .ds %>% filter(UID == .seq_uid) %>% 
+    add_bin_mins(.bins) %>%
+    select(UID, TIME, RWY, DIST_2GO, TIME_2GO, MIN_DIST_2GO, MIN_TIME_2GO) %>% 
+    mutate(TIME_WARP = TIME - .t_trail)  %>%
+    rename(SEQ_UID = UID, SEQ_DIST_2GO = DIST_2GO, SEQ_TIME_2GO = TIME_2GO
+          ,SEQ_MIN_DIST_2GO = MIN_DIST_2GO, SEQ_MIN_TIME_2GO = MIN_TIME_2GO)
+
+  spacedev <- leader %>% 
+    left_join(trailer, by = c("TIME" = "TIME_WARP", "RWY")) %>% 
+    mutate(MIN_TIME_OFF = MIN_TIME_2GO - SEQ_MIN_TIME_2GO
+         , MIN_DIST_OFF = MIN_DIST_2GO - SEQ_MIN_DIST_2GO
+         , DIST_OFF     = DIST_2GO - SEQ_DIST_2GO
+         , STEP = row_number()) %>% 
+    select(UID, SEQ_UID, STEP, MIN_TIME_OFF, MIN_DIST_OFF, DIST_OFF, everything())
+
+return(spacedev)
+}
+
+#' Smoothing function 
+#' wrapper for slider::slide()
+
+roll_median <- function(.df, .var, .whs, .target_var){
+  df <- .df %>% 
+    mutate(
+     {{.target_var}} := slider::slide_dbl({{.var}}, median, .before = .whs, .after = .whs)
+  )
+}
+
+```

### 2023-04-03 paper.qmd
pipeline in place ... 2 weeks too late ;)

@@
 Despite the introduction of an initial market-based mechanism, immediate action to curb fuel burn and CO2 emissions rests with improvements of operational efficiency.
-<!-- verbatim -->
-ICAO has introduced the trajectory-based operations concept, which promises operational and environmental benefits to aviation. 
+<!-- verbatim --> ICAO has introduced the trajectory-based operations concept, which promises operational and environmental benefits to aviation.
 where the flight trajectory of an aircraft is flown as close as possible to the userpreferred route, with as little disruptions as possible through collaborative decision-making mechanism.
 This includes reducing potential conflicts and resolving demand/capacity imbalances earlier and more efficiently.
-TBO can therefore bring significant operational and environmental benefits to aviation.
+TBO can therefore bring significant operational and environmental benefits to aviation.\
 <!-- end verbatim copy from ICAO paper -->
@@
 
-## Arrival Sequencing
+## Generalisation of additional time in terminal airspace
 
-Spacing deviation
+The additional time in terminal airspace (c.f.
+ICAO KPI08, EUROCONTROL Performance Review System & European Single European Sky Performance Scheme) is typically expressed as the difference between the observed travel time of a flight entering the terminal area (e.g.
+40NM or 100NM from the aerodrome reference point) and a respective reference time.
+The latter is defined as the travel time in non-congested situations.
+For practical reasons, the reference (or uncongested) travel times are determined based on a historical data analysis and sub-sampling the population of arrival flights per entry sector/fix, aircraft wake turbulence category and propulsion type (as a proxy of approach speed), and landing runway.
 
-Let us consider a pair of consecutive landing aircraft denoted leader and trailer, with s their required time spacing1.
-Using the constant time delay principle, the spacing deviation (or spacing error) at time t considers the current position of trailer at time t, and the past position of leader at time t -- s.
-Precisely, it is defined as the difference between the respective minimum times from these two positions (see figure below): spacing deviation (t) = min time (trailer (t)) -- min time (leader (t -- s))
+This notion of additional time can be generalised for every subset of points within the arrival airspace.
+On the basis of historic data, the travel time between such a set of points can be determined and compared to the flight time of an uncongested trajectory.
+
+```{r}
+#| label: fig-spacing-deviation-concept
+#| fig-cap: Spacing deviation concept
+#| out-width: 45%     # 2 column half-size
+knitr::include_graphics("./figures/concept.png")
+```
+
+For practical reasons, this paper discretises the arrival airspace into cells.
+In particular, we are using hexogonal cells and the H3-indexing system [@ubertechnologiesinc2018].
+Uber originally developed the Hexagonal Hierarchical Geospatial Indexing System (H3) as a grid system for supporting the optimisation of ride planning (and associated pricing and dispatch) and for visualisation purposes.
+H3 is available as an open-source library written in C and with bindings in several languages.
+We use hexagon-cells of resolution 8 with an average edge length of 0.5314 km.
+
+## Arrival Sequencing - Spacing Deviation
+
+Let us consider a pair of consecutive landing aircraft denoted leader and trailer, with $s$ being their temporal spacing (inter-arrival time).
+From an operational perspective, highly efficient operations will maximise the utilisation of the available runway (system) capacity.
+Accordingly, we can assume that during peak times the temporal spacing at landing $s$ accounts for the operational concept applied at the airport, and only comprises a minimal spacing error.
+Potential error sources include additional safety margins applied by the air traffic controller (or aircrew when landing under visual flight rules), variations of the arrival spacing due to reduction of airspeed during the flare.
+We consider the observed temporal spacing $s_{ij}$ as a lower bound for safe operations.
+
+Using the constant time delay principle, the spacing deviation (or spacing error) at time t considers the current position of trailer at time $t$, and the past position of leader at time $t-s$.
+Based on our approach, the spacing deviation is defined as the temporal difference between the respective reference times for the position of the leader and trailer:
+
+$$spacing deviation (t) = ref{\_}time(trailer_{(t)}) - ref{\_}time(leader_{(t - s)})$$
 

### 2023-04-03 x_replicate.qmd
pipeline in place ... 2 weeks too late ;)

@@
 source("~/RProjects/paper-2023-ICNS/R/altitude_outlier_filter.R")
+source("~/RProjects/paper-2023-ICNS/R/trj_identify_legs.R")
 
@@
 source("~/RProjects/paper-2023-ICNS/R/add_landing_time_to_nested_data.R")
+source("~/RProjects/paper-2023-ICNS/R/trj_landing_runway.R")
+
+source("~/RProjects/paper-2023-ICNS/R/add_distance_rowwise.R")
+source("~/RProjects/paper-2023-ICNS/R/interpolate_lin_4D.R")
 
@@
     extract_4d_nest() |> 
-    confirm_arrival_at(apt_box) |> 
+    confirm_arrival_at(apt_box, rwy_centerline_ls) |> 
     check_arrival_orphan()
@@
 ```{r}
+apt <- "EGLL"
+aip <- readr::read_csv("./data/aip-arp-rwys.csv", show_col_types = FALSE) |> 
+    dplyr::filter(ICAO == apt)
+rwys_egll <- aip |> filter(REF != "ARP") |> rename(ELEV = GEOM_ALT)
+
 test_landing_time <- function(phase, rwys = rwys_egll){
@@
 ```{r}
+# check for date of loaded data
+ptn <- "2019-08-01"
+
 tessel_fn <- here::here("data", "EGLL-tessel-db.csv.gz")
@@
 ```{r}
-study_sequences <- mickey |> filter(ARR, !ORPHAN) |> select(UID, RWY, ALDT) |> 
-  arrange(RWY, ALDT, UID) |> 
-  group_by(RWY) |> 
-  mutate(LEAD_DELTA = difftime(ALDT, lag(ALDT, default = first(ALDT)), units = "sec"))
 study_sequences
@@
 ```{r}
-add_bin_mins <- function(.trj, .bins){
-  trj <- .trj %>%
-    left_join(.bins %>% select(H3_BIN, RWY, MIN_DIST_2GO, MIN_TIME_2GO)
-              , by = c("RWY", "H3_BIN"))
-  return(trj)
-}
+source("~/RProjects/paper-2023-ICNS/pairwise_spacing.R")
+```
 
-calc_pairwise_spacing <- function(.uid, .seq_uid, .t_trail , .ds, .bins = bins){
-
-  leader <- .ds %>% filter(UID == .uid) %>% 
-    add_bin_mins(.bins) %>%
-    select(UID, TIME, RWY, DIST_2GO, TIME_2GO, MIN_DIST_2GO, MIN_TIME_2GO)
-
-  trailer<- .ds %>% filter(UID == .seq_uid) %>% 
-    add_bin_mins(.bins) %>%
-    select(UID, TIME, RWY, DIST_2GO, TIME_2GO, MIN_DIST_2GO, MIN_TIME_2GO) %>% 
-    mutate(TIME_WARP = TIME - .t_trail)  %>%
-    rename(SEQ_UID = UID, SEQ_DIST_2GO = DIST_2GO, SEQ_TIME_2GO = TIME_2GO
-          ,SEQ_MIN_DIST_2GO = MIN_DIST_2GO, SEQ_MIN_TIME_2GO = MIN_TIME_2GO)
-
-  spacedev <- leader %>% 
-    left_join(trailer, by = c("TIME" = "TIME_WARP", "RWY")) %>% 
-    mutate(MIN_TIME_OFF = MIN_TIME_2GO - SEQ_MIN_TIME_2GO
-         , MIN_DIST_OFF = MIN_DIST_2GO - SEQ_MIN_DIST_2GO
-         , DIST_OFF     = DIST_2GO - SEQ_DIST_2GO
-         , STEP = row_number()) %>% 
-    select(UID, SEQ_UID, STEP, MIN_TIME_OFF, MIN_DIST_OFF, DIST_OFF, everything())
-
-return(spacedev)
-}
+```{r}
 
@@
 ```
+
+do for 1 pair
+```{r}
+rq <- my_pairwise_spacing("400550-0", "89505a-0",.t_trail = 134, .bins = tessel_lookup)
+
+# can we smooth the results ======================
+#' Smoothing function 
+#' wrapper for slider::slide()
+
+roll_median <- function(.df, .var, .whs, .target_var){
+  df <- .df %>% 
+    mutate(
+     {{.target_var}} := slider::slide_dbl({{.var}}, median, .before = .whs, .after = .whs)
+  )
+} #-----------------------------------------------
+whs <- 45
+rqq <- rq |> group_by(UID, SEQ_UID, RWY) %>% 
+  roll_median(MIN_TIME_OFF, whs, SPACING) |> 
+  filter(STEP < 2500)
+
+ggplot() + 
+  geom_line(data = rq,  aes(x = STEP, y = MIN_TIME_OFF), alpha = 0.5) +
+  geom_line(data = rqq, aes(x = STEP, y = SPACING), color = "blue")
+```
+
+do for more couples
+
+```{r}
+couples <- study_sequences |> 
+  group_by(RWY) |> 
+  filter(!is.na(LEADER), LEAD_DELTA <= 300) |> 
+  ungroup()
+couples
+
+sdp <- couples |> 
+  dplyr::select(LEAD_UID = LEADER, TRAIL_UID = UID, T_TRAIL = LEAD_DELTA) |> 
+  tibble::rowid_to_column("GID")
+sdp
+
+sdp_res <- sdp |> group_by(GID) |> 
+  purrr::pmap(.f = ~ my_pairwise_spacing(..2, ..3, ..4, .bins = tessel_lookup)) |> 
+  bind_rows() |> 
+  tidyr::fill(SEQ_UID, .direction = "up")
+#
+write_csv(sdp_res, "./data/spd-res-EGLL-20190801.csv.gz")
+```
+
+```{r}
+sdp_smooth <- sdp_res |> group_by(UID, SEQ_UID, RWY) %>% 
+  roll_median(MIN_TIME_OFF, whs, SPACING) |> 
+  filter(STEP < 2500) 
+sdp_smooth <- sdp_smooth |> mutate(GID = dplyr::cur_group_id())
+
+ggplot() + 
+  geom_line(  data = sdp_res |> group_by(UID,SEQ_UID) |> 
+                              mutate(GID = dplyr::cur_group_id())
+            , aes(x = STEP, y = MIN_TIME_OFF, group = GID), alpha = 0.2) +
+  geom_line(data = sdp_smooth, aes(x = STEP, y = SPACING, group = GID), color = "blue", alpha = 0.5)
+```

### 2023-04-03 R/trj_confirm_arrival_at.R

@@ -10,7 +10,7 @@
 #' @export
 #'
 #' @examples
-confirm_arrival_at <- function(.df, .apt_box, .time_window = 500, ...){
+confirm_arrival_at <- function(.df, .apt_box, .rwy_centerline_ls = rwy_centerline_ls, .time_window = 500,...){
   df <- .df %>%
     # extract last (aka oldest) pos reports
     dplyr::mutate(last_4d = map(
@@ -32,7 +32,7 @@ confirm_arrival_at <- function(.df, .apt_box, .time_window = 500, ...){
             .x = last_4d
           , .f = ~ estimate_rwy(
                     .x |> cast_latlon_to_pts()
-                  , rwy_centerline_ls
+                  , .rwy_centerline_ls
                   ) |> 
             arrival_runway_assignment()
           )

### 2023-04-03 pairwise_spacing.R

@@ -1,0 +1,83 @@
+my_pairwise_spacing <- function(.lead_uid, .trail_uid, .t_trail, .mickey = mickey, .bins,...){
+  
+  leader <- .mickey |> 
+    dplyr::filter(UID == .lead_uid) |> 
+    dplyr::select(UID, RWY2 = RWY, study_trj) |> 
+    tidyr::unnest(study_trj) |> 
+    add_bin_mins(.bins) |> 
+    dplyr::select(UID, RWY, TIME, TIME_2GO, DIST_2GO, H3_BIN, TIME_MIN, TIME_Q5, DIST_MIN, DIST_Q5)|> 
+    interpolate_lin_spacdev()
+  
+  trailer <- .mickey |> 
+    dplyr::filter(UID == .trail_uid) |> 
+    dplyr::select(UID, RWY2 = RWY, study_trj) |> 
+    tidyr::unnest(study_trj) |> 
+    add_bin_mins(.bins) |> 
+    dplyr::select(UID, RWY, TIME, TIME_2GO, DIST_2GO, H3_BIN, TIME_MIN, TIME_Q5, DIST_MIN, DIST_Q5) |> 
+    interpolate_lin_spacdev()
+  
+  # do time-warp for trailer and
+  # rename variables
+  trailer <- trailer |>
+    dplyr::mutate(TIME_WARP = TIME - .t_trail) |>
+    dplyr::rename(
+        SEQ_UID = UID, SEQ_DIST_2GO = DIST_2GO, SEQ_TIME_2GO = TIME_2GO
+       ,SEQ_TIME = TIME
+       ,SEQ_H3_BIN = H3_BIN
+       ,SEQ_MIN_DIST_2GO   = DIST_MIN, SEQ_MIN_TIME_2GO   = TIME_MIN
+       ,SEQ_MIN_DIST_2GO_Q = DIST_Q5,  SEQ_MIN_TIME_2GO_Q = TIME_Q5
+       )
+  
+  # join and calc spacing 
+  spacedev <- leader |>
+    dplyr::left_join(trailer, by = c("TIME" = "TIME_WARP", "RWY")) |> 
+    dplyr::mutate(  
+             MIN_TIME_OFF   = TIME_MIN - SEQ_MIN_TIME_2GO
+           , MIN_TIME_OFF_Q = TIME_Q5  - SEQ_MIN_TIME_2GO_Q
[…]

### 2023-04-04 paper.qmd
slowly getting going

@@
 abstract: |
-  This document is a sample illustrating the Quarto `ieeetran` template.
-  It includes the key elements of a scientific articles 
-  (references, equations, figures, tables, code, cross references).
-  The template enables the generation of IEEE-formatted article from a Jupypter notebook.
+  In light of the on-going discussion on climate change, operational efficiency of air navigation has gained a higher visibility.
+  Efficient operations will positively influence the fuel burn by airspace users. 
+  While the development and introduction of novel aircraft airframe and propulsion technologies, 
+  the global pickup rate of sustainable aviation fuel, and benefits mechanism of market-based measures promise a positive effect
+  over time, >> reducing unnecessary contraints on airspace users can be immediately deployed. 
+  There is still a considerable benefit pool for arrival operations. 
+  This proof-of-concept paper explores the use of open data for the analysis of operational efficiency within the extended terminal airspace.
+  To characterise the potentail benefit pool, a method to quantify the spacing deviation between successive flights is developed.
+  This paper summarises the conceptual approach and reports on the first findings of an use-case analysis for a subset of European airports.
+  The results show the general feasibility of the approach and help to identify requirements for the further development of operational performance measures. 
 keywords: [template, demo]
@@
 
-# Data and Conceptual Approach
+# Conceptual Approach and Data
+
+## Research Approach
+
+This paper follows an exploratory research process (c.f. @fig-research-approach).
+We frame the work as a operational performance analytical problem.
+
+```{r}
+#| label: fig-research-approach
+#| fig-cap: Exploratory research approach
+#| out-width: 45%     # 2 column half-size
+knitr::include_graphics("./figures/exploratory-research-concept.png")
+```
 
@@
 
-$$spacing deviation (t) = ref{\_}time(trailer_{(t)}) - ref{\_}time(leader_{(t - s)})$$
+$$\begin{aligned}
+spacing \text{ } deviation (t) =  ref{\_}time(trailer_{(t)}) - \\ref{\_}time(leader_{(t - s)})
+\end{aligned}$${#eq-spacing-deviation}
+
+```{r}
+#| label: fig-follower-leader-example
+#| fig-cap: Example timeline of spacing deviations between interacting arrivals
+#| out-width: 45%     # 2 column half-size
+knitr::include_graphics("./figures/example-follower-leader-pairs.png")
+```
+
+@eq-spacing-deviation does something.
 

### 2023-04-04 study_pipeline.qmd
slowly getting going

---
title: "Industrialise Pipeline"
format: pdf
---

```{r}
list_my_functions <- list.files(path = "./R", pattern = "\\.R", full.names = TRUE)
list_my_functions |> purrr::walk(.f = ~ source(.x))
```

## Overview

This Quarto document provides a staged approach to generating the data for the ICNS2023 paper.

1. load and extract test data for the studied European airports
2. .... generate mickey & store interim results
3. ... do results

## Load and extract study data

```{r}
#| message: false
load_osn_dumps <- function(.ptn, .pth_to_dumps){
    # list downloaded dumps
    fns <- list.files(path = .pth_to_dumps, pattern = .ptn, full.names = TRUE)

    # load list of chosen dumps
    ds  <- fns |> 
      purrr::map_dfr(.f = read_osn_tar_file) |> 
      make_nice_names_osn()
    ds <- ds |> 
      dplyr::mutate(
            dplyr::across(
                  .cols = c("ALT_B","ALT_G")
                , .fns  = coerce_meter_to_feet)
            ) 
    ds <- ds |> coerce_unixepoch_to_datetime()
    return(ds)
}

extract_and_save_jormungand_data <- function(
    .apt, .range = 205, .ds, .save_path = "./data-study/"
    , .save_label = ptn
    , .aip_file = "./data/aip-arp-rwys.csv", ...
    ){
  # load AIP data for study  
  aip <- read_csv(.aip_file, show_col_types = FALSE)
  # extract airport ARP
  arp <- filter(aip, ICAO == .apt, REF == "ARP") |> dplyr::rename(ELEV = GEOM_ALT)
  jormungand <- establish_radial_jormungand(arp, .range) |> 
    dplyr::mutate(REF = paste(ICAO,REF,.range, sep = "-")) |> 
    dplyr::select(ICAO, REF, geometry)

  # subset study data for range
  svs <- spatial_subset_jormungand(.ds, jormungand)

  # write out study data by airport range
  fn_out <- paste0(.save_path, .apt, "-200NM-", .save_label, ".parquet")
  message(paste0("writing ", fn_out))
  arrow::write_parquet(svs |> sf::st_drop_geometry(), fn_out)
}

extract_and_save_study_dataset_for_airport <- function(
    .study_apts
    ,.ptn, .pth_to_dumps, .range = 205, ...
){
  # get osn dump

}

```

Loaded: 
 2019-08-01, 02, 03, 04, 05, 06, 07
 2022-08-01, ... 07, 08, 09

```{r}
# European study data
pth_to_dumps <- "../__DATA/OSN-dumps"
ptn <- "2022-08-09"
study_apts <- c("EGLL","EDDM","EDDF", "LSZH","EHAM")

 osn <- load_osn_dumps(ptn, pth_to_dumps)
  #subset for study airports
  study_apts |> 
    purrr::walk( .f = ~ extract_and_save_jormungand_data(.x, .ds = osn) )
```

### 2023-04-04 x_replicate.qmd
slowly getting going

@@
 rq <- my_pairwise_spacing("400550-0", "89505a-0",.t_trail = 134, .bins = tessel_lookup) 
+rq1 <- my_pairwise_spacing("89505a-0","40054d-0", .t_trail = 122, .bins = tessel_lookup)
+rq2<- my_pairwise_spacing("71005e-0", "4007f2-0", .t_trail = 91, .bins = tessel_lookup)
+rq3 <- my_pairwise_spacing("4007f2-0","896464-0", .t_trail = 98, .bins = tessel_lookup)
+
+rq <- rq |> bind_rows(rq1, rq2, rq3) |> 
+  mutate(GID = paste(UID, SEQ_UID, RWY, sep = "-"))
 
@@
 
+rqq <- rqq |> filter(!grepl(pattern = "-NA-", x = GID), STEP < 1250)
+
 ggplot() + 
-  geom_line(data = rq,  aes(x = STEP, y = MIN_TIME_OFF), alpha = 0.5) +
-  geom_line(data = rqq, aes(x = STEP, y = SPACING), color = "blue")
+ # geom_line(data = rq,  aes(x = STEP, y = MIN_TIME_OFF, group = GID), alpha = 0.5) +
+  geom_line(data = rqq, aes(x = STEP, y = SPACING, group = GID, color = GID) ) +  #  = "blue")
+  labs(x = "time to fly [sec]", y = "spacing deviation [sec]"
+       ,color = "follower-leader-pair") +
+  theme(legend.position = c(0.2,0.2))
 ```

### 2023-04-04 R/spatial_subset_jormungand.R

@@ -25,3 +25,8 @@ spatial_subset_jormungand <- function(.df_latlon, .outer_poly, ...){
   }
    df <- df[.outer_poly, ]
 }
+
+establish_radial_jormungand <- function(.center_latlon, .radius_NM = 205, ...){
+  center <- cast_latlon_to_pts(.center_latlon)
+  circle <- center |> sf::st_buffer(dist = .radius_NM * 1852, nQuadSegs = 60 )
+}

### 2023-04-04 paper.tex
push .tex

@@
 \begin{abstract}
-    This document is a sample illustrating the Quarto \texttt{ieeetran}
-    template. It includes the key elements of a scientific articles
-    (references, equations, figures, tables, code, cross references).
-    The template enables the generation of IEEE-formatted article from a
-    Jupypter notebook.
+    In light of the on-going discussion on climate change, operational
+    efficiency of air navigation has gained a higher visibility.
+    Efficient operations will positively influence the fuel burn by
+    airspace users. While the development and introduction of novel
+    aircraft airframe and propulsion technologies, the global pickup
+    rate of sustainable aviation fuel, and benefits mechanism of
+    market-based measures promise a positive effect over time,
+    \textgreater\textgreater{} reducing unnecessary contraints on
+    airspace users can be immediately deployed. There is still a
+    considerable benefit pool for arrival operations. This
+    proof-of-concept paper explores the use of open data for the
+    analysis of operational efficiency within the extended terminal
+    airspace. To characterise the potentail benefit pool, a method to
+    quantify the spacing deviation between successive flights is
+    developed. This paper summarises the conceptual approach and reports
+    on the first findings of an use-case analysis for a subset of
+    European airports. The results show the general feasibility of the
+    approach and help to identify requirements for the further
+    development of operational performance measures.
 \end{abstract}
@@
 goal for international aviation
-\protect\hyperlink{ref-icao2022}{{[}1{]}}. In support to the Paris
-Agreement, the goal is to achieve net-zero carbon emissions by 2050.
+\protect\hyperlink{ref-icao2022}{\textbf{icao2022?}}. In support to the
+Paris Agreement, the goal is to achieve net-zero carbon emissions by
+2050.
 
@@
 action to curb fuel burn and CO2 emissions rests with improvements of
-operational efficiency.
+operational efficiency. ICAO has introduced the trajectory-based
+operations concept, which promises operational and environmental
+benefits to aviation. where the flight trajectory of an aircraft is
+flown as close as possible to the userpreferred route, with as little
+disruptions as possible through collaborative decision-making mechanism.
+This includes reducing potential conflicts and resolving demand/capacity
+imbalances earlier and more efficiently. TBO can therefore bring
+significant operational and environmental benefits to aviation.\\
 
@@
 
-\hypertarget{data-and-conceptual-approach}{%
-\section{Data and Conceptual
-Approach}\label{data-and-conceptual-approach}}
+\hypertarget{conceptual-approach-and-data}{%
+\section{Conceptual Approach and
+Data}\label{conceptual-approach-and-data}}
+
+\hypertarget{research-approach}{%
+\subsection{Research Approach}\label{research-approach}}
+
+This paper follows an exploratory research process (c.f.
+Figure~\ref{fig-research-approach}). We frame the work as a operational
+performance analytical problem.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{./figures/exploratory-research-concept.png}
+
+}
+
+\caption{\label{fig-research-approach}Exploratory research approach}
+
+\end{figure}
 
@@
 phases, e.g.~recent machine learning algorithm
-\protect\hyperlink{ref-sun2017flightphase}{{[}2{]}}. This paper
+\protect\hyperlink{ref-sun2017flightphase}{{[}1{]}}. This paper
 implements a heuristic approach with a focus on the detection of arrival
@@
 
-\hypertarget{arrival-sequencing}{%
-\subsection{Arrival Sequencing}\label{arrival-sequencing}}
+\begin{figure}
 
-Spacing deviation
+{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{./figures/KATL-traffic-single-day.png}
+
+}
+
+\caption{\label{fig-KATL-arrivals}Arrivals at Atlanta Hartsfiled (KATL)
+on sample day}
+
+\end{figure}
+
+\hypertarget{generalisation-of-additional-time-in-terminal-airspace}{%
+\subsection{Generalisation of additional time in terminal
+airspace}\label{generalisation-of-additional-time-in-terminal-airspace}}
+
+The additional time in terminal airspace (c.f. ICAO KPI08, EUROCONTROL
+Performance Review System \& European Single European Sky Performance
+Scheme) is typically expressed as the difference between the observed
+travel time of a flight entering the terminal area (e.g.~40NM or 100NM
+from the aerodrome reference point) and a respective reference time. The
+latter is defined as the travel time in non-congested situations. For
+practical reasons, the reference (or uncongested) travel times are
+determined based on a historical data analysis and sub-sampling the
+population of arrival flights per entry sector/fix, aircraft wake
+turbulence category and propulsion type (as a proxy of approach speed),
+and landing runway.
+
+This notion of additional time can be generalised for every subset of
+points within the arrival airspace. On the basis of historic data, the
+travel time between such a set of points can be determined and compared
+to the flight time of an uncongested trajectory.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{./figures/concept.png}
+
+}
+
+\caption{\label{fig-spacing-deviation-concept}Spacing deviation concept}
+
+\end{figure}
+
+For practical reasons, this paper discretises the arrival airspace into
+cells. In particular, we are using hexogonal cells and the H3-indexing
+system \protect\hyperlink{ref-ubertechnologiesinc2018}{{[}2{]}}. Uber
+originally developed the Hexagonal Hierarchical Geospatial Indexing
+System (H3) as a grid system for supporting the optimisation of ride
+planning (and associated pricing and dispatch) and for visualisation
+purposes. H3 is available as an open-source library written in C and
+with bindings in several languages. We use hexagon-cells of resolution 8
+with an average edge length of 0.5314 km.
+
+\hypertarget{arrival-sequencing---spacing-deviation}{%
+\subsection{Arrival Sequencing - Spacing
+Deviation}\label{arrival-sequencing---spacing-deviation}}
 
 Let us consider a pair of consecutive landing aircraft denoted leader
-and trailer, with s their required time spacing1. Using the constant
-time delay principle, the spacing deviation (or spacing error) at time t
-considers the current position of trailer at time t, and the past
-position of leader at time t -- s. Precisely, it is defined as the
-difference between the respective minimum times from these two positions
-(see figure below): spacing deviation (t) = min time (trailer (t)) --
-min time (leader (t -- s))
+and trailer, with \(s\) being their temporal spacing (inter-arrival
+time). From an operational perspective, highly efficient operations will
+maximise the utilisation of the available runway (system) capacity.
+Accordingly, we can assume that during peak times the temporal spacing
+at landing \(s\) accounts for the operational concept applied at the
+airport, and only comprises a minimal spacing error. Potential error
+sources include additional safety margins applied by the air traffic
+controller (or aircrew when landing under visual flight rules),
+variations of the arrival spacing due to reduction of airspeed during
+the flare. We consider the observed temporal spacing \(s_{ij}\) as a
+lower bound for safe operations.
+
+Using the constant time delay principle, the spacing deviation (or
+spacing error) at time t considers the current position of trailer at
+time \(t\), and the past position of leader at time \(t-s\). Based on
+our approach, the spacing deviation is defined as the temporal
+difference between the respective reference times for the position of
+the leader and trailer:
+
+\begin{equation}\protect\hypertarget{eq-spacing-deviation}{}{\begin{aligned}
+spacing deviation (t) =  ref{\_}time(trailer_{(t)}) - \\ref{\_}time(leader_{(t - s)})
+\end{aligned}}\label{eq-spacing-deviation}\end{equation}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{./figures/example-follower-leader-pairs.png}
+
+}
+
+\caption{\label{fig-follower-leader-example}Example timeline of spacing
+deviations between interacting arrivals}
+
+\end{figure}
+
+Equation~\ref{eq-spacing-deviation} does something.
 
@@
 \begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-icao2022}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-sun2017flightphase}{}}%
 \CSLLeftMargin{{[}1{]} }%
-\CSLRightInline{ICAO, {``Resolution A41-21: Consolidated statement of
-continuing ICAO policies and practices related to environmental
-protection {\textemdash} climate change,''} 2022 {[}Online{]}.
-Available:
-\url{https://www.icao.int/environmental-protection/Documents/Assembly/Resolution_A41-21_Climate_change.pdf}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sun2017flightphase}{}}%
-\CSLLeftMargin{{[}2{]} }%
 \CSLRightInline{J. Sun, J. Ellerbroek, and J. Hoekstra, {``Flight
@@
 Aerospace Information Systems}, vol. 14, no. 10, pp. 566--572, 2017. }
+
+\leavevmode\vadjust pre{\hypertarget{ref-ubertechnologiesinc2018}{}}%
+\CSLLeftMargin{{[}2{]} }%
+\CSLRightInline{Uber Technologies Inc, {``H3: Hexagonal hierarchical
+geospatial indexing system.''} 2018 {[}Online{]}. Available:
+\url{https://h3geo.org/}}
 

### 2023-04-04 paper.qmd
initial stab at background section

@@
     keep-tex: true
+  #docx: default
 
@@
 
-## header 2
+Trajectory-based operations (TBO) are a core element of the ongoing transformation programs for air navigation services (e.g. Europe/SESAR, US/Next Gen, Japan/CARATS).
+On the conceptual level TBO describe a move from clearance-based operations to a highly predictable and negotiated trajectory execution between airspace users and air traffic control.
+The TBO concept entails the successive refinement (i.e. negotiation) between the parties starting from an inital planned trajectory through operations and the actual execution of flight.
+TBO builds on higher levels of information exchange between planning and flow management systems and the airborne and ground systems during flight operations.
+@fig-TBO-concept shows the overall timeline and evolution of a trajectory.
+
+```{r}
+#| label: fig-TBO-concept
+#| fig-cap: Trajectory-based operations - conceptual perspective
+#| out-width: 45%  # 2 column IEEE
+knitr::include_graphics("./figures/TBO-concept.png")
+```
+
+Conceptual approach := technology agnostic
+CONOPS may entail technological enablers and/or advanced procedures
+Performance objectives observed on concept functional level
 
@@
 <!-- what did this paper show -->
+Ambitious political targets for 2050 have been formulated to curb the contribution of air transportation to climate change and work towards climate neutrality.
+Policy and industry action revolve around a variety of measures. 
+Within this context - and in light of the required lead time for a global pickup on sustainable aviation fuels, the introduction of novel airframe designs, or propulsion technology - increasing the operational efficiency is a declared now-term measure.
+There is a substantial benefit pool for the wider arrival phase covering the airspace within 200NM of the destination airport.
+Trajectory-based operations promise higher levels of predictability and reduced inefficiencies. 
+The benefits of reducing spacing and flow synchronisation measures in the terminal airspace are therefore dependent on the capability of balancing out operational variations from delivery of the aircraft to the wider arrival airspace.
 
-This paper aimed at exploring a data-driven approach to measuring arrival management based on open data.
+This proof-of-concept paper aims at operationalising a performance measure to characterise arrival management techniques, but also provide a basis to discuss the handover from the enroute phase to the wider arrival airspace.
+
+This paper aimed at exploring a data-driven approach to measuring operational efficiency during the arrival phase. 
+In light of the ongoing discussions about the climate impact of air transportation and the contr
+Arrival management based on open data.
 
 <!-- strength and weaknesses -->
+The proposed approach builds on the generalisation of a well-understood key performance metric for the wider arrival airspace.
+It follows a data-driven approach that is independent of detailed local context modelling.
+For example, local airspace restrictions or procedures are embedded in the temporal-spatial framework.
+Measuring the spacing deviation between interacting arrival flights provides a basis for the analysis of approach concepts that can help to streamline the arrival sequencing and contribute to the further reduction of unnecessary fuel burn and emissions.
+
+This proof-of-concept addresses the aspect of transparency by making use of openly available air transport trajectory data and open (analytical) software.
+The approach is readily transferrable to other regions.
 
 <!-- application -->
+The goal of this work was the initial development of an operational performance measure.
+Further validation is required at a larger set of European (and global airports).
+This will allow for the refinement of the parameterization of the approach.
+It is planned to roll-out the approach under the umbrella of the EUROCONTROL Performance Review System.
+This will provide European stakeholders with a means to address enhancement of arrival operations, and may help to build a catalogue of arrival management techniques that support higher levels of efficiency.
 
 <!-- further research -->
+This paper demonstrated feasibility of the approach.
+While the proposed performance monitoring of the spacing deviation envelope provides a basis for identifying inefficiencies during the arrival flow, there is a need to account also for factors driving and impacting the observed performance.
+The temporal assessment of spacing deviation
+
+* weather, wind
+* runway system configuration, adjacent aerodromes/level capping
+
+On the algorithmic level, future work should address the impact of the geospatial dimensions of cells and the method of their definition.
+
+This paper applied a observational approach to determining the reference times per cell. 
+This will pose challenges for longer time horizons as the continual update and processing of arrival trajectories would require substantial resources. 
+A possibility is the use of data structures to support the continual online accumulation of statistics without the need to store the complete historic dataset or processed artefacts.
 
 <!-- final appraisal -->
+The results demonstrate the general feasibility of the approach.
+The proposed approach offers a measure to start the dialogue between operational planners and stakeholders to identify benefit pools during the arrival phase.
 
@@
 
+<!--
 The script to download the weekly global datasets are included.
 The cleaned trajectory data is stored at: \<\<tbd\>\>.
+-->
 
@@
 
-The authors thank the comments by XX YY ZZ.
+The authors thank the comments by colleagues from the Performance Section of the Department of Airspace Control (DECEA) Brazil and the Performance Review Unit of EUROCONTROL. 
+This study would have not been feasible without the generous support of Opensky Network and the many community members collecting and feeding open ADSB data.
 
-The views expressed are the authors' own and do not represent a policy or position of EUROCONTROL <!-- and co-authors organisations -->.
+The views expressed are the authors' own and do not represent a policy or position of DECEA or EUROCONTROL <!-- and co-authors organisations -->.

### 2023-04-04 paper.tex
initial stab at background section

@@
 
-\hypertarget{header-2}{%
-\subsection{header 2}\label{header-2}}
+Trajectory-based operations (TBO) are a core element of the ongoing
+transformation programmes (e.g.~Europe/SESAR, US/Next Gen,
+Japan/CARATS). On the conceptual level TBO describe a move from
+clearance-based operations to a highly predictable and negotiated
+trajectory execution between airspace users and air traffic control. The
+TBO concept entails the successive refinement (i.e.~negotiation) between
+the parties starting from an inital planned trajectory through
+operations and the actual execution of flight. TBO builds on higher
+levels of information exchange between planning and flow management
+systems and the airborne and ground systems during flight operations.
+Figure~\ref{fig-TBO-concept} shows the overall timeline and evolution of
+a trajectory.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{./figures/TBO-concept.png}
+
+}
+
+\caption{\label{fig-TBO-concept}Trajectory-based operations - conceptual
+perspective}
+
+\end{figure}
+
+Conceptual approach := technology agnostic CONOPS may entail
+technological enablers and/or advanced procedures Performance objectives
+observed on concept functional level
 
@@
 \begin{equation}\protect\hypertarget{eq-spacing-deviation}{}{\begin{aligned}
-spacing deviation (t) =  ref{\_}time(trailer_{(t)}) - \\ref{\_}time(leader_{(t - s)})
+spacing \text{ } deviation (t) =  ref{\_}time(trailer_{(t)}) - \\ref{\_}time(leader_{(t - s)})
 \end{aligned}}\label{eq-spacing-deviation}\end{equation}
@@
 
+Ambitious political targets for 2050 have been formulated to curb the
+contribution of air transportation to climate change and work towards
+climate neutrality. Policy and industry action revolve around a variety
+of measures. Within this context - and in light of the required lead
+time for a global pickup on sustainable aviation fuels, the introduction
+of novel airframe designs, or propulsion technology - increasing the
+operational efficiency is a declared now-term measure. There is a
+substantial benefit pool for the wider arrival phase covering the
+airspace within 200NM of the destination airport. Trajectory-based
+operations promise higher levels of predictability and reduced
+inefficiencies. The benefits of reducing spacing and flow
+synchronisation measures in the terminal airspace are therefore
+dependent on the capability of balancing out operational variations from
+delivery of the aircraft to the wider arrival airspace.
+
+This proof-of-concept paper aims at operationalising a performance
+measure to characterise arrival management techniques, but also provide
+a basis to discuss the handover from the enroute phase to the wider
+arrival airspace.
+
 This paper aimed at exploring a data-driven approach to measuring
-arrival management based on open data.
+operational efficiency during the arrival phase. In light of the ongoing
+discussions about the climate impact of air transportation and the contr
+Arrival management based on open data.
+
+The proposed approach builds on the generalisation of a well-understood
+key performance metric for the wider arrival airspace. It follows a
+data-driven approach that is independent of detailed local context
+modelling. For example, local airspace restrictions or procedures are
+embedded in the temporal-spatial framework. Measuring the spacing
+deviation between interacting arrival flights provides a basis for the
+analysis of approach concepts that can help to streamline the arrival
+sequencing and contribute to the further reduction of unnecessary fuel
+burn and emissions.
+
+This proof-of-concept addresses the aspect of transparency by making use
+of openly available air transport trajectory data and open (analytical)
+software. The approach is readily transferrable to other regions.
+
+The goal of this work was the initial development of an operational
+performance measure. Further validation is required at a larger set of
+European (and global airports). This will allow for the refinement of
+the parameterization of the approach. It is planned to roll-out the
+approach under the umbrella of the EUROCONTROL Performance Review
+System. This will provide European stakeholders with a means to address
+enhancement of arrival operations, and may help to build a catalogue of
+arrival management techniques that support higher levels of efficiency.
+
+This paper demonstrated feasibility of the approach. While the proposed
+performance monitoring of the spacing deviation envelope provides a
+basis for identifying inefficiencies during the arrival flow, there is a
+need to account also for factors driving and impacting the observed
+performance. The temporal assessment of spacing deviation
+
+\begin{itemize}
+\tightlist
+\item
+  weather, wind
+\item
+  runway system configuration, adjacent aerodromes/level capping
+\end{itemize}
+
+On the algorithmic level, future work should address the impact of the
+geospatial dimensions of cells and the method of their definition.
+
+This paper applied a observational approach to determining the reference
+times per cell. This will pose challenges for longer time horizons as
+the continual update and processing of arrival trajectories would
+require substantial resources. A possibility is the use of data
+structures to support the continual online accumulation of statistics
+without the need to store the complete historic dataset or processed
+artefacts.
+
+The results demonstrate the general feasibility of the approach. The
+proposed approach offers a measure to start the dialogue between
+operational planners and stakeholders to identify benefit pools during
+the arrival phase.
 
@@
 
-The script to download the weekly global datasets are included. The
-cleaned trajectory data is stored at:
-\textless\textless tbd\textgreater\textgreater.
-
 \hypertarget{acknowledgment-and-disclaimer}{%
@@
 
-The authors thank the comments by XX YY ZZ.
+The authors thank the comments by colleagues from the Performance
+Section of the Department of Airspace Control (DECEA) Brazil and the
+Performance Review Unit of EUROCONTROL. This study would have not been
+feasible without the generous support of Opensky Network and the many
+community members collecting and feeding open ADSB data.
 
 The views expressed are the authors' own and do not represent a policy
-or position of EUROCONTROL .
+or position of DECEA or EUROCONTROL .
 

### 2023-04-04 paper.tex
clean-up / evening stop

@@
 Trajectory-based operations (TBO) are a core element of the ongoing
-transformation programmes (e.g.~Europe/SESAR, US/Next Gen,
-Japan/CARATS). On the conceptual level TBO describe a move from
-clearance-based operations to a highly predictable and negotiated
+transformation programs for air navigation services (e.g.~Europe/SESAR,
+US/Next Gen, Japan/CARATS). On the conceptual level TBO describe a move
+from clearance-based operations to a highly predictable and negotiated
 trajectory execution between airspace users and air traffic control. The
@@
 \end{figure}
+
+For this study we postulate the following goal of trajectory based
+arrival management: ``\ldots{} to reduce inefficiencies in the
+(extended) arrival phase by establishing an arrival sequence to require
+less costly sequencing close to the airport absorbing time/distance in
+more fuel efficient altitudes.''
 

### 2023-04-05 paper.qmd
clean-up / evening stop
update to background

@@
 On the conceptual level TBO describe a move from clearance-based operations to a highly predictable and negotiated trajectory execution between airspace users and air traffic control.
-The TBO concept entails the successive refinement (i.e. negotiation) between the parties starting from an inital planned trajectory through operations and the actual execution of flight.
+The TBO concept entails the successive refinement (i.e.
+negotiation) between the parties starting from an inital planned trajectory through operations (i.e.
+the actual execution of flight and agreed trajectory).
 TBO builds on higher levels of information exchange between planning and flow management systems and the airborne and ground systems during flight operations.
@@
 
-Conceptual approach := technology agnostic
-CONOPS may entail technological enablers and/or advanced procedures
-Performance objectives observed on concept functional level
+For this study we postulate the following goal of trajectory based arrival management:
+“… to reduce inefficiencies in the (extended) arrival phase by establishing an arrival sequence to require less costly sequencing close to the airport absorbing time/distance in more fuel efficient altitudes.”
+
+The focus of this paper is to evaluate the feasibility of studying arrival management techniques at different airports with a spatio-temporal approach.
+Previous work focussed on developing this concept, e.g.  liu & chrisinasen.
+Both groups studies the evolution of spacing over time and laid the foundation for the conceptualisation of spacing deviation concept and sequencing pressure 
+The benefit of this approach is to provide an operational performance measure that supports the assessment/comparison of different approach techniques.
+The existing performance framework (ICAO GANP, ECTROL , EU) apply the additioanl time concept to discrete points of the arrival airspace.
+These discrete positions represent the entry into the (near-airport) airspace defined by a concentric circle of 40NM or 100NM.
+
+The approach provides for the following performance measurement / monitoring benefits:
+* the conceptual approach is technology agnostic;
+* the local concept of operations specifies the respective techniques used and may entail technological enablers and/or advanced procedures; and
+* ultimately, performance objectives are observabel at the on concept functional level.
 

### 2023-04-05 paper.qmd
update of background

@@
 @fig-TBO-concept shows the overall timeline and evolution of a trajectory.
+ICAO is in the process of harmonising the global interpretation and application of the TBO concept (c.f.
+ICAO).
+Within that context a variety of scenarios and applications are described.
+This paper focusses on the arrival phase of flights and can be broadly assigned to the TBO-airport domain.
 
@@
 “… to reduce inefficiencies in the (extended) arrival phase by establishing an arrival sequence to require less costly sequencing close to the airport absorbing time/distance in more fuel efficient altitudes.”
+
+At the time being, there exists a group of predominant arrival management techniques broadly fullfilling this goal scope.
+
+* runway pressure: this concept aims at maximizing the utilisation of the arrival runway capacity in combination with increasing the flexibility of air traffic controllers to influence the sequence of arrivals.
+This can be achieved by holding aircraft in close(r) proximity to the airport allowing for close queue-based control of the arrival sequence.
+* (classical) vectoring: this concept builds on the mental model of an air traffic controller to establish the sequence as part of the standard approach procedures in combination with vectoring to adapt the desired landing sequence.
+* tromboning: in analogy of the music instrument, tromboning is based on area navigation points that can extend the arrival leg to build the sequence of successive arrivals.
+Aircraft are typically (re)cleared to follow one of the pathes defined by the respective turning points (trombone extensions).
+* point-merge: Point merge is designed to work in high traffic loads without radar vectoringpoint.
+It builds on a fixed gate through which a sequence of flights will proceed.
+The equidistant sequencing legs ensure the inter-arrival spacing and aircraft are cleared to the gate once the required separation is established.
 

### 2023-04-05 study_pipeline.qmd
initial stab at background section
next step towards putting paper together

@@
 
-```{r}
+```{}
 # European study data
@@
 ```
+
+to-do: read-in Brazil use-case, if time permits US use-case: KATL
+
+## Prepare study data set and write it out
+
+```{r}
+study_loop <- function(.apt, .fn, .debug, ...){
+
+  #----------- read airport data -------------------------
+  aip <- readr::read_csv("./data/aip-arp-rwys.csv", show_col_types = FALSE) |> 
+    dplyr::filter(ICAO == .apt)
+  if("GEOM_ALT" %in% names(aip)){
+    aip <- aip |> dplyr::rename(ELEV = GEOM_ALT)
+    }
+
+  arp     <- aip |> dplyr::filter(ICAO == .apt,   REF == "ARP")
+  rwys    <- aip |> dplyr::filter(ICAO == .apt, ! REF == "ARP")
+  rwy_centerline_ls <- cast_rwy_ctr_line_ls(rwys)
+  apt_box <- airport_centerline_box(rwys, ...)
+
+  # ---------------------- load trajectory data --------------------------------
+  warning(paste(.apt, " - reading - ", .fn))
+  # check entry file format
+  if(file_extension(.fn) == "parquet"){
+    ds <- arrow::read_parquet(.fn)
+  }
+  if(file_extension(.fn) == "feather"){
+    ds <- arrow::read_feather(.fn)
+  }
+  ds <- ds %>%
+    tibble::as_tibble() |>
+    #---- enforce above threshold ~ ELEV arp
+    dplyr::filter(ALT_B >= arp$ELEV + 300) |>
+     fix_altitude_heuristic() |>
+     Alt_Outlier_filter(.alt_var = ALT_B)
+
+  #---------------------- check for local traffic ----------------------------
+  # subset icao24s inside airport box
+  seen_at_airport <- ds |> 
+    dplyr::filter(ALT_B <= arp$ELEV + 4000) |> 
+    spatial_subset_jormungand(apt_box, ...) |> 
+    dplyr::distinct(ICAO24)
+
+  ds <- ds |> 
+    dplyr::filter(ICAO24 %in% seen_at_airport$ICAO24) |> 
+    identify_trajectory_legs(.max_gap = 5) |> 
+    mutate(UID = paste(ICAO24, LEG, sep = "-"))
+
+  phase <- ds |> 
+    extract_4d_nest() |> 
+    confirm_arrival_at(apt_box, rwy_centerline_ls) |> 
+    check_arrival_orphan()
+
+  # # # add landing time
+  phase_and_ldgtime <- add_landing_time(phase, rwys)
+}
+
+add_landing_time <- function(phase, rwys){
+
+  arrs <- phase |> filter(ARR & !ORPHAN)
+  rest <- phase |> filter(! UID %in% arrs$UID )
+
+  with_ldg_time <-  arrs |>
+    # estimate and assign landing time to arrival trajectories
+    dplyr::mutate(
+      study_trj = pmap(
+              .l = list(data, UID, RWY)
+            , .f = ~ add_landing_time_to_nested_data(..1, ..2, ..3, rwys)
+            )
+      ) |> 
+    # pullout ALDT as separate column
+    dplyr::mutate(
+      ALDT = map(.x = study_trj
+               , .f = ~ filter(.x, MST == "ALDT") |> pull(TIME)
+               )
+      ) |> tidyr::unnest(ALDT) |> 
+    # interpolate to 5sec
+    dplyr::mutate(
+      study_trj = map(.x = study_trj
+                    , .f = ~ interpolate_lin_4D(.x, .source = "OSN") |> 
+                             add_time_and_distance_to_go() |> 
+                             dplyr::mutate(H3_BIN = h3::geo_to_h3(c(LAT, LON), res = 8))
+                    )
+      )
+
+  return(bind_rows(with_ldg_time, rest))
+}
+# safe_add_landing_time <- purrr::possibly(.f = add_landing_time, otherwise = NULL)
+# 
+# # make this fail safe
+# safe_study_loop <- purrr::safely(.f = study_loop)
+```
+
+```{r}
+#| message: false
+apt <- "EGLL"
+rwys <- readr::read_csv("./data/aip-arp-rwys.csv", show_col_types = FALSE) |> 
+  filter(ICAO == apt) |> filter(TYPE == "THR") |> rename(ELEV = GEOM_ALT)
+
+#fn  <- list.files(path = "./data", pattern = paste0(apt,".*\\.feather"), full.names = TRUE)
+fn  <- list.files(path = "./data-study", pattern = paste0("^", apt, "-200NM-"), full.names = TRUE)[-c(1:2)]
+ptn <- as.Date(str_extract(fn, "\\d{4}-\\d{2}-\\d{2}"), format="%Y-%m-%d")
+
+purrr::walk2(.x = fn[5], .y[5] = ptn
+             ,  .f = ~ study_loop(apt, .x) |> 
+                     readr::write_rds(file = paste0("./data-study/mickey-", apt, "-", .y[5], ".rds") )
+)
+
+```

### 2023-04-05 R/add_distance_rowwise.R

@@ -38,7 +38,7 @@ add_time_and_distance_to_go <- function(.trj, .mst = "ALDT", .time_unit = "sec")
     dplyr::mutate(
       TIME_2GO = difftime(
           TIME[which(MST == .mst)], TIME, units = .time_unit)
-    , DIST_2GO = DIST_FLOWN[which(MST == .mst) - DIST_FLOWN])
+    , DIST_2GO = DIST_FLOWN[which(MST == .mst)] - DIST_FLOWN)
   
   return(df)
 }

### 2023-04-05 xhelper_get_file_extension.R

@@ -1,0 +1,3 @@
+file_extension <- function(filenames) {
+  sub(pattern = "^(.*\\.|[^.]+)(?=[^.]*)", replacement = "", filenames, perl = TRUE)
+}

### 2023-04-05 paper.qmd
Update paper.qmd

@@
 
-data preparation
+This study builds on open trajectory data from Opensky Network. 
+Opensky Network is a non-profit organisation and community-based surveillance data collection network. 
+The collected air traffic data is available to researchers and government organisations on a license-free basis.
+Performance monitoring is based on the principle of reproducibility and transparency. 
+Open access to the underlying data and enrusing reproducibility of the analysis is a key requirement [@PRC_OPDI2022).
 

### 2023-04-05 paper.qmd
Update paper.qmd

@@
 Performance monitoring is based on the principle of reproducibility and transparency. 
-Open access to the underlying data and enrusing reproducibility of the analysis is a key requirement [@PRC_OPDI2022).
+Open access to the underlying data and enrusing reproducibility of the analysis is a key requirement [@PRC_OPDI2022].
+For this feasibilty study we used a prepared download sample of the Opensky Network data comprising 10 days of operations in 2019 and 2022.
 
--   trajectory data - Opensky Network, weekly downloads
--   airport information - Openstreet Map
+The airport information required for this study was derived from Openstreet Map and validated by publicly available AIP information.
+Openstreet Map also builds on a community driven effort to provide accurate and timely maps. 
+It revolves around seeing the public as a central source of geographical information and voluteering to its maintenance.
+Over the past years, the information on airports has iteratively improved. 
+The principal information required for this paper comrpised the aerodrome reference points and runway thresholds and associated elevations.
 
-data downloaded & script development, data cleaning
+The primary input to this paper was the air traffic data downloaded from Opensky Network.
+The following sections describe the data preparatory and analytical steps taken and motivates the chosen heuristics and approaches.
 
